---
title: "KYOTO CHERRY BLOSSOM TREND (812-2015): RECREATING OUR WORLD IN DATA’S VISUALIZATION"
format: 
  html:
    theme: darkly
    code-fold: show
editor: source
---

Original blog post: <https://alierwaidatastudio.com/tutorials/posts/2024/03/09/tutorial-03>.

This tutorial aims to recreate an intriguing visual by Our World in Data named [Day of the year with peak cherry blossom in Kyoto, Japan](https://ourworldindata.org/grapher/date-of-the-peak-cherry-tree-blossom-in-kyoto?time=earliest..2023).

## Load packages

```{r}
#| label: laod-pkgs
#| output: false

rm(list = ls(all.names = TRUE))

library(tidyverse)
library(scales)
library(ggtext)

file_kyoto  <- "data/KyotoFullFlower7.xls"
icon_sakura <- "images/blossom.png"

invisible(Sys.setlocale("LC_TIME", "en_GB.UTF-8"))
```

## Import the data

```{r}
#| label: load-data

cherry_raw <- readxl::read_excel(path = file_kyoto, skip = 25) |> 
  set_names(nm = c("year", "full_flowering_date_doy", "full_flowering_date", "source", "type", "reference")) |> 
  mutate(
    date = str_c(year, str_sub(full_flowering_date, 1, 1), str_sub(full_flowering_date, 2, 3), sep = "-")
  ) |> 
  filter(!is.na(date)) |> 
  mutate(
    rolling_day_of_year = slider::slide_dbl(.x = full_flowering_date_doy, .f = mean, .before = 20, .complete = TRUE)
  )

tail(cherry_raw)
```

## Writing a function

We can write a function so that we can easily play with other moving average windows.

```{r}
#| label: define-function

tweak_cherry_trees_data <- function(data, window, cols) {
  
  data |> 
    janitor::clean_names() |> 
    select({{ cols }}) |> 
    mutate(
      date = str_c(
        year,
        str_sub(flowering_month, start = 1, end = 1),
        str_sub(flowering_month, start = 1, end = 1),
        sep = "-"
      )
    ) |> 
    filter(!is.na(date)) |> 
    mutate(
      rolling_day_of_year = slider::slide_dbl(
        .x = flowering_day_of_year,
        .f = mean,
        .before = {{ window }},
        .complete = TRUE
      )
    )
}
```

## Test the function

Let's test our new function

```{r}
#| label: test-function

cherry_trees_raw <- readxl::read_excel(path = file_kyoto, skip = 25)

# Clean and transform the data
cherry_trees_data_tbl <- tweak_cherry_trees_data(
  data = cherry_trees_raw, 
  window = 20, 
  cols = c(
    year = ad,
    flowering_day_of_year = full_flowering_date_doy,
    flowering_month = full_flowering_date
  )
)
```

## Plotting cherry trees data

```{r}
#| label: plot-cherry-trees
#| warning: false

# Plot cherry trees data ------
myplot <- ggplot(
  data = cherry_trees_data_tbl, 
  mapping = aes(x = year, y = flowering_day_of_year)
  ) +
  geom_point(na.rm = TRUE, size = 0.8) +
  geom_line(
    mapping = aes(y = rolling_day_of_year),
    na.rm = TRUE,
    linewidth = 1.2,
    color = "#5e2ce8"
  ) +
  annotate(
    geom = "text",
    x = 1990,
    y = 80,
    label = "20-Year\n Average",
    color = "#5e2ce8",
    size = 4,
    na.rm = TRUE
  ) +
  scale_x_continuous(
    limits = c(812, 2015),
    breaks = c(812, 1000, 1200, 1400, 1600, 1800, 2023)
  ) +
  scale_y_continuous(
    limits = c(70, 120),
    labels = c(
      "March 11", "March 21", "March 31",
      "April 10", "April 20", "April 30"
    )
  ) +
  ggthemes::theme_wsj() +
  labs(
    title = "Cherry trees have been blossoming earlier due to warmer spring temperatures",
    subtitle = "Date of peak cherry tree (***Prunus jamasakura***) in Kyoto, Japan",
    caption = "**Data Source: Yasuk Aono (2021:2024)**<br>**Note:** The <span style='color: #5e2ce8'>20-year average</span> is calculated when there are at least five years with data in the <span style='color: #5e2ce8'>20-year window</span>. The dates are off by one day on the years.<br><br>
        OurWorldInData.org/biodiversity|CC BY"
  ) +
  theme(
    plot.title = element_textbox_simple(
      size = 12,
      margin = margin(t = 15, r = 0, b = 5, l = 0),
      lineheight = 1.2,
      linewidth = 0.3
    ),
    plot.title.position = "plot",
    plot.subtitle = element_textbox_simple(
      size = 10,
      hjust = 0,
      margin = margin(t = 15, r = 0, b = 15, l = 0),
      lineheight = 1.2,
      linewidth = 0.1
    ),
    axis.text = element_text(
        size = 8,
        color = "darkgray"
    ),
    plot.caption = element_textbox_simple(
      size = 8,
      margin = margin(t = 15, r = 0, b = 10, l = 0),
      lineheight = 1.2,
      linewidth = 0.1
    ),
    axis.line = element_line(color = "darkgray"),
    axis.ticks = element_line(color = "darkgray"),
    plot.caption.position = "plot",
    plot.margin = margin(15, 15, 15, 15)
  )

print(myplot)
```

Save the plot

```{r}
#| label: save-plot
#| output: false

ggsave("images/cherry_bomb_04.png", width = 25, height = 18, units = "cm", dpi = 300)
```
