---
title: "Peak Bloom Prediction"
author: "econmaett"
lang: en-US
format:
  html:
    embed-resources: true
fig-cap-location: bottom
tbl-cap-location: bottom
toc: true
editor: source
---

Learn more about phenology at <https://en.wikipedia.org/wiki/Phenology>.

## Packages

We start by loading all the packages necessary to run this analysis

```{r}
#| label: load-pkgs
#| output: false
#| warning: false
#| message: false

invisible(Sys.setlocale("LC_ALL","English"))

library(tidymodels)
library(tidyverse)
library(arrow)
library(bonsai)
library(rules)
library(finetune)
library(infer)
library(probably)
library(kernlab)
library(corrr)
library(rpart.plot)
library(vip)
library(brulee)
library(stacks)
library(gt)
library(nasapower)
library(openmeteo)

tidymodels_prefer()
```

## Data

We begin by loading the data necessary for this analysis

### Clean data

The data for the five sites is provided as a simple text file in CSV format. Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site. The six columns in each data file are:
- `location`: a human-readable location identifier (`string`).
- `lat`: (approximate) latitude of the cherry trees (`double`).
- `long`: (approximate) longitude of the cherry trees (`double`).
- `alt`: (approximate) altitude of the cherry trees (`double`).
- `year`: year of the observation (`integer`).
- `bloom_date`: date of peak bloom of the cherry trees ([ISO 8601](https://xkcd.com/1179/) date `string`). The "peak bloom date" may be defined differently for different sites
- `bloom_doy`: days since January 1st until peak bloom (`integer`). January 1st is `1`.

#### Kyoto

The file `kyoto.csv` has been obtained from Yasuyuki Aono (`aono(at)envi.osakafu-u.ac.jp`) at <http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/>.

The geographical location at `35.01198°N` (latitude), `135.6761°E` (longitude), `44` meters above sea levels (altitude) roughly corresponds to the **Nakanoshima** area of the **Arashiyama Park** in Kyoto, Japan. Peak bloom is defined as **80%** of flowers being in full bloom.

The peak bloom date of the **Prunus jamasakura (Japanese mountain cherry)** is determined by a local news paper in Arashiyama (Kyoto, Japan). Data prior to 1888 is extracted from various descriptions or estimated:
- Data from the 9th to the 14th centuries was acquired and analyzed by Aono and Saito (2010; International Journal of Biometeorology, 54, 211-219).
- Phenology for 15th to 21st centuries was acquired and analyzed by Aono and Kazui (2008; International Journal of Climatology, 28, 905-914).

Note that the data set for Kyoto includes implicit missing values (`NA`s). [Chapter 18 "Missing Values"](https://r4ds.hadley.nz/missing-values) of Hadley Wickham's book "R for Data Science (2e)" describes strategies for dealing with implicit `NA`s.

```{r}
#| label: load-kyoto
#| output: false
#| warning: false
#| message: false

kyoto <- read_csv("data/kyoto.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(country = "japan") |> 
  fill(location, lat, long, alt, .direction = "downup")
```

Plot the time series for Kyoto

```{r}
#| label: fig-kyoto
#| fig-cap: "Peak cherry blossom days in Kyoto, Japan"
#| warning: false
#| message: false

fig_kyoto <- kyoto |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.5) +
  geom_point(color = "#ff0054", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(800, 2000, by = 100), NA)) +
  labs(
    title = "Peak cherry blossom days in Kyoto, Japan", 
    subtitle = "Cultivar: Prunus jamasakura (Japanese mountain cherry)",
    x = NULL, y = NULL,
    caption = "Source: Yasuyuki Aono"
  ) +
  theme_light()

print(fig_kyoto)

ggsave(filename = "figures/fig_kyoto.png", plot = fig_kyoto, bg = "white")
```

Here we have used **locally estimated scatterplot smoothing (LOESS)** or **locally weighted scatterplot smoothing (LOWESS)** to highlight how the average peak bloom day changed over the years. 

We could explicitly model this with a **local level (+ trend)** state-space model that we fit with the **Kalman filter and Smoother**, which also allows us to effectively deal with **missing observations**.

MeteoSwiss suggests using a **Gaussian Kernel** with a *20-year window* to extract the trend, which is equivalent to using a **low-pass filter** that shuts down all higher frequencies.

It also looks like the observations exhibit **heteroskedasticity**, that is, the variance changes over time, and it looks like the variance has become more narrow over time.

We can use **changepoint detection** methods to test for the changing time trend and the changing variance.

Assuming that the data in the past is correct, there do seem to be seasonal cycles, perhaps connected to solar activity or the tilt of the earth.

There is a clear trend towards earlier blooming, perhaps connected to anthropogenic climate change and urbanization. Studies attribute about half of the earlier blooming times in recent years to the **urban heat island effect**, and the other half to **climate change**.

Modelling the local level and the trend would be useful if we are to forecast a couple years into the future, since an individual prediction may be incorrect, but on average, they may be good. 

However, since we are only interested in forecasting a single moment in time, which may be greatly influenced by local weather, we probably are better off using historical weather data that we augment with forecasts, either by professional forecasters or by extrapolating the previous predictors.

#### Liestal

The file `liestal.csv` contains observations for **a single wild cherry (prunus avium) tree** in **Liestal-Weideli**, at `7.4814°N` (latitude), `7.730519°E` (longitude) and `350` meters above sea levels (altitude). 

```{r}
#| label: load-liestal
#| output: false
#| warning: false
#| message: false

liestal <- read_csv("data/liestal.csv") |> 
  mutate(country = "switzerland")
```

The **peak bloom date** is defined as the day when **25%** of the blossoms are in full bloom. The date is determined by [MeteoSwiss](https://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html). This means that 70% of buds would be open a some days later. This raises the question of whether we should model the locations separately.

The data is provided by **Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss**.

Let's plot the time series for Liestal, Switzerland:

```{r}
#| label: fig-liestal
#| fig-cap: "Peak cherry blossom days for Liestal-Weideli, Switzerland"
#| warning: false
#| message: false

fig_liestal <- liestal |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Liestal-Weideli, Switzerland",
    subtitle = "Cultivar: Wild cherry (prunus avium)",
    x = NULL, y = NULL,
    caption = "Source: Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss"
  ) +
  theme_light()

print(fig_liestal)

ggsave(filename = "figures/fig_liestal.png", plot = fig_liestal, bg = "white")
```

For Liestal you can see the impact of the infamous heat summer of 2003 which I remember from my childhood, although interestingly, there's another record in 1990.

#### New York

The file `newyorkcity.csv` contains **only a single observation** for the year 2024. It concerns the **Yoshino cherry (prunus × yedoensis)** tree in **Washington Square Park**, New York City, at `40.7304 °N` (latitude), `-73.99809°E` (longitude) and `8.5` meters above sea level (altitude). Peak bloom is defined as **70%** of flowers being in full bloom.

```{r}
#| label: load-newyorkcity
#| output: false
#| warning: false
#| message: false

newyorkcity <- read_csv("data/nyc.csv") |> 
  mutate(country = "unitedstates")

nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789, Species_ID == 228) |> 
  mutate(Observation_Date = mdy(Observation_Date))

nyc_data <- nyc_data_npn |>
  arrange(Observation_Date) |>
  mutate(year = year(Observation_Date)) |>
  group_by(year) |>
  summarize(
    first_flower_index = min(which((Intensity_Value %in% c("75-94%", "95% or more")))),
    bloom_date = as_date(strftime(Observation_Date[first_flower_index], format = "%Y-%m-%d")),
    bloom_doy = Day_of_Year[first_flower_index],
    .groups = "drop"
  ) |>
  filter(!is.na(bloom_doy)) |>
  select(-first_flower_index) |>
  mutate(country = "unitedstates", location = "newyorkcity") |> 
  arrange(year)

newyorkcity <- newyorkcity |>
  bind_rows(nyc_data |> filter(year != 2024)) |>  
  fill(lat, long, alt)
```

Luckily, there are some historical observations dating *back to 2019* in the data provided by **USA-NPN**. The Washington Square Park has site id `32789` and the **Yoshino** cherry you should predict has species id `228`.

This file now contains individual observations of the phenophase and not the actual peak bloom date. We use the variable `Intensity_Value` which is either `-9999`, `95% or more`, `Less than 5%`, `25-49%`, `75-94%`, `50-74%`. Since according to the variable `Individual_ID`, only one tree, `183140` is being observed, we can use the first date where `Intensity_Value` is `50-74%` as our `bloom_doy`.

#### Vancouver

The file `vancouver.csv` contains observations for cherry trees of the cultivar  **Prunus × yedoensis "Akebono" (daybreak cherry)** in **Maple Grove Park**, Vancouver, BC located approximately at `49.2236916°N` (latitude), `-123.1636251°E` (longitude), `24` meters above sea levels (altitude). Peak bloom is defined as **70%** of flowers being in full bloom.

```{r}
#| label: load-vancouver
#| output: false
#| warning: false
#| message: false

vancouver <- read_csv("data/vancouver.csv") |> 
  mutate(country = "canada")

vancouver_forum <- tribble(
  ~ year, ~ bloom_date,
  2021, "2021-03-31", # Start: 2021-03-21, Full: 2021-03-31, Full: 2021-04-07
  2020, "2020-04-03", # Full bloom
  2018, "2018-03-18", # Full bloom
  2017, "2017-04-11", # Full bloom 
  2015, "2015-03-19", # Full bloom
  2011, "2011-04-01", # Full bloom was over on 2011-04-15
  2010, "2010-03-14", # 60% bloom on 2010-03-07
  2008, "2008-04-14" # Start: 2008-04-06, Full: 2008-04-14, finished: 2008-04-22
  ) |> 
  mutate(
    bloom_date = as_date(bloom_date),
    bloom_doy = as.numeric(1 + as_date(bloom_date) - as_date(paste0(year, "-01-01"))),
    location = "vancouver",
    country = "canada"
  )

vancouver <- vancouver |> 
  bind_rows(vancouver_forum) |> 
  fill(lat, long, alt) |> 
  arrange(year)
```

Unfortunately, there are **only three observations** for 2022, 2023, and 2024, respectively. But casual observations have been recorded in the form of **photos posted** to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/). You can **search the forum** for the keywords `"Akebono"` (i.e., the name of the cultivar) and `"Maple Grove Park"` (i.e., the location of the trees).

We can do some additional plausibility checks. For example, spring 2010 was indeed unusually warm, [causing cherry blossoms to bloom during the Vancouver 2010 Winter Olympics](https://www.upi.com/News_Photos/view/upi/33393f9d20159ecb347de9aa1d670922/2010-Winter-Olympics-in-Vancouver/). National Geographic claimed it was the [warmest Olympics to date](https://www.nationalgeographic.com/adventure/article/100212-vancouver-2010-warmest-winter-olympics).

#### Washington

The file `washingtondc.csv` has been obtained from the EPA's Climate Change Indicators in the United States <https://www.epa.gov/climate-indicators/cherry-blossoms>.

```{r}
#| label: load-washingtondc
#| message: false
#| warning: false
#| output: false

washingtondc <- read_csv("data/washingtondc.csv") |> 
  mutate(country = "unitedstates")
```

The geographical location at `38.88535°N` (latitude), `-77.03863°E` (longitude) and `0` meters above sea level (altitude) correspond to the location of the [Tidal Basin in Washington, DC](https://www.nps.gov/articles/dctidalbasin.htm).

The peak bloom date is defined as the day when **70%** of the **Prunus × yedoensis "Somei-yoshino"** are in full bloom, as determined by the [National Park Service](https://www.nps.gov/subjects/cherryblossom/bloom-watch.htm).

Let's plot the data for Washington D.C

```{r}
#| label: fig-washingtondc
#| fig-cap: "Peak cherry blossom days for Washington DC, USA"
#| warning: false
#| message: false

fig_washingtondc <- washingtondc |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Washington DC, USA",
    subtitle = "Cultivar: Yoshino cherry (prunus × yedoensis )",
    x = NULL, y = NULL,
    caption = "Source: Environmental Protection Agency (EPA) Climate Change Indicators in the United States"
  ) +
  theme_light()

print(fig_washingtondc)

ggsave(filename = "figures/fig_washingtondc.png", plot = fig_washingtondc, bg = "white")
```

Now that we have loaded the datasets for the five locations, we can combine them into a single CSV file called `cherry.csv`.

```{r}
#| label: combine-cherry-csv
#| output: false
#| warning: false
#| message: false

cherry <- bind_rows(kyoto, liestal, newyorkcity, vancouver, washingtondc) |> 
  relocate(year, location, country, lat, long, alt, bloom_date)
```

We can visualize the time series for our five main locations next to each other. Notice the use of the `{stringr}` function `str_to_title()` to convert the first letter of the locations to uppercase.

```{r}
#| label: fig-cherries_1880
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at five different sites.

df_cherry_1880 <- cherry |> 
  filter(year >= 1880)

fig_cherries <- ggplot(data = df_cherry_1880, aes(year, bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab") +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 40), NA)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(
    title = "Peak bloom (since Jan 1st)",
    x = NULL, 
    y = NULL
  ) +
  theme_light()

print(fig_cherries)

ggsave(filename = "figures/fig_cherries.png", plot = fig_cherries, bg = "white")
```

### Additional data

The competition organizers provide additional data sets for sites other than the main sites relevant for the competition. We can use these time series in your modeling to help with spatial and temporal extrapolation. For example, we can add predictors to these data sets and train models on additional sites.

#### Japan

The data file `japan.csv` is provided by the [Japanese Meteorological Agency (JMA)](https://www.data.jma.go.jp/sakura/data/) and contains recorded peak bloom dates for cherry trees at various sites across Japan. The sample trees are located within a 5km radius of the location indicated in the data file. 

Load the data set, make implicit missing values explicit and change the `location` variable to only include the city names in lowercase letters.

```{r}
#| label: read-japan
#| warning: false
#| message: false
#| output: false

japan <- read_csv("data/japan.csv") |>  
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Japan/")) |> 
  mutate(location = tolower(location), country = "japan") |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))

sendai <- japan |> 
  filter(location == "sendai", alt == 38.85)

hakodate <- japan |> 
  filter(location == "hakodate", alt == 35)

niigata <- japan |> 
  filter(location == "niigata", alt == 4.1)
```

Similar climate to Vancouver, BC: Sendai, Hakodate, and Niigata

Let's plot peak bloom dates for selected locations in Japan

```{r}
#| label: fig-bloom-japan
#| fig-cap: "Peak bloom dates for selected locations Japan"
#| warning: false
#| message: false

fig_japan <- japan |>
  filter(location %in% c("sendai", "hakodate", "niigata")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 20), NA)) +
  facet_wrap(facets = vars(location)) +
  labs(
    title = "Peak bloom (since Jan 1st) for selected locations in Japan",
    subtitle = NULL,
    x = NULL, y = NULL,
    caption = "Source: Japanese Meteorological Agency (JMA)"
  ) +
  theme_light()

print(fig_japan)

ggsave(filename = "figures/fig_japan.png", plot = fig_japan, bg = "white")
```

#### MeteoSwiss

The data file `meteoswiss.csv` contains peak bloom dates for various sites across Switzerland, obtained from <https://opendata.swiss/en/dataset/phanologische-beobachtungen>.

Note that when you use the `{utils}` function `read.csv()` you need to specify the option `encoding = "UTF-8"` to allow for special characters that are common in the German language.

```{r}
#| label: load-meteoswiss
#| warning: false
#| message: false
#| output: false

meteoswiss <- read_csv("data/meteoswiss.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Switzerland/")) |> 
  mutate(location = tolower(location), country = "switzerland") |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))

binningen <- meteoswiss |> 
  filter(location == "basel-binningen") |> 
  filter(alt == 315) |> 
  mutate(location = str_remove(location, "basel-"))

laufenburg <- meteoswiss |> 
  filter(location == "laufenburg") |> 
  filter(alt == 330)

albisgutli <- meteoswiss |> 
  filter(location == "zürich-albisgütli") |> 
  mutate(location = str_remove(location, "zürich-")) |> 
  filter(alt == 530)

meteoschweiz <- meteoswiss |> 
  filter(location == "zürich-meteoschweiz") |> 
  mutate(location = str_remove(location, "zürich-")) |> 
  filter(alt == 555)

witikon <- meteoswiss |> 
  filter(location == "zürich-witikon") |> 
  mutate(location = str_remove(location, "zürich-")) |> 
  filter(alt == 620)
```

The locations have up to 70 observations, but most are at very high altitudes, much higher than the five main locations. I therefore select two locations which have a similar altitude to Liestal and are in close geographical proximity.

I enhance the dataset of the five main locations with observations from the selected locations in Japan and Switzerland. I do not use data from Korea because the observations are few and the phenological event observed is first bloom and not peak bloom.

```{r}
#| label: enhance-cherry-with-selected-locations
#| warning: false
#| message: false

cherry <-bind_rows(
  kyoto,
  liestal,
  newyorkcity,
  vancouver,
  washingtondc,
  sendai
  hakodate,
  niigata,
  binningen,
  laufenburg,
  albisgutli,
  meteoschweiz,
  witikon
)
```

## Predictors

We have seen that a linear time trend alone is not sufficient to explain the variance in the peak bloom times. It therefore makes sense to include external data on local weather conditions in our model. 

### POWER

We use the `{nasapower}` package described at <https://docs.ropensci.org/nasapower/> to access the [NASA Prediction Of Worldwide Energy Resources (POWER) API](https://power.larc.nasa.gov/). We create a custom function called `get_power()` to repeatedly query the API with the `nasapower::get_power()` function.

```{r}
#| label: get_power

get_power <- function(
    vars, 
    start = "1981-01-01", 
    end = as.character(Sys.Date()), 
    freq = "daily", 
    lat, long
  ) {
  power <- map2(
    .x  = long,
    .y = lat,
    .f = function(long = .x, lat = .y) {
      res <- nasapower::get_power(
        community = "ag",
        lonlat = c(long, lat),
        pars = vars,
        dates = c(start, end),
        temporal_api = freq
        )
      return(res)
      }
    ) |>
    list_rbind()
  return(power)
}
```

We load the following variables:

- `T2M`: Temperature at 2 Meters. The average air (dry bulb) temperature at 2 meters above the surface of the earth.

- `T2M_MAX:` Temperature at 2 Meters Maximum. The maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `T2M_MIN`: Temperature at 2 Meters Minimum. The minimum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `PRECTOTCORR` Corrected precipitation

- `RH2M`: Relative Humidity at 2 Meters. The ratio of vapor pressure to the saturation vapor pressure with respect to a plane surface of pure water, expressed in percent.

```{r}
#| label: specify-vars

vars <- c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "PRECTOTCORR",
  "RH2M"
)
```

We retrieve the **coordinates** which we will use to match to the nearest weather stations

```{r}
#| label: get-coordinates

crd_kyoto        <- distinct(kyoto, location, country, lat, long, alt)
crd_liestal      <- distinct(liestal, location, country, lat, long, alt)
crd_nyc          <- distinct(newyorkcity, location, country, lat, long, alt)
crd_vancouver    <- distinct(vancouver, location, country, lat, long, alt)
crd_washingtondc <- distinct(washingtondc, location, country, lat, long, alt)

crd_sendai <- distinct(sendai, location, country, lat, long, alt)
crd_hakodate <- distinct(hakodate, location, country, lat, long, alt)
crd_niigata <- distinct(niigata, location, country, lat, long, alt)

crd_binningen    <- distinct(binningen, location, country, lat, long, alt)
crd_laufenburg   <- distinct(laufenburg, location, country, lat, long, alt)
crd_albisgutli   <- distinct(albisgutli, location, country, lat, long, alt)
crd_meteoschweiz <- distinct(meteoschweiz, location, country, lat, long, alt)
crd_witikon      <- distinct(witikon, location, country, lat, long, alt)
```

Load the predictors for the locations

```{r}
#| label: load-nasa-power
#| eval: false

# Five main locations
power_kyoto <- get_power(vars, lat = crd_kyoto$lat, long = crd_kyoto$long) |> 
  left_join(crd_kyoto, by = c("LAT" = "lat", "LON" = "long"))

power_liestal <- get_power(vars, lat = crd_liestal$lat, long = crd_liestal$long) |> 
  left_join(crd_liestal, by = c("LAT" = "lat", "LON" = "long"))

power_nyc <- get_power(vars, lat = crd_nyc$lat, long = crd_nyc$long) |> 
  left_join(crd_nyc, by = c("LAT" = "lat", "LON" = "long"))

power_vancouver <- get_power(vars, lat = crd_vancouver$lat, long = crd_vancouver$long) |> 
  left_join(crd_vancouver, by = c("LAT" = "lat", "LON" = "long"))

power_washingtondc <- get_power(vars, lat = crd_washingtondc$lat, long = crd_washingtondc$long) |> 
  left_join(crd_washingtondc, by = c("LAT" = "lat", "LON" = "long"))

# Selected locations in Japan
power_sendai <- get_power(vars, lat = crd_sendai$lat, long = crd_sendai$long) |>
  left_join(crd_sendai, by = c("LAT" = "lat", "LON" = "long"))

power_hakodate <- get_power(vars, lat = crd_hakodate$lat, long = crd_hakodate$long) |>
  left_join(crd_hakodate, by = c("LAT" = "lat", "LON" = "long"))

power_niigata <- get_power(vars, lat = crd_niigata$lat, long = crd_niigata$long) |>
  left_join(crd_niigata, by = c("LAT" = "lat", "LON" = "long"))

# Selected locations in Switzerland
power_binningen <- get_power(vars, lat = crd_binningen$lat, long = crd_binningen$long) |>
  left_join(crd_binningen, by = c("LAT" = "lat", "LON" = "long"))
 
power_laufenburg <- get_power(vars, lat = crd_laufenburg$lat, long = crd_laufenburg$long) |>
  left_join(crd_laufenburg, by = c("LAT" = "lat", "LON" = "long"))

power_albisgutli <- get_power(vars, lat = crd_albisgutli$lat, long = crd_albisgutli$long) |>
  left_join(crd_albisgutli, by = c("LAT" = "lat", "LON" = "long"))

power_meteoschweiz <- get_power(vars, lat = crd_meteoschweiz$lat, long = crd_meteoschweiz$long) |>
  left_join(crd_meteoschweiz, by = c("LAT" = "lat", "LON" = "long"))

power_witikon <- get_power(vars, lat = crd_witikon$lat, long = crd_witikon$long) |>
  left_join(crd_witikon, by = c("LAT" = "lat", "LON" = "long"))
```

Combine the downloaded data sets and save them to a CSV file

```{r}
#| label: combine-power_cherry
#| eval: false

power_cherry <- bind_rows(
  power_kyoto,
  power_liestal,
  power_nyc,
  power_vancouver,
  power_washingtondc,
  power_sendai,
  power_hakodate,
  power_niigata,
  power_binningen,
  power_laufenburg,
  power_albisgutli,
  power_meteoschweiz,
  power_witikon
  ) |> 
  rename_with(.fn = tolower, .cols = everything()) |> 
  rename(date = yyyymmdd, long = lon, month = mm, day = dd, prcp = prectotcorr) |> 
  relocate(location, country, date, lat, long, alt, year, month, day, doy)

write_csv(power_cherry, file = "data/power_cherry.csv")
```

### Open Meteo

Use the [`{openmeteo}`](https://CRAN.R-project.org/package=openmeteo) package to retrieve weather forecasts from the [Open Meteo API](https://open-meteo.com/).

We have already used the coordinates of the five main locations so all we have to do is to combine them into a list. Because I only want to forecast the five main locations for the current year, I will only load weather forecasts for these locations.

Note that fore locations in Japan and Switzerland, the coordinates for individual trees are given, and I have to select a single location to draw weather forecasts from, hence I choose either the most common location or the one in the middle. Since the coordinates are matched to the nearest weather station, such a choice should not matter much.

```{r}
#| label: list-locations

locations <- list(
  crd_kyoto,
  crd_liestal,
  crd_nyc,
  crd_vancouver,
  crd_washingtondc,
  crd_sendai,
  crd_hakodate,
  crd_niigata,
  crd_binningen,
  crd_laufenburg,
  crd_albisgutli,
  crd_meteoschweiz,
  crd_witikon
)
```

We create a `get_fcst()` function that calls `openmeteo::weather_forecast()`. We can only get 15 days ahead weather forecasts, and we are about 3 days behind with the NASA POWER data.

```{r}
#| label: get_fcst

get_fcst <- function(loc_gc) {
  
  fcst <- weather_forecast(
    location = c(loc_gc$lat, loc_gc$long), 
    start = as.character(today() - days(3)), 
    end = as.character(today() + days(12)), 
    hourly = c(
      "temperature_2m", 
      "relative_humidity_2m"
    ),
    daily = c(
      "temperature_2m_max",
      "temperature_2m_min"
    )
  )
  
  fcst$location <- tolower(loc_gc$location)
  
  return(fcst)
}
```

Use the `get_fcst()` function to load the weather forecasts for the five main locations

```{r}
#| label: load-weather-forecasts
#| warning: false
#| message: false
#| eval: false

locations_fcst <- map_dfr(.x = locations, .f = get_fcst) |> 
  rename(
    t2m = hourly_temperature_2m,
    t2m_max = daily_temperature_2m_max,
    t2m_min = daily_temperature_2m_min,
    rh2m = hourly_relative_humidity_2m
  ) |> 
  mutate(rh2m = rh2m * 28/26) |> 
  group_by(location, date) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  relocate(location, date, t2m, t2m_max, t2m_min) |> 
  mutate(across(where(is.numeric), ~ round(.x, 2))) |> 
  mutate(prcp = NA)

write_csv(locations_fcst, file = "data/locations_fcst.csv")
```

We need to forecast the precipitation variable from NASA POWER because it does not align with the rainfall variable from Open Meteo. Relative humidity also has to be re-weighted by a factor of 28/26.

Read in the forecasts

```{r}
#| label: combine-power-openmeteo
#| output: false
#| warning: false
#| message: false

power_cherry <- read_csv("data/power_cherry.csv")

locations_fcst <- read_csv("data/locations_fcst.csv")

last_obs_date <- power_cherry |> 
  filter(!is.na(t2m)) |> 
  slice_max(order_by = date) |> 
  select(date) |> 
  pull() |> 
  unique()

power_cherry_fcst <- bind_rows(
  power_cherry,
  locations_fcst |> 
    filter(date > last_obs_date)
  ) |> 
  filter(!is.na(t2m))

write_csv(power_cherry_fcst, file = "data/power_cherry_fcst.csv")
```


## Feature engineering

### Midrange temperature

We create the midrange temperature, the average of the maximum and the minimum average temperature measured on a specific day.

```{r}
#| label: create-t2m_range

power_cherry_fcst <- power_cherry_fcst |> 
  mutate(t2m_range = (t2m_max + t2m_min) / 2) |> 
  relocate(location, date, t2m, t2m_max, t2m_min, t2m_range)
```

### Aggregated precipitation since January 1st

```{r}
#| label: create-aprcp

power_cherry_fcst <- power_cherry_fcst |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(
    aprcp = cumsum(prcp)
  ) |> 
  ungroup()
```


### Aggregated growing degree days since January 1st

We create the accumulated growing degree days (AGDD) as the sum of growing degree days (GDD) since January 1st, where GDD are calculated as the number of degrees C by which each day's midrange temperature exceeds a base temperature, which can be 0 degrees C, or a species-specific base temperature, such as 4.3 degrees C for Yoshino cherry trees. If the midrange temperature does not exceed the base temperature on a given day, no growing degree days are counted. See more at <https://en.wikipedia.org/wiki/Growing_degree-day>.

```{r}
#| label: create-agdd

base_temp <- 4.0

power_cherry_fcst <- power_cherry_fcst |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(
    gdd = pmax(t2m_range - base_temp, 0),
    agdd = cumsum(gdd)
  ) |> 
  ungroup()
```

## Select time range

I select only days between October 1st and the last available forecast date of the current year. That way, the model won't rely on data which is not available on the day of the submission deadline.

```{r}
#| label: select-time-range
#| warning: false
#| message: false

last_fcst_date <- power_cherry_fcst |> 
  filter(!is.na(t2m_max)) |> 
  select(date) |>
  pull() |> 
  max()

power_cherry_fcst <- power_cherry_fcst |> 
  mutate(doy = yday(date)) |> 
  relocate(location, country, date, year, month, day, doy, lat, long, alt) |> 
  filter(doy <= yday(last_fcst_date))
```

## Temporal aggregation

```{r}
#| label: aggregate-to-months

df_months <- power_cherry_fcst |> 
  select(-c(country, year, month, day, doy, lat, long, alt, prcp, gdd, t2m_range, t2m)) |>
  mutate(year = year(date), month = month(date, label = TRUE, abbr = TRUE)) |> 
  group_by(location, year, month) |> 
  summarise(across(where(is.numeric), mean), .groups = "drop") |> 
  ungroup() |> 
  pivot_longer(cols = -c(location, year, month), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", month)) |> 
  select(-c(month)) |> 
  pivot_wider(names_from = variable, values_from = value) |> 
  mutate(
    # prcp_Feb = if_else(is.na(prcp_Feb), mean(prcp_Feb, na.rm = TRUE), prcp_Feb),
    aprcp_Feb = if_else(is.na(aprcp_Feb), mean(aprcp_Feb, na.rm = TRUE), aprcp_Feb),
    # prcp_Mar = if_else(is.na(prcp_Mar), mean(prcp_Mar, na.rm = TRUE), prcp_Mar),
    aprcp_Mar = if_else(is.na(aprcp_Mar), mean(aprcp_Mar, na.rm = TRUE), aprcp_Mar)
  )

cherry_months <- cherry |> 
  full_join(y = df_months, join_by(location, year)) |> 
  select(-c(bloom_date, country, lat, long, alt)) |>
  left_join(
    y = locations |> 
      list_rbind(), 
    by = join_by(location)
  ) |> 
  arrange(year, location) |>
  filter(year >= 1981) |> 
  select(-c(lat, long)) |> 
  relocate(bloom_doy, country, location, alt, year) |> 
  arrange(country, location, year)

write_csv(cherry_months, file = "data/cherry_months.csv")

cherry_months_2024 <- cherry_months |> 
  filter(!is.na(bloom_doy))

cherry_months_2025 <- cherry_months |> 
  filter(year == 2025)
```

## Exploratory data analysis (EDA)

### Linear model

A simple linear model can help us understand how predictive a feature is

Plot the average maximum temperature in winter `t2m_max_winter`

```{r}
#| label: fig-t2m_max_winter
#| fig-cap: "Average maximum winter temperature (Dec. 1st of the previous year until end of February)"

cherry_seasons |>
  ggplot(aes(x = year, y = t2m_max_winter)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum winter temperature (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```

We fit a linear model with different intercepts and time trends for each location and a weather effect. We can add interaction effects to the temperature data too if we want.

```{r}
#| label: fit-lin-t2m_max_winter

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:t2m_max_winter, 
    data = cherry_seasons
  )
```

Create fitted values for the bast observations

```{r}
#| label: predictions-t2m_max_winter

mean_pred <- predict(object = lm_fit, new_data = cherry_seasons)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-t2m_max_winter
#| fig-cap: "Predictions from linear time trend and average maximum winter temperature"

cherry_seasons |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

Because `t2m_max_winter` shows a clear upwards trend over the past twenty years, we are able model the trend towards earlier peak bloom dates quite well.

We see that the model with the linear time trend and the predictor `t2m_max_winter` does predict the data better than with only the linear time trend. We can assume that a model with `t2m_max_spring` would explain the data even better, but unfortunately, we don't have data for the spring of this year.

We also see that the average temperature in Kyoto is too mild to explain much of the variance in peak bloom times.

### Graphical

Plot a linear time trend

```{r}
cherry_seasons |>
  ggplot(aes(x = year, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

Plot peak dates vs aggregated growing degree days in spring

```{r}
cherry_seasons |>
  ggplot(aes(x = agdd_spring, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```


### Correlations

We use the `{corr}` package to check the correlations of the predictors.

```{r}
cherry_seasons |>
  correlate() |>
  gt() |>
  fmt_number() |>
  tab_header(title = "correlations") |>
  tab_footnote(
  footnote = ""
)
```

### Decision tree

Fit a decision tree for exploratory data analysis. We use the `{parsnip}` package to create a model that runs on the `"rpart"` engine, and we use the `rpart.plot` package to plot the decision tree.

```{r}
set.seed(456)

# Create a {parsnip} model that runs on the "rpart" engine
cart_specs <- decision_tree(
  tree_depth = 10,
  cost_complexity = 0.001
  ) |>
  set_mode("regression") |>
  set_engine("rpart")

# fit the model 
cart_fit <- cart_specs |>
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_days
  )

# Visualize with {rpart.plot}
cart_fit$fit |> 
  rpart.plot()
```

On a seasonal aggregation level, aggregated growing degree days (AGDD) are the most important predictor.

### Variable importance plot (VIP)

Use the `{vip}` package to create a variable importance plot (VIP)

```{r}
cart_fit |> 
  vip()
```

## Modeling

### Data spending

We need to split these observations randomly into
- a training set that will be used to train the model
- a test set that will be used to compare the performance different models against each other

Because the number of observations are so small, any random split will have huge consequences. Stratified splitting may somewhat help here.

We use the `{rsample}` function `initial_split()` or `initial_time_split()` to create a test set and a training set. The first let's you use the argument `strata = bloom_doy`, the latter will consider the time when splitting the data.

```{r}
#| label: initial-split
#| eval: true

set.seed(123)

cherry_data <- cherry_months_2024

data_split <- initial_split(data = cherry_data, prop = 2/3)

data_train <- training(data_split)
data_test  <- testing(data_split)

cherry_recipe <- recipe(formula = bloom_doy ~ ., data = data_train) |>
  step_dummy(all_nominal_predictors()) |> 
  step_harmonic(year, frequency = 1 / 11, cycle_size = 1) |> 
  step_zv(all_nominal_predictors()) |> 
  step_normalize(all_numeric_predictors())

folds <- vfold_cv(data_train, v = 10)
```

Set the model engines with the `{parsnip}` package.

```{r}
linear_mod <- linear_reg() |> 
  set_mode("regression") |> 
  set_engine("lm")

pca_tune_mod <- linear_reg() |> 
  set_mode("regression") |> 
  set_engine("lm")

ridge_tune_mod <- linear_reg(
  mixture = 0, 
  penalty = tune()
  ) |> 
  set_mode("regression") |> 
  set_engine("glmnet")

lasso_tune_mod <- linear_reg(
  mixture = 1, 
  penalty = tune()
  ) |> 
  set_mode("regression") |> 
  set_engine("glmnet")

rf_tune_mod <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
  ) |> 
  set_engine("ranger", num.threads = 6, importance = "impurity") |> 
  set_mode("regression")

xgb_tune_mod <- boost_tree(
  mtry = tune(),
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
  ) |>
  set_engine("xgboost", num.threads = 6) |> 
  set_mode("regression")

lgb_tune_mod <- boost_tree(
  mtry = tune(),
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
  ) |> 
  set_engine("lightgbm", num.threads = 6) |> 
  set_mode("regression")

cub_tune_mod <- cubist_rules(
  committees = tune(),
  neighbors = tune(),
  max_rules = tune()
  ) |>
  set_engine("Cubist") |>
  set_mode("regression")

ridge_grid <- grid_random(
  penalty(range = c(-10, 10)),
  size = 50
)

lasso_grid <- grid_random(
  penalty(range = c(-10, 10)),
  size = 50
)

rf_grid <- grid_random(
  trees(range = c(1000, 2000)),
  mtry(range = c(1, 5)),
  min_n(range = c(1, 10)),
  size = 50
)

xgb_grid <- grid_random(
  trees(range = c(1000, 2000)),
  tree_depth(range = c(1, 5)),
  learn_rate(range = c(0.1, 0.005)),
  mtry(range = c(1, 6)),
  loss_reduction(range = c(0, 0.1)),
  size = 50
)

lgb_grid <- grid_random(
  trees(range = c(1000, 2000)),
  tree_depth(range = c(1, 5)),
  learn_rate(range = c(0.1, 0.005)),
  mtry(range = c(1, 6)),
  loss_reduction(range = c(0, 0.1)),
  size = 50
)

cub_grid <- grid_random(
  committees(range = c(30, 45)),
  neighbors(range = c(1, 10)),
  max_rules(range = c(5, 30)),
  size = 50
)
```

Tune the models using the k-fold CV on the grid that we have just created using the `{tune}` function `tune_grid()`. Define the metrics with the `{yardstick}` function `metric_set()`, specifying `mae()`. To speed up the tuning process, we use the `tune_race_anova()` function from the `{finetune}` package to speed up the tuning process.

```{r}
ridge_tune_res <- tune_grid(
  object = workflow() |> 
    add_recipe(cherry_recipe) |> 
    add_model(ridge_tune_mod),
  resamples = folds, 
  grid = ridge_grid,
  control = control_race(),
  metrics = metric_set(mae)
)

lasso_tune_res <- tune_grid(
  object = workflow() |> 
    add_recipe(cherry_recipe) |> 
    add_model(lasso_tune_mod),
  resamples = folds, 
  grid = lasso_grid,
  control = control_race(),
  metrics = metric_set(mae)
)

rf_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(rf_tune_mod),
    resamples = folds,
    grid = rf_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

xgb_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(xgb_tune_mod),
    resamples = folds,
    grid = xgb_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

lgb_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lgb_tune_mod),
    resamples = folds,
    grid = lgb_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

cub_tune_res <- tune_race_anova(
  object = workflow() |>
    add_recipe(cherry_recipe) |>
    add_model(cub_tune_mod),
  resamples = folds,
  grid = cub_grid,
  control = control_race(),
  metrics = metric_set(mae)
)
```

Use `finalize_workflow()` to fit the models with the best hyperparameter combination as assessed using k-fold cross validation to the whole training set.

```{r}
final_linear_fit <- fit(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(linear_mod), 
  data = data_train
  )

final_ridge_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(ridge_tune_mod), 
      select_best(ridge_tune_res)
      ), 
  data = data_train
  )

final_lasso_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lasso_tune_mod), 
      select_best(lasso_tune_res)
      ), 
  data = data_train
  )

final_rf_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(rf_tune_mod), 
      select_best(rf_tune_res)
      ), 
  data = data_train
  )

final_xgb_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(xgb_tune_mod), 
      select_best(xgb_tune_res)
      ), 
  data = data_train
  )

final_lgb_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lgb_tune_mod), 
      select_best(lgb_tune_res)
      ), 
  data = data_train
  )

final_cub_fit <- fit(
  object = finalize_workflow(
    workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(cub_tune_mod),
    select_best(cub_tune_res)
    ),
  data = data_train
  )

final_linear_res <- predict(final_linear_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_ridge_res <- predict(final_ridge_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_lasso_res <- predict(final_lasso_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_rf_res <- predict(final_rf_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_xgb_res <- predict(final_xgb_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_lgb_res <- predict(final_lgb_fit, new_data = data_test) |> 
  bind_cols(data_test)

final_cub_res <- predict(final_cub_fit, new_data = data_test) |> 
  bind_cols(data_test)
```

Compute the mae for the "final" model. You can use the `{yardstick}` functions `mae()`.

```{r}
final_linear_mae <- mae(final_linear_res, truth = bloom_doy, estimate = .pred)

final_ridge_mae <- mae(final_ridge_res, truth = bloom_doy, estimate = .pred)

final_lasso_mae <- mae(final_lasso_res, truth = bloom_doy, estimate = .pred)

final_rf_mae <- mae(final_rf_res, truth = bloom_doy, estimate = .pred)

final_xgb_mae <- mae(final_xgb_res, truth = bloom_doy, estimate = .pred)

final_lgb_mae <- mae(final_lgb_res, truth = bloom_doy, estimate = .pred)

final_cub_mae <- mae(final_cub_res, truth = bloom_doy, estimate = .pred)

tribble(
  ~ model, ~ mae,
  "linear", final_linear_mae$.estimate,
  "ridge", final_ridge_mae$.estimate,
  "lasso", final_lasso_mae$.estimate,
  "rf", final_rf_mae$.estimate,
  "xgb", final_xgb_mae$.estimate,
  "lgb", final_lgb_mae$.estimate,
  "cub", final_cub_mae$.estimate
  ) |> 
  arrange(mae)
```


```{r}
vip(final_rf_fit)
vip(final_xgb_fit)
vip(final_lgb_fit)
```

### Stacking

```{r}
ctrl_grid <- control_stack_grid()
ctrl_res <- control_stack_resamples()

model_1_res <- extract_workflow(ridge_tune_res) |> 
  fit_resamples(folds, metrics = metric_set(mae))

model_2_res <- extract_workflow(lasso_tune_res) |> 
  fit_resamples(ctrl_res, metrics = metric_set(mae))

model_3_res <- extract_workflow(rf_tune_res) |> 
  fit_resamples(ctrl_res, metrics = metric_set(mae))

model_4_res <- extract_workflow(xgb_tune_res) |> 
  fit_resamples(ctrl_res, metrics = metric_set(mae))

model_5_res <- extract_workflow(lgb_tune_res) |> 
  fit_resamples(ctrl_res, metrics = metric_set(mae))

model_6_res <- extract_workflow(cub_tune_res) |> 
  fit_resamples(ctrl_res, metrics = metric_set(mae))

cherry_stack <- stacks() |> 
  add_candidates(ridge_tune_res) |> 
  add_candidates(lasso_tune_res) |> 
  add_candidates(rf_tune_res) |> 
  add_candidates(xgb_tune_res) |> 
  add_candidates(lgb_tune_res) |> 
  add_candidates(cub_tune_res) |> 
  blend_predictions() |> 
  fit_members()




boot_preds <- test_data %>%
  bootstrap(times = 1000) %>%
  mutate(pred = map(splits, ~ predict(ensemble_stack, analysis(.x)))) %>%
  unnest(pred)

pi_90 <- boot_preds %>%
  group_by(id) %>%
  summarize(
    lower = quantile(.pred, 0.05),
    upper = quantile(.pred, 0.95)
  )

ensemble_preds <- predict(ensemble_stack, new_data = test_data, type = "pred_int", level = 0.90)
```


### Forecast

```{r}
fcst <- predict(object = final_rf_fit, new_data = cherry_months_2025)
```


### Validation

The validation set is used to validate, and calibrate the models that were selected using k-fold CV and fitted on the whole training data set.

We use the `{probably}` package to add prediction intervals to our point forecasts.

Post-processing is important to handle any adjustments and bias for predictions.

We then use the calibrated models on our test dataset for a final evaluation.

Create the predictions for validation data

```{r}
rf_vpreds <- predict(rf_fit, validate)
lgb_vpreds <- predict(lgb_fit, validate)
cub_vpreds <- predict(cub_fit, validate)
```

Make a large dataframe with all predictions and the actual observations

```{r}
validate_preds <- validate |>
  cbind(
    rf = rf_vpreds$.pred,
    lgb = lgb_vpreds$.pred,
    cub = cub_vpreds$.pred
  ) |>
  relocate(year, location, bloom_doy, rf, lgb, cub) |> 
  pivot_longer(cols = c(rf, lgb, cub), names_to = "model", values_to = "value") |> 
  select(year, location, bloom_doy, model, value)
```

Visualize the certainty in predictions

```{r}
validate_preds |>
  ggplot(aes(x = bloom_doy, y = value, color = model)) +
  geom_point() +
  coord_equal() +
  geom_abline(lty = 2) +
  scale_x_continuous(limits = c(80, 110)) +
  scale_y_continuous(limits = c(80, 110)) +
  labs(
    title = "bloom_doy & predictions",
    x = "bloom_doy", y = "predictions"
  ) +
  theme_bw() +
  theme(legend.position = "top")
```

Quick error measurement

```{r}
validate_preds |>
  mutate(abs_error = abs(bloom_doy - value)) |>
  group_by(model) |>
  summarize(avg_abs_error = mean(abs_error)) |> 
  arrange(avg_abs_error)
```

### Calibration

The top performing models on validation are cubist and random forest

Cubist is already pretty precise and accurate but we will calibrate our random forest to see if we increase accuracy on the validation set.

Calibrate random forest with validation data

```{r}
calibrate_rf <- data.frame(
  bloom_doy = validate$bloom_doy,
  .pred = rf_vpreds$.pred
)

rf_fit_cal <- calibrate_rf |>
  cal_estimate_linear(bloom_doy)

calibrated_rf_preds <- calibrate_rf |>
  cal_apply(rf_fit_cal)

calibrated_rf_preds |>
  mutate(abs_error = abs(bloom_doy - .pred)) |>
  summarize(avg_abs_error = mean(abs_error))
```

There is an improvement but note we need to **test** the calibration

### Intervals

While we have optimized our point forecasts with pre-processing, tuning, and post-processing, it's also important to capture our model uncertainty and to create robust prediction intervals with the validation set.

Let's create confidence intervals for our models. The simplest form being is the conformal split method implemented in the `{probably}` function `int_conformal_split()`.

```{r}
rf_int_model <- probably::int_conformal_split(rf_fit, validate)
lgb_int_model <- probably::int_conformal_split(lgb_fit, validate)
cub_int_model <- probably::int_conformal_split(cub_fit, validate)
```

We have obtained conformal interval models.

### Testing

We use the test data set to evaluate the performance of the models that we have trained on the test set and calibrated on the validation set.

Create the predictions and prediction intervals on the test set

```{r}
cubist_preds <- predict(cub_int_model, data_test, level = 0.90)

rf_cal_preds <- predict(rf_int_model, data_test, level = 0.90) |> 
  cal_apply(rf_fit_cal)

lgb_int_preds <- predict(lgb_int_model, data_test, level = 0.90)
```

Combine the test set and the predictions

```{r}
test_df <- data_test |>
  cbind(
    cubist = cubist_preds$.pred,
    cubist_upper = cubist_preds$.pred_upper,
    cubist_lower = cubist_preds$.pred_lower,
    rf_cal = rf_cal_preds$.pred,
    rf_cal_upper = rf_cal_preds$.pred_upper,
    rf_cal_lower = rf_cal_preds$.pred_lower,
    lgb = lgb_int_preds$.pred,
    lgb_upper = lgb_int_preds$.pred_upper,
    lgb_lower = lgb_int_preds$.pred_lower
  ) |> 
  select(year, location, bloom_doy, cubist, cubist_upper, cubist_lower, rf_cal, rf_cal_upper, rf_cal_lower, lgb, lgb_upper, lgb_lower)
```


```{r}
test_df |>
  ggplot() +
  geom_point(aes(x = bloom_doy, y = bloom_doy, color = "actual")) +
  geom_errorbar(aes(xmin = lgb_lower, xmax = lgb_upper, y = bloom_doy, color = "lgb_interval"), alpha = 0.4) +
  geom_point(aes(x = cubist, y = bloom_doy, color = "cubist")) +
  geom_point(aes(x = rf_cal, y = bloom_doy, color = "rf_calibrated")) +
  coord_equal() +
  scale_x_continuous(limits = c(80, 110)) +
  scale_y_continuous(limits = c(80, 110)) +
  geom_abline(lty = 2) +
  labs(x = "predictions", y = "bloom_doy") +
  theme_bw()
```
Test accuracy

```{r}
test_df |>
  mutate(
    cub_abs_error = abs(bloom_doy - cubist),
    rf_abs_error = abs(bloom_doy - rf_cal),
    lgb_coverage = ifelse(bloom_doy >= lgb_lower & bloom_doy <= lgb_upper, 1, 0),
    n = row_number()
  ) |>
  summarize(
    cub_total_abs_error = sum(cub_abs_error),
    rf_total_abs_error = sum(rf_abs_error),
    lgb_coverage_pct = sum(lgb_coverage) / max(n)
  )
```

## Submission

To submit your entries, enter the predicted bloom dates from your models in the **submission form**.

```{r}
#| label: tbl-predictions-for-submission
#| tbl-cap: "Predictions for Submission"

predictions |>
  filter(year == 2025) |>
  mutate(prediction_date = strptime(paste(year, prediction), "%Y %j") |> as_date()) |>
  kable() |> 
  kable_styling()
```

**NOTE**: The question is if we need to round the point estimates or if we can submit partial days.

## Time trend

We can explore models without external predictors that rely solely on past observations and on time trends.

A simple method to predict peak bloom dates is to fit an **ordinary least-squares (OLS)** regression on the observed years and extrapolate the regression function into future. We want to have a separate line for each location, hence we tell R to estimate *interaction* effects. Note that R will automatically exclude `NA`s from the regression.

We only use data **from 1880** to fit the trends, as prior data may not be as reliable/relevant.

```{r}
#| label: fit-linear-time-trend

cherry_ls <- cherry |> 
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc"))

ls_fit <- lm(
  formula = bloom_doy ~ 0 + location + location:year,
  data = cherry_ls, 
  subset = year >= 1880
)
```

This simple linear regression functions suggest a trend toward earlier peak bloom at the sites.

```{r}
#| label: summary-linear-model

summary(ls_fit)
```

Unsurprisingly, neither the intercept nor the time trend are statistically significant for Vancouver and New York City, where we have few observations.

We can compute predictions for both the historical observations and for the current year with the `predict()` function. We first use the `expand_grid()` function to create all possible combinations for the years between 1880 and 2025 and the five locations, so that we have the input variables for both the fitted values for the past years and the prediction for the current year

```{r}
#| label: create-grid

cherry_grid <- expand_grid(location = unique(cherry_ls$location), year = 1880:2025)
```

Then we use these input variables to create both point estimates and 90% confidence intervals with `predict()`

```{r}
#| label: create-predictions

predictions <- predict(
  object = ls_fit, newdata = cherry_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(cherry_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  relocate(year, location)
```

Plot only the fitted values, i.e., the predictions for the actual observations between 2015 and 2024

```{r}
#| label: fig-fitted
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from simple linear regression models fitted to
#|   five sites.

cherry_ls |>
  right_join(predictions, by = c("year", "location")) |>
  filter((location == "vancouver" & year >= 2015) | (location != "vancouver" & year >= 2015)) |>
  filter((location == "newyorkcity" & year >= 2019) | (location != "newyorkcity" & year >= 2015)) |> 
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that the slope of the time trend has become steeper since 2015. If we assume that this trend continues, our prediction for the current year will be too high.

We create a function called `day_to_date()` that converts `bloom_doy` into an ISO date

```{r}
#| label: day_to_date

#' Small helper function to convert the day of year to the actual date.
#'
#' @param year year as an integer
#' @param doy day of the year as integer (1 means January 1st)
#' @return date string

doy_to_date <- function(year, doy) {
  strptime(paste(year, doy, sep = "-"), "%Y-%j") |> # create date object
    strftime("%Y-%m-%d") # translate back to date string in ISO 8601 format
}
```


Based on this simple model, the peak bloom dates at these sites in 2025 are predicted to occur on.

```{r}
#| label: tbl-predictions
#| tbl-cap: "Predicted peak bloom dates at the four sites"

predictions |>
  filter(year == 2025) |>
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date) |> 
  gt()
```

### Averaging

We have only a few observations for Vancouver, BC and for New York City. Although the climate in Vancouver and NYC is different from the other locations, the simplest way to borrow information from the other locations is to average across these three sites.

Hence, we want to fit a straight line through the peak bloom dates, ignoring the actual site. That is, we fit a simple linear time trend on all available observations, but we want to use larger **weights** for observations from the target sites.

#### Vancouver

To run **weighted least squares (WLS)**, the `weights` inside `stats::lm()` argument must be a vector of the same length as the input data and we can use the names of the `location` column to create it. 

We could have created a new variable explicitly with:

```{r}
#| eval: false

cherry$weight <- (cherry_ls$location == "vancouver") + 0.2 * (cherry_ls$location != "vancouver")

cherry$weight <- NULL
```

So the weight put on observations from Vancouver are while observations from other locations receive a weight of only one fifth.

```{r}
#| label: fit-wls-vancouver

ls_fit_for_van <- lm(
  formula = bloom_doy ~ year,
  data = cherry_ls, subset = year >= 1880,
  weights = (location == "vancouver") + 0.2 * (location != "vancouver")
)
```

Create a grid with the input variables for Vancouver only

```{r}
#| label: grid-vancouver

vancouver_grid <- tibble(location = "vancouver", year = 2008:2025)
```

Then we use these inputs to create predictions for Vancouver, 

```{r}
#| label: predict-vancouver

predictions_vancouver <- predict(
  object = ls_fit_for_van, newdata = vancouver_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(vancouver_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date)
```

Not surprisingly, the predicted peak bloom dates for Vancouver are now very similar to the other 3 sites:

```{r}
#| label: tbl-predictions-vancouver
#| tbl-cap: "Predictions for Vancouver"

predictions_vancouver |> 
  gt()
```

Due to **rounding** and the **slope** of the time trend being very flat for such a long time series, the predictions are almost the same for every year.

We can plot the fitted values against the actual data from previous years

```{r}
#| label: fig-fitted-vancouver
#| warning: false
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for Vancouver using data from all four sites.

cherry_ls |>
  right_join(y = predictions_vancouver, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2008:2025) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that without including weather data, we are not able to capture the unusually warm springs in Vancouver, such as the one from 2010.

We have made this very cautious prediction since we don't know much about the conditions in Vancouver.

If we were satisfied with these weighted predictions for Vancouver, we could use them to replace the predictions that were made using only data from Vancouver.

```{r}
#| label: replace-vancouver-predictions

predictions <- predictions |>
  filter(location != "vancouver") |>
  bind_rows(predictions_vancouver)
```

#### New York

We once again fit a model on the whole data set, weighting observations from New York City with one and observations from other locations with one fifth.

```{r}
#| label: wls-for-nyc

ls_fit_for_nyc <- lm(
  formula = bloom_doy ~ year,
  data = cherry_ls, subset = year >= 1880,
  weights = (location == "newyorkcity") + 0.2 * (location != "newyorkcity")
)
```

Add inputs since 2019 and the current year

```{r}
#| label: grid-nyc

nyc_grid <- tibble(location = "newyorkcity", year = 2019:2025)
```

Create the fitted values for NYC

```{r}
#| label: fitted-nyc

predictions_nyc <- predict(
  object = ls_fit_for_nyc, newdata = nyc_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(nyc_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date)
```

Show the fitted values for New York City

```{r}
#| label: tbl-predictions-nyc
#| tbl-cap: "Predictions for New York City"

predictions_nyc |> 
  gt()
```

Due to rounding, and because the slope of the linear time trend being very flat for such a long time series, the predictions are the same for every year.

We plot the fitted values for New York City and the actual observations

```{r}
#| label: fig-fitted-nyc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for NYC using data from all five sites.

cherry_ls |>
  right_join(predictions_nyc, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2019:2024) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

Again, this model does not capture the steepness of the slope of the time trend accurately.

If we were satisfied with the averaged predictions we could replace the predictions that were made using only data from New York City

```{r}
#| label: replace-nyc-predictions

predictions <- predictions |>
  filter(location != "newyorkcity") |>
  bind_rows(predictions_nyc)
```