---
title: "Peak Bloom Prediction"
author: "econmaett"
lang: en-US
format:
  html:
    embed-resources: true
fig-cap-location: bottom
tbl-cap-location: bottom
toc: true
editor: source
---

Learn more about phenology at <https://en.wikipedia.org/wiki/Phenology>.

## Packages

We start by loading all the packages necessary to run this analysis

```{r}
#| label: load-pkgs
#| output: false

invisible(Sys.setlocale("LC_ALL","English"))

library(tidymodels)
library(tidyverse)
library(bonsai)
library(rules)
library(finetune)
library(probably)
library(kernlab)
library(corrr)
library(rpart.plot)
library(vip)
library(gt)
library(nasapower)
library(openmeteo)

tidymodels_prefer()
```

## Data

We begin by loading the data necessary for this analysis

### Clean data

The data for the five sites is provided as a simple text file in CSV format. Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site. The six columns in each data file are:
- `location`: a human-readable location identifier (`string`).
- `lat`: (approximate) latitude of the cherry trees (`double`).
- `long`: (approximate) longitude of the cherry trees (`double`).
- `alt`: (approximate) altitude of the cherry trees (`double`).
- `year`: year of the observation (`integer`).
- `bloom_date`: date of peak bloom of the cherry trees ([ISO 8601](https://xkcd.com/1179/) date `string`). The "peak bloom date" may be defined differently for different sites
- `bloom_doy`: days since January 1st until peak bloom (`integer`). January 1st is `1`.

#### Kyoto

The file `kyoto.csv` has been obtained from Yasuyuki Aono (`aono(at)envi.osakafu-u.ac.jp`) at <http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/>.

The geographical location at `35.01198°N` (latitude), `135.6761°E` (longitude), `44` meters above sea levels (altitude) roughly corresponds to the **Nakanoshima** area of the **Arashiyama Park** in Kyoto, Japan.

The peak bloom date of the **Prunus jamasakura (Japanese mountain cherry)** is determined by a local news paper in Arashiyama (Kyoto, Japan). Data prior to 1888 is extracted from various descriptions or estimated:
- Data from the 9th to the 14th centuries was acquired and analyzed by Aono and Saito (2010; International Journal of Biometeorology, 54, 211-219).
- Phenology for 15th to 21st centuries was acquired and analyzed by Aono and Kazui (2008; International Journal of Climatology, 28, 905-914).

Note that the data set for Kyoto includes implicit missing values (`NA`s). [Chapter 18 "Missing Values"](https://r4ds.hadley.nz/missing-values) of Hadley Wickham's book "R for Data Science (2e)" describes strategies for dealing with implicit `NA`s.

```{r}
#| label: load-kyoto
#| output: false

kyoto <- read_csv("data/kyoto.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  fill(location, lat, long, alt, .direction = "downup")
```

Plot the time series for Kyoto

```{r}
#| label: fig-kyoto
#| fig-cap: "Peak cherry blossom days in Kyoto, Japan"
#| warning: false

fig_kyoto <- kyoto |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.5) +
  geom_point(color = "#ff0054", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(800, 2000, by = 100), NA)) +
  labs(
    title = "Peak cherry blossom days in Kyoto, Japan", 
    subtitle = "Cultivar: Prunus jamasakura (Japanese mountain cherry)",
    x = NULL, y = NULL,
    caption = "Source: Yasuyuki Aono"
  ) +
  theme_light()

print(fig_kyoto)

ggsave(filename = "figures/fig_kyoto.png", plot = fig_kyoto, bg = "white")
```

Here we have used **locally estimated scatterplot smoothing (LOESS)** or **locally weighted scatterplot smoothing (LOWESS)** to highlight how the average peak bloom day changed over the years. 

We could explicitly model this with a **local level (+ trend)** state-space model that we fit with the **Kalman filter and Smoother**, which also allows us to effectively deal with **missing observations**.

MeteoSwiss suggests using a **Gaussian Kernel** with a *20-year window* to extract the trend, which is equivalent to using a **low-pass filter** that shuts down all higher frequencies.

It also looks like the observations exhibit **heteroskedasticity**, that is, the variance changes over time, and it looks like the variance has become more narrow over time.

We can use **changepoint detection** methods to test for the changing time trend and the changing variance.

Assuming that the data in the past is correct, there do seem to be seasonal cycles, perhaps connected to solar activity or the tilt of the earth.

There is a clear trend towards earlier blooming, perhaps connected to anthropogenic climate change and urbanization. Studies attribute about half of the earlier blooming times in recent years to the **urban heat island effect**, and the other half to **climate change**.

Modelling the local level and the trend would be useful if we are to forecast a couple years into the future, since an individual prediction may be incorrect, but on average, they may be good. 

However, since we are only interested in forecasting a single moment in time, which may be greatly influenced by local weather, we probably are better off using historical weather data that we augment with forecasts, either by professional forecasters or by extrapolating the previous predictors.

#### Liestal

The file `liestal.csv` contains observations for **a single wild cherry (prunus avium) tree** in **Liestal-Weideli**, at `7.4814°N` (latitude), `7.730519°E` (longitude) and `350` meters above sea levels (altitude). 

```{r}
#| label: load-liestal
#| output: false

liestal <- read_csv("data/liestal.csv")
```

The **peak bloom date** is defined as the day when **25%** of the blossoms are in full bloom. The date is determined by [MeteoSwiss](https://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html). This means that 70% of buds would be open a some days later. This raises the question of whether we should model the locations separately.

The data is provided by **Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss**.

Let's plot the time series for Liestal, Switzerland:

```{r}
#| label: fig-liestal
#| fig-cap: "Peak cherry blossom days for Liestal-Weideli, Switzerland"
#| warning: false

fig_liestal <- liestal |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Liestal-Weideli, Switzerland",
    subtitle = "Cultivar: Wild cherry (prunus avium)",
    x = NULL, y = NULL,
    caption = "Source: Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss"
  ) +
  theme_light()

print(fig_liestal)

ggsave(filename = "figures/fig_liestal.png", plot = fig_liestal, bg = "white")
```

For Liestal you can see the impact of the infamous heat summer of 2003 which I remember from my childhood, although interestingly, there's another record in 1990.

#### New York

The file `newyorkcity.csv` contains **only a single observation** for the year 2024. It concerns the **Yoshino cherry (prunus × yedoensis)** tree in **Washington Square Park**, New York City, at `40.7304 °N` (latitude), `-73.99809°E` (longitude) and `8.5` meters above sea level (altitude).

```{r}
#| label: load-newyorkcity
#| output: false

newyorkcity <- read_csv("data/nyc.csv")
```

Luckily, there are some historical observations dating *back to 2019* in the data provided by **USA-NPN**. The Washington Square Park has site id `32789` and the **Yoshino** cherry you should predict has species id `228`.

```{r}
#| label: load-historical-data-nyc
#| output: false

nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789, Species_ID == 228) |> 
  mutate(Observation_Date = mdy(Observation_Date))
```

This file now contains individual observations of the phenophase and not the actual peak bloom date. We use the variable `Intensity_Value` which is either `-9999`, `95% or more`, `Less than 5%`, `25-49%`, `75-94%`, `50-74%`. Since according to the variable `Individual_ID`, only one tree, `183140` is being observed, we can use the first date where `Intensity_Value` is `50-74%` as our `bloom_doy`.

```{r}
#| label: clean-npn-nyc-data
#| warning: true

nyc_data <- nyc_data_npn |>
  arrange(Observation_Date) |>
  mutate(year = year(Observation_Date)) |>
  group_by(year) |>
  summarize(
    first_flower_index = min(which((Intensity_Value %in% c("75-94%", "95% or more")))),
    bloom_date = as_date(strftime(Observation_Date[first_flower_index], format = "%Y-%m-%d")),
    bloom_doy = Day_of_Year[first_flower_index],
    .groups = "drop"
  ) |>
  filter(!is.na(bloom_doy)) |>
  select(-first_flower_index) |>
  mutate(location = "newyorkcity") |> 
  arrange(year)
```

Then we combine the data sets. Make sure not to add any dates twice.

```{r}
#| label: augment-nyc-data

newyorkcity <- newyorkcity |>
  bind_rows(nyc_data |> filter(year != 2024)) |>  
  fill(lat, long, alt)
```

We now have a total of four observations for New York City

```{r}
#| label: tbl-nyc-data
#| tbl-cap: "Augmented NYC observations"

newyorkcity |> 
  filter(location == "newyorkcity") |> 
  gt()
```

Plot the time series for New York City, USA

```{r}
#| label: fig-newyorkcity
#| fig-cap: "Peak cherry blossom days for New York City, USA"
#| warning: false

fig_newyorkcity <- newyorkcity |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  labs(
    title = "Peak cherry blossom days for New York City, USA", 
    subtitle = "Cultivar: Yoshino cherry (prunus × yedoensis)",
    x = NULL, y = NULL,
    caption = "Source: Geroge Mason University and USA-NPN"
  ) +
  theme_light()

print(fig_newyorkcity)

ggsave(filename = "figures/fig_newyorkcity.png", plot = fig_newyorkcity, bg = "white")
```


#### Vancouver

The file `vancouver.csv` contains observations for cherry trees of the cultivar  **Akebono (daybreak cherry)** in **Maple Grove Park**, Vancouver, BC located approximately at `49.2236916°N` (latitude), `-123.1636251°E` (longitude), `24` meters above sea levels (altitude).

Akebono (daybreak cherry). A medium sized tree growing 8 m × 8 m (25 ft) with a stiff, upright-spreading crown, eventually becoming umbrella shaped. ‘Akebono’ flowers in late March or early April, usually following the purple-leaf plums by about a week. Flowers are produced abundantly, shell-pink fading to nearly white. ‘Akebono’ is a seedling of Prunus yedoensis ‘Somei-yoshino‘(Tokyo cherry) selected in 1925 at a California nursery. ‘Akebono’ is by far the most commonly planted early spring flowering cherry in this area, and is celebrated for its essentially rainproof flowers and freedom from disease. Autumn colour is yellow to pumpkin orange. <https://vcbf.ca/cherry_cultivars/akebono/>.

```{r}
#| label: load-vancouver
#| output: false

vancouver <- read_csv("data/vancouver.csv")
```

Unfortunately, there are **only three observations** for 2022, 2023, and 2024, respectively. But casual observations have been recorded in the form of **photos posted** to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/). You can **search the forum** for the keywords `"Akebono"` (i.e., the name of the cultivar) and `"Maple Grove Park"` (i.e., the location of the trees).

After searching through the web forum, I have decided to add the following observations to the data set:

```{r}
#| label: define-vancouver_forum

vancouver_forum <- tribble(
  ~ year, ~ bloom_date,
  2021, "2021-03-31", # Start: 2021-03-21, Full: 2021-03-31, Full: 2021-04-07
  2020, "2020-04-03", # Full bloom
  2018, "2018-03-18", # Full bloom
  2017, "2017-04-11", # Full bloom 
  2015, "2015-03-19", # Full bloom
  2011, "2011-04-01", # Full bloom was over on 2011-04-15
  2010, "2010-03-14", # 60% bloom on 2010-03-07
  2008, "2008-04-14" # Start: 2008-04-06, Full: 2008-04-14, finished: 2008-04-22
  ) |> 
  mutate(
    bloom_date = as_date(bloom_date),
    bloom_doy = as.numeric(1 + as_date(bloom_date) - as_date(paste0(year, "-01-01"))),
    location = "vancouver"
  )
```

We can do some additional plausibility checks. For example, spring 2010 was indeed unusually warm, [causing cherry blossoms to bloom during the Vancouver 2010 Winter Olympics](https://www.upi.com/News_Photos/view/upi/33393f9d20159ecb347de9aa1d670922/2010-Winter-Olympics-in-Vancouver/). National Geographic claimed it was the [warmest Olympics to date](https://www.nationalgeographic.com/adventure/article/100212-vancouver-2010-warmest-winter-olympics).

Once we are statsfied with the plausibility, we can add these observations to the data set

```{r}
#| label: augment-vancouver

vancouver <- vancouver |> 
  bind_rows(vancouver_forum) |> 
  fill(lat, long, alt) |> 
  arrange(year)
```

Now we have a total of eleven observations for Vancouver, BC.

```{r}
#| label: tbl-augmented-vancouver-data
#| tbl-cap: "Augmented Vancouver data"

vancouver |> 
  filter(location == "vancouver") |> 
  gt()
```

Plot the data for Vancouver BC, Canada

```{r}
#| label: fig-vancouver
#| fig-cap: "Peak cherry blossom days for Vancouver BC, Canada"
#| warning: false

fig_vancouver <- vancouver |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  # scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Vancouver BC, Canada", 
    subtitle = "Cultivar: Akebono (daybreak cherry)",
    x = NULL, y = NULL,
    caption = "Source: Geroge Mason University and VCBF Neighbourhood Blog"
  ) +
  theme_light()

print(fig_vancouver)

ggsave(filename = "figures/fig_vancouver.png", plot = fig_vancouver, bg = "white")
```


#### Washington

The file `washingtondc.csv` has been obtained from the EPA's Climate Change Indicators in the United States <https://www.epa.gov/climate-indicators/cherry-blossoms>.

```{r}
#| label: load-washingtondc
#| output: false

washingtondc <- read_csv("data/washingtondc.csv")
```

The geographical location at `38.88535°N` (latitude), `-77.03863°E` (longitude) and `0` meters above sea level (altitude) correspond to the location of the [Tidal Basin in Washington, DC](https://www.nps.gov/articles/dctidalbasin.htm).

The peak bloom date is defined as the day when **70%** of the **Yoshino cherry (prunus x yedoensis)** are in full bloom, as determined by the [National Park Service](https://www.nps.gov/subjects/cherryblossom/bloom-watch.htm).

Let's plot the data for Washington D.C

```{r}
#| label: fig-washingtondc
#| fig-cap: "Peak cherry blossom days for Washington DC, USA"
#| warning: false

fig_washingtondc <- washingtondc |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#a53860", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Washington DC, USA",
    subtitle = "Cultivar: Yoshino cherry (prunus × yedoensis )",
    x = NULL, y = NULL,
    caption = "Source: Environmental Protection Agency (EPA) Climate Change Indicators in the United States"
  ) +
  theme_light()

print(fig_washingtondc)

ggsave(filename = "figures/fig_washingtondc.png", plot = fig_washingtondc, bg = "white")
```

Now that we have loaded the datasets for the five locations, we can combine them into a single CSV file called `cherry.csv`.

```{r}
#| label: combine-cherry-csv
#| eval: false

cherry <- bind_rows(kyoto, liestal, newyorkcity, vancouver, washingtondc)

write_csv(cherry, file = "data/cherry.csv")
```

Now we can load the `cherry.csv` file from our `data/` folder

```{r}
#| label: load-cherry-csv
#| output: false

cherry <- read_csv("data/cherry.csv")
```

We can show the last three observation for each location

```{r}
#| label: tbl-01
#| tbl-cap: "latest 3 observations for each location"

cherry |> 
  group_by(location) |> 
  slice_tail(n = 3) |> 
  gt()
```

We can visualize the time series for our five main locations next to each other. Notice the use of the `{stringr}` function `str_to_title()` to convert the first letter of the locations to uppercase.

```{r}
#| label: fig-cherries_1880
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at five different sites.

df_cherry_1880 <- cherry |> 
  filter(year >= 1880)

fig_cherries <- ggplot(data = df_cherry_1880, aes(year, bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab") +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 40), NA)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(
    title = "Peak bloom (since Jan 1st)",
    x = NULL, 
    y = NULL
  ) +
  theme_light()

print(fig_cherries)

ggsave(filename = "figures/fig_cherries.png", plot = fig_cherries, bg = "white")
```

Create summary statistics to get an overview over the data. Specifically, we are interested in the range of the `year` variable, the total number of observations at each location, and the mean, median, and standard deviation of the `bloom_doy` variable.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for the main data set"

cherry |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location) |> 
  summarise(
    min_year = min(year),
    max_year = max(year),
    n_obs = n(),
    mean = round(mean(bloom_doy)),
    median = round(median(bloom_doy)),
    sd = round(sd(bloom_doy))
  ) |> 
  gt()
```

We see that the standard deviation for Liestal is much larger than the rest, where peak bloom dates are on average about a weak apart from their historical average.

We check for missing observations in the most recent years:

```{r}
#| label: tbl-check-kyoto-missing
#| tbl-cap: "The most recent missing values in the Kyoto dataset"

cherry |> 
  filter(location == "kyoto") |> 
  filter(is.na(bloom_doy)) |> 
  tail() |> 
  gt()
```

Kyoto has multiple implicit `NA`s, including in 1730, 1872, 1895, 1919, 1921, and 1945. So if we use data since 1880, as the demo analysis suggests, we still have four `NA`s. For time series analysis, we usually assume equally-spaced observations, and therefore need to either cut the data to start in 1946, or to replace the `NA`s with a sensible value. The other four locations do not contain any implicit `NA`.

Plot the distribution of the peak bloom days for the locations:

```{r}
#| label: fig-histogram
#| fig-cap: "Histogram of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location)) +
  geom_histogram() +
  labs(title = "Histogram of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's use a density plot instead:

```{r}
#| label: fig-density-plot
#| fig-cap: "Denisty plots of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location, colour = location)) +
  geom_density(alpha = 0.5) +
  labs(title = "Denisty plots of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's plot all five time series in the same plot.

```{r}
#| label: fig-cherry-blossoms-combined
#| fig-cap: "Peak cherry blossom days for different locations"
#| warning: false

fig_combined <- cherry |> 
  filter(year >= 1880) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy, colour = location)) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  geom_step(linetype = "dotted", lwd = 0.5) +
  geom_point(alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(title = "Peak cherry blossom days by locations", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())

print(fig_combined)

ggsave(filename = "figures/fig_combined.png", plot = fig_combined, bg = "white")
```

This graphic highlights the negative linear time trend for all locations.

It also highlights that a linear time trend will likely underestimate the accelerating trend towards earlier peak bloom dates.

A simple method such as **ETS**, which places a lower weight on observations far in the past, may be better at modelling the recent trend.

In the M-comptetitions, a combination of ARIMA, which takes seasonality into account, and ETS, which takes the changing trend into account, usually fared well.

In any case, we should compare all our models against **benchmark models** such as the **naive model**, where the prediction is equal to the last observation, and a simple **AR(1)** model.

### Additional data

The competition organizers provide additional data sets for sites other than the main sites relevant for the competition. We can use these time series in your modeling to help with spatial and temporal extrapolation. For example, we can add predictors to these data sets and train models on additional sites.

#### Japan

The data file `japan.csv` is provided by the [Japanese Meteorological Agency (JMA)](https://www.data.jma.go.jp/sakura/data/pdf/005.pdf) and contains recorded peak bloom dates for cherry trees at various sites across Japan. The sample trees are located within a 5km radius of the location indicated in the data file. 

Load the data set, make implicit missing values explicit and change the `location` variable to only include the city names in lowercase letters.

```{r}
#| label: read-japan
#| output: false

japan <- read_csv("data/japan.csv") |>  
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Japan/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `japan.csv` file contains observations between `r min(japan$year)` and `r max(japan$year)` for `r length(unique(japan$location))` locations in Japan.

Let's plot peak bloom dates for Kyoto, Japan

```{r}
#| label: fig-bloom-kyoto2
#| fig-cap: "Peak bloom dates in Kyoto, Japan"
#| warning: false

fig_kyoto2 <- japan |>
  filter(location == "kyoto") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 10), NA)) +
  labs(
    title = "Peak bloom (since Jan 1st) in Kyoto, Japan",
    subtitle = NULL,
    x = NULL, y = NULL,
    caption = "Source: Japanese Meteorological Agency (JMA)"
  ) +
  theme_light()

print(fig_kyoto2)

ggsave(filename = "figures/fig_kyoto2.png", plot = fig_kyoto2, bg = "white")
```

#### Korea

The file `south_korea.csv` contains **first flowering dates** of cherry trees at various sites across South Korea, curated by the **Korean Meteorological Administration (KMA)**.

Because these are **first flowering dates**, they are likely to occur earlier than hypothetical full bloom dates would be.

Trees in South Korea were assumed to be of the **Prunus × yedoensis** variety based on previous work (Hur, Ahn, and Shim 2014; Chung et al. 2009) and its frequent occurrence.

Again, we make implicit missing values explicit, and I change the location names to lowercase so that to match the style of the main data set.

```{r}
#| label: load-korea
#| output: false

korea <- read_csv("data/south_korea.csv") |> 
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "South Korea/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `south_korea.csv` file contains observations between `r min(korea$year)` and `r max(korea$year)` at `r length(unique(korea$location))` locations in South Korea.

Let's plot the first bloom dates in Seoul, South Korea:

```{r}
#| label: fig-seoul
#| fig-cap: "First bloom in Seoul, South Korea"
#| warning: false

fig_seoul <- korea |>
  filter(location == "seoul") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1980, 2020, by = 5), NA)) +
  labs(
    title = "First bloom (since Jan 1st) in Seoul, Korea", 
    subtitle = "Cultivar: Prunus × yedoensis", 
    x = NULL, y = NULL, 
    caption = "Source: Korean Meteorological Administration (KMA)"
  ) +
  theme_light()

print(fig_seoul)

ggsave(filename = "figures/fig_seoul.png", plot = fig_seoul, bg = "white")
```

#### MeteoSwiss

The data file `meteoswiss.csv` contains peak bloom dates for various sites across Switzerland, obtained from <https://opendata.swiss/en/dataset/phanologische-beobachtungen>.

Note that when you use the `{utils}` function `read.csv()` you need to specify the option `encoding = "UTF-8"` to allow for special characters that are common in the German language.

```{r}
#| label: load-meteoswiss
#| output: false

meteoswiss <- read_csv("data/meteoswiss.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Switzerland/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `meteoswiss.csv` file contains observations between `r min(meteoswiss$year)` and `r max(meteoswiss$year)` at `r length(unique(meteoswiss$location))` locations.

Plot the data for Liestal, Switzerland. Remove the observation for the year 1952. Note that for this data set, peak bloom is defined as **50%** of blossoms being in bloom.

```{r}
#| label: fig-liestal2
#| fig-cap: "Peak bloom dates in Liestal, Switzerland"
#| warning: false

fig_liestal2 <- meteoswiss |>
  filter(location == "liestal", year != 1952) |>
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(
    limits = c(1985, NA), 
    breaks = c(seq(1985, 2020, by = 5), NA)
  ) +
  labs(
    title = "Peak bloom (since Jan 1st) in Liestal, Switzerland",
    subtitle = "Peak bloom defined as 50% of blossoms being in bloom",
    x = NULL, y = NULL,
    caption = "Source: MeteoSwiss"
  ) +
  theme_light()

print(fig_liestal2)

ggsave(filename = "figures/fig_liestal2.png", plot = fig_liestal2, bg = "white")
```


Plot the data for Zurich, Switzerland

```{r}
#| label: fig-zurich
#| fig-cap: "Peak bloom (since Jan 1st) in Zurich, Switzerland"
#| warning: false

fig_zurich <- meteoswiss |> 
  filter(str_detect(location, "zürich-")) |>
  mutate(location = str_remove(location, "zürich-")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 10), NA)) +
  labs(
    title = "Peak bloom (since Jan 1st) in Zurich, Switzerland",
    subtitle = "Peak bloom defined as 50% of blossoms being in bloom",
    x = NULL, y = NULL,
    caption = "Source: MeteoSwiss"
  ) +
  theme_light() +
  facet_grid(rows = vars(str_to_title(location)))

print(fig_zurich)

ggsave(filename = "figures/fig_zurich.png", plot = fig_zurich, bg = "white")
```

#### NPN

Two additional files are provided by the **USA National Phenology Network (USA-NPN)** and the many participants who contribute to its *Nature’s Notebook* program.

The file `USA-NPN_individual_phenometrics_data.csv`contains observations on **individual phenometrics** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-individual-phenometrics
#| output: false

npn_ind_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  mutate(Observation_Date = mdy(Observation_Date))

npn_ind_des <- read_csv("data/USA-NPN_individual_phenometrics_datafield_descriptions.csv")
```

The file `USA-NPN_status_intensity_data.csv` contains observations on the **status and intensity** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-status-intensity
#| output: false

npn_int_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv")

npn_int_des <- read_csv("data/USA-NPN_status_intensity_datafield_descriptions.csv")
```

These files can help augment the existing data sets for which you don't have many observations such as New York City (and perhaps even Vancouver). 

## Time trend

We can explore models without external predictors that rely solely on past observations and on time trends.

A simple method to predict peak bloom dates is to fit an **ordinary least-squares (OLS)** regression on the observed years and extrapolate the regression function into future. We want to have a separate line for each location, hence we tell R to estimate *interaction* effects. Note that R will automatically exclude `NA`s from the regression.

We only use data **from 1880** to fit the trends, as prior data may not be as reliable/relevant.

```{r}
#| label: fit-linear-time-trend

ls_fit <- lm(
  formula = bloom_doy ~ 0 + location + location:year,
  data = cherry, 
  subset = year >= 1880
)
```

This simple linear regression functions suggest a trend toward earlier peak bloom at the sites.

```{r}
#| label: summary-linear-model

summary(ls_fit)
```

Unsurprisingly, neither the intercept nor the time trend are statistically significant for Vancouver and New York City, where we have few observations.

We can compute predictions for both the historical observations and for the current year with the `predict()` function. We first use the `expand_grid()` function to create all possible combinations for the years between 1880 and 2025 and the five locations, so that we have the input variables for both the fitted values for the past years and the prediction for the current year

```{r}
#| label: create-grid

cherry_grid <- expand_grid(location = unique(cherry$location), year = 1880:2025)
```

Then we use these input variables to create both point estimates and 90% confidence intervals with `predict()`

```{r}
#| label: create-predictions

predictions <- predict(
  object = ls_fit, newdata = cherry_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(cherry_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  relocate(year, location)
```

Plot only the fitted values, i.e., the predictions for the actual observations between 2015 and 2024

```{r}
#| label: fig-fitted
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from simple linear regression models fitted to
#|   five sites.

cherry |>
  right_join(predictions, by = c("year", "location")) |>
  filter((location == "vancouver" & year >= 2015) | (location != "vancouver" & year >= 2015)) |>
  filter((location == "newyorkcity" & year >= 2019) | (location != "newyorkcity" & year >= 2015)) |> 
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that the slope of the time trend has become steeper since 2015. If we assume that this trend continues, our prediction for the current year will be too high.

We create a function called `day_to_date()` that converts `bloom_doy` into an ISO date

```{r}
#| label: day_to_date

#' Small helper function to convert the day of year to the actual date.
#'
#' @param year year as an integer
#' @param doy day of the year as integer (1 means January 1st)
#' @return date string

doy_to_date <- function(year, doy) {
  strptime(paste(year, doy, sep = "-"), "%Y-%j") |> # create date object
    strftime("%Y-%m-%d") # translate back to date string in ISO 8601 format
}
```


Based on this simple model, the peak bloom dates at these sites in 2025 are predicted to occur on.

```{r}
#| label: tbl-predictions
#| tbl-cap: "Predicted peak bloom dates at the four sites"

predictions |>
  filter(year == 2025) |>
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date) |> 
  gt()
```

### Averaging

We have only a few observations for Vancouver, BC and for New York City. Although the climate in Vancouver and NYC is different from the other locations, the simplest way to borrow information from the other locations is to average across these three sites.

Hence, we want to fit a straight line through the peak bloom dates, ignoring the actual site. That is, we fit a simple linear time trend on all available observations, but we want to use larger **weights** for observations from the target sites.

#### Vancouver

To run **weighted least squares (WLS)**, the `weights` inside `stats::lm()` argument must be a vector of the same length as the input data and we can use the names of the `location` column to create it. 

We could have created a new variable explicitly with:

```{r}
#| eval: false

cherry$weight <- (cherry$location == "vancouver") + 0.2 * (cherry$location != "vancouver")

cherry$weight <- NULL
```

So the weight put on observations from Vancouver are while observations from other locations receive a weight of only one fifth.

```{r}
#| label: fit-wls-vancouver

ls_fit_for_van <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "vancouver") + 0.2 * (location != "vancouver")
)
```

Create a grid with the input variables for Vancouver only

```{r}
#| label: grid-vancouver

vancouver_grid <- tibble(location = "vancouver", year = 2008:2025)
```

Then we use these inputs to create predictions for Vancouver, 

```{r}
#| label: predict-vancouver

predictions_vancouver <- predict(
  object = ls_fit_for_van, newdata = vancouver_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(vancouver_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date)
```

Not surprisingly, the predicted peak bloom dates for Vancouver are now very similar to the other 3 sites:

```{r}
#| label: tbl-predictions-vancouver
#| tbl-cap: "Predictions for Vancouver"

predictions_vancouver |> 
  gt()
```

Due to **rounding** and the **slope** of the time trend being very flat for such a long time series, the predictions are almost the same for every year.

We can plot the fitted values against the actual data from previous years

```{r}
#| label: fig-fitted-vancouver
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for Vancouver using data from all four sites.

cherry |>
  right_join(y = predictions_vancouver, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2008:2025) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that without including weather data, we are not able to capture the unusually warm springs in Vancouver, such as the one from 2010.

We have made this very cautious prediction since we don't know much about the conditions in Vancouver.

If we were satisfied with these weighted predictions for Vancouver, we could use them to replace the predictions that were made using only data from Vancouver.

```{r}
#| label: replace-vancouver-predictions

predictions <- predictions |>
  filter(location != "vancouver") |>
  bind_rows(predictions_vancouver)
```

#### New York

We once again fit a model on the whole data set, weighting observations from New York City with one and observations from other locations with one fifth.

```{r}
#| label: wls-for-nyc

ls_fit_for_nyc <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "newyorkcity") + 0.2 * (location != "newyorkcity")
)
```

Add inputs since 2019 and the current year

```{r}
#| label: grid-nyc

nyc_grid <- tibble(location = "newyorkcity", year = 2019:2025)
```

Create the fitted values for NYC

```{r}
#| label: fitted-nyc

predictions_nyc <- predict(
  object = ls_fit_for_nyc, newdata = nyc_grid,
  interval = "prediction", level = 0.9
  ) |> 
  bind_cols(nyc_grid) |> 
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(across(c(prediction, lower, upper), round)) |> 
  mutate(prediction_date = doy_to_date(year, prediction)) |> 
  relocate(year, location, prediction_date)
```

Show the fitted values for New York City

```{r}
#| label: tbl-predictions-nyc
#| tbl-cap: "Predictions for New York City"

predictions_nyc |> 
  gt()
```

Due to rounding, and because the slope of the linear time trend being very flat for such a long time series, the predictions are the same for every year.

We plot the fitted values for New York City and the actual observations

```{r}
#| label: fig-fitted-nyc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for NYC using data from all five sites.

cherry |>
  right_join(predictions_nyc, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2019:2024) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

Again, this model does not capture the steepness of the slope of the time trend accurately.

If we were satisfied with the averaged predictions we could replace the predictions that were made using only data from New York City

```{r}
#| label: replace-nyc-predictions

predictions <- predictions |>
  filter(location != "newyorkcity") |>
  bind_rows(predictions_nyc)
```

## Predictors

We have seen that a linear time trend alone is not sufficient to explain the variance in the peak bloom times. It therefore makes sense to include external data on local weather conditions in our model. 

### POWER

We use the `{nasapower}` package described at <https://docs.ropensci.org/nasapower/> to access the [NASA Prediction Of Worldwide Energy Resources (POWER) API](https://power.larc.nasa.gov/). We create a custom function called `get_power()` to repeatedly query the API with the `nasapower::get_power()` function.

```{r}
#| label: get_power

get_power <- function(vars, start = "1981-01-01", end = as.character(Sys.Date()), freq = "daily", lat, long) {
  
  power <- map2(
      .x = long,
      .y = lat,
      .f = function(long = .x, lat = .y) {
        res <- nasapower::get_power(
          community = "ag",
          lonlat = c(long, lat),
          pars = vars,
          dates = c(start, end),
          temporal_api = freq
        )
        return(res)
      }
    ) |>
    list_rbind()
  
  return(power)
}
```


We load the following variables:

- `T2M`: Temperature at 2 Meters. The average air (dry bulb) temperature at 2 meters above the surface of the earth.

- `T2M_MAX:` Temperature at 2 Meters Maximum. The maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `T2M_MIN`: Temperature at 2 Meters Minimum. The minimum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `PRECTOTCORR`: Precipitation Corrected. The average MERRA-2 bias corrected total precipitation at the surface of the earth.

- `RH2M`: Relative Humidity at 2 Meters. The ratio of vapor pressure to the saturation vapor pressure with respect to a plane surface of pure water, expressed in percent.

- `PS`: Surface pressure (kPa)

- `WS10M`: Wind Speed at 10 Meters. The average of wind speed at 10 meters above the surface of the earth.

```{r}
#| label: specify-vars

vars <- c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "PRECTOTCORR",
  "RH2M",
  "PS",
  "WS10M"
)
```

We retrieve the **coordinates** which we will use to match to the nearest weather stations

```{r}
#| label: get-coordinates

crd_kyoto        <- distinct(kyoto, location, lat, long)
crd_liestal      <- distinct(liestal, location, lat, long)
crd_nyc          <- distinct(newyorkcity, location, lat, long)
crd_vancouver    <- distinct(vancouver, location, lat, long)
crd_washingtondc <- distinct(washingtondc, location, lat, long)

crd_japan      <- distinct(japan, location, lat, long)
crd_korea      <- distinct(korea, location, lat, long)
crd_meteoswiss <- distinct(meteoswiss, location, lat, long)
```

Load the predictors for the five main locations

```{r}
#| label: load-main-nasa-power
#| eval: false

power_kyoto <- get_power(vars, lat = crd_kyoto$lat, long = crd_kyoto$long)
power_kyoto$location <- "kyoto"

power_liestal <- get_power(vars, lat = crd_liestal$lat, long = crd_liestal$long)
power_liestal$location <- "liestal"

power_nyc <- get_power(vars, lat = crd_nyc$lat, long = crd_nyc$long)
power_nyc$location <- "newyorkcity"

power_vancouver <- get_power(vars, lat = crd_vancouver$lat, long = crd_vancouver$long)
power_vancouver$location <- "vancouver"

power_washingtondc <- get_power(vars, lat = crd_washingtondc$lat, long = crd_washingtondc$long)
power_washingtondc$location <- "washingtondc"

power_cherry <- bind_rows(
  power_kyoto,
  power_liestal,
  power_nyc,
  power_vancouver,
  power_washingtondc
)

write_csv(power_cherry, file = "data/power_cherry.csv")
```

We also load predictors for the additional data sets. Because these take a long time to load, instead load data from a previous run.

```{r}
#| label: load-additional-nasa-power
#| eval: false

power_japan <- get_power(vars, lat = crd_japan$lat, long = crd_japan$long)
write_csv(power_japan, file = "data/power_japan.csv")

power_korea <- get_power(vars, lat = crd_korea$lat, long = crd_korea$long)
write_csv(power_korea, file = "data/power_korea.csv")

power_meteoswiss <- get_power(vars, lat = crd_meteoswiss$lat, long = crd_meteoswiss$long)
write_csv(power_meteoswiss, file = "data/power_meteoswiss.csv")

power_npn <- get_power(vars, lat = crd_npn$lat, long = crd_npn$long)
write_csv(power_npn, file = "data/power_npn.csv")
```

Let's have a look at the NASA POWER dataset for the five locations. We change some of the column names and convert the variable names to lowercase letters.

```{r}
#| label: load-power_cherry
#| output: false

power_cherry <- read_csv("data/power_cherry.csv") |> 
  rename_with(.fn = tolower, .cols = everything()) |> 
  rename(date = yyyymmdd, prcp = prectotcorr) |> 
  select(-c(lat, lon, year, mm, dd, doy)) |> 
  relocate(location, date)
```

## Open Meteo

Use the [`{openmeteo}`](https://CRAN.R-project.org/package=openmeteo) package to retrieve weather forecasts from the [Open Meteo API](https://open-meteo.com/).

We have already used the coordinates of the five main locations so all we have to do is to combine them into a list.

```{r}
#| label: list-locations

locations <- list(
  crd_kyoto,
  crd_liestal,
  crd_nyc,
  crd_vancouver,
  crd_washingtondc
)
```

We create a `get_fcst()` function that calls `openmeteo::weather_forecast()`. We can only get 15 days ahead weather forecasts, and we are about 3 days behind with the NASA POWER data.

```{r}
#| label: get_fcst

get_fcst <- function(loc_gc) {
  
  fcst <- weather_forecast(
    location = c(loc_gc$lat, loc_gc$long), 
    start = as.character(today() - days(3)), 
    end = as.character(today() + days(12)), 
    hourly = c(
      "temperature_2m", 
      "relative_humidity_2m", 
      "pressure_msl",
      "wind_speed_10m"
    ),
    daily = c(
      "temperature_2m_max",
      "temperature_2m_min",
      "precipitation_sum"
    )
  )
  
  fcst$location <- tolower(loc_gc$location)
  
  return(fcst)
}
```

Use the `get_fcst()` function to load the weather forecasts for the five main locations

```{r}
#| label: load-weather-forecasts
#| eval: false

locations_fcst <- purrr::map_dfr(.x = locations, .f = get_fcst) |> 
  rename(
    t2m = hourly_temperature_2m,
    t2m_max = daily_temperature_2m_max,
    t2m_min = daily_temperature_2m_min,
    prcp = daily_precipitation_sum,
    rh2m = hourly_relative_humidity_2m,
    ps = hourly_pressure_msl,
    ws10m = hourly_wind_speed_10m
  ) |> 
  group_by(location, date) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  relocate(location, date, t2m, t2m_max, t2m_min) |> 
  mutate(across(where(is.numeric), ~ round(.x, 2)))

write_csv(locations_fcst, file = "data/locations_fcst.csv")
```

Read in the forecasts

```{r}
#| label: read-locations_fcst-csv
#| output: false

locations_fcst <- read_csv("data/locations_fcst.csv")
```

Add the weather forecast to the predictors

```{r}
#| label: combine-power-openmeteo

last_obs_date <- power_cherry |> 
  filter(!is.na(t2m)) |> 
  slice_max(order_by = date) |> 
  select(date) |> 
  pull() |> 
  unique()

power_cherry <- bind_rows(
  power_cherry,
  locations_fcst |> filter(date > last_obs_date)
  ) |> 
  filter(!is.na(t2m))
```

Check out if the forecast was appended correctly. If not, there will be duplicates.

```{r}
power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  summarise(n = dplyr::n(), .by = c(location, date, variable)) |>
  filter(n > 1L)
```

Unfortunately, this will only give us a couple days of weather into March. We can either extrapolate this or combine it with additional sources.

## Feature engineering

We create some additional features from our predictors

### Midrange temperature

We create the midrange temperature, the average of the maximum and the minimum average temperature measured on a specific day.

```{r}
#| label: create-t2m_range

power_cherry <- power_cherry |> 
  mutate(t2m_range = (t2m_max + t2m_min) / 2) |> 
  relocate(location, date, t2m, t2m_max, t2m_min, t2m_range)
```

### Aggregated precipitation since January 1st

We use the variable daily measured precipitation (in mm) to calculate accumulated precipitation starting from January 1st of the year of the observation date. 

```{r}
#| label: create-aprcp

power_cherry <- power_cherry |> 
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(aprcp = cumsum(prcp)) |> 
  ungroup() |> 
  select(-year)
```

### Aggregated growing degree days since January 1st

We create the accumulated growing degree days (AGDD) as the sum of growing degree days (GDD) since January 1st, where GDD are calculated as the number of degrees C by which each day's midrange temperature exceeds a base temperature, which can be 0 degrees C, or a species-specific base temperature, such as 4.3 degrees C for Yoshino cherry trees. If the midrange temperature does not exceed the base temperature on a given day, no growing degree days are counted. See more at <https://en.wikipedia.org/wiki/Growing_degree-day>.

```{r}
#| label: create-agdd

base_temp <- 4.3

power_cherry <- power_cherry |> 
  mutate(gdd = pmax(t2m_range - base_temp, 0)) |>
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(agdd = cumsum(gdd)) |> 
  ungroup() |>
  select(-c(year, gdd))
```


### Aggregated chilling degree days since October 1st

We can calculate chilling degree days (CDD)** and accumulated chilling degree days (ACDD) since October 1st. We count CDDs whenever the midrange temperature is below a threshold of 7 degrees Celsius, and we accumulate until December 31st of each year. If the midrange temperature stays above the threshold, no GDDs are counted. See more at <https://en.wikipedia.org/wiki/Chilling_requirement>.

```{r}
#| label: create-acdd

temp_threshold <- 7

power_acdd <- power_cherry |> 
  mutate(year = year(date)) |> 
  filter(date >= paste0(year, "-09-01")) |> 
  mutate(cdd = pmax(0, temp_threshold - t2m_range)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(acdd = cumsum(cdd)) |> 
  ungroup() |> 
  select(-year)

power_cherry <- power_cherry |> 
  left_join(y = power_acdd |> select(location, date, acdd), join_by(location, date))

power_cherry <- power_cherry |> 
  mutate(acdd = if_else(is.na(acdd), 0, acdd))
```

## Select time range

I select only days between October 1st and the last available forecast date of the current year. That way, the model won't rely on data which is not available on the day of the submission deadline.

```{r}
#| label: select-time-range

last_fcst_date <- power_cherry |> 
  select(date) |>
  pull() |> 
  max()

power_cherry <- power_cherry |> 
  mutate(day = yday(date)) |> 
  relocate(location, date, day) |> 
  # filter(day %in% c(1:yday(last_fcst_date)))
  filter(day %in% c(yday(date("1980-12-01")):366, 1:yday(last_fcst_date)))
```

## Temporal aggregation

We need to decide at which temporal granularity we want our predictors.

### Seasons

```{r}
#| label: aggregate-to-seasons

df_seasons <- power_cherry |> 
  mutate(year = year(date), month = month(date)) |> 
  mutate(
    season = case_when(
      month %in% c(12, 1, 2) ~ "winter",
      month %in% c(3, 4, 5) ~ "spring",
      month %in% c(6, 7, 8) ~ "summer",
      month %in% c(9, 10, 11) ~ "winter"
    )
  ) |> 
  mutate(year = if_else(month == 12, year + 1, year)) |> 
  group_by(location, year, season) |> 
  summarise(across(where(is.numeric), mean), .groups = "drop") |> 
  ungroup() |> 
  pivot_longer(cols = -c(location, year, season), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", season)) |> 
  select(-season) |> 
  pivot_wider(names_from = variable, values_from = value)

cherry_seasons <- cherry |> 
  inner_join(y = df_seasons, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Months

```{r}
#| label: aggregate-to-months

df_months <- power_cherry |> 
  mutate(year = year(date), month = month(date)) |> 
  group_by(location, year, month) |> 
  summarise(across(where(is.numeric), mean), .groups = "drop") |> 
  ungroup() |> 
  pivot_longer(cols = -c(location, year, month), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", month)) |> 
  select(-c(month)) |> 
  pivot_wider(names_from = variable, values_from = value)

cherry_months <- cherry |> 
  inner_join(y = df_months, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Weeks

```{r}
#| label: aggregate-to-weeks

df_weeks <- power_cherry |> 
  mutate(year = year(date), week = week(date)) |> 
  select(-date) |> 
  group_by(location, year, week) |> 
  summarise(across(where(is.numeric), mean), .groups = "drop") |> 
  ungroup() |> 
  pivot_longer(cols = -c(location, year, week), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", week)) |> 
  select(-week) |> 
  pivot_wider(names_from = variable, values_from = value)

cherry_weeks <- cherry |> 
  inner_join(y = df_weeks, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Days

```{r}
#| label: aggregate-to-days
#| warning: true

df_days <- power_cherry |> 
  mutate(year = year(date), day = yday(date)) |> 
  relocate(location, year, day, date) |>
  pivot_longer(cols = -c(location, year, day, date), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", day)) |> 
  select(-c(day, date)) |> 
  pivot_wider(id_cols = c(location, year), names_from = variable, values_from = value)

cherry_days <- cherry |> 
  inner_join(y = df_days, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

## Exploratory data analysis (EDA)

### Linear model

A simple linear model can help us understand how predictive a feature is

Plot the average maximum temperature in winter `t2m_max_winter`

```{r}
#| label: fig-t2m_max_winter
#| fig-cap: "Average maximum winter temperature (Dec. 1st of the previous year until end of February)"

cherry_seasons |>
  ggplot(aes(x = year, y = t2m_max_winter)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum winter temperature (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```

We fit a linear model with different intercepts and time trends for each location and a weather effect. We can add interaction effects to the temperature data too if we want.

```{r}
#| label: fit-lin-t2m_max_winter

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:t2m_max_winter, 
    data = cherry_seasons
  )

lm_fit_months <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:(t2m_max_1 + t2m_max_2), 
    data = cherry_months
  )

lm_fit<- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:(t2m_max_1 + t2m_max_2 + t2m_max_3 + t2m_max_4 + t2m_max_5 + t2m_max_6 + t2m_max_7 + t2m_max_8), 
    data = cherry_weeks
  )

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_days
  )
```

Create fitted values for the bast observations

```{r}
#| label: predictions-t2m_max_winter

mean_pred <- predict(object = lm_fit, new_data = cherry_seasons)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-t2m_max_winter
#| fig-cap: "Predictions from linear time trend and average maximum winter temperature"

cherry_seasons |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

Because `t2m_max_winter` shows a clear upwards trend over the past twenty years, we are able model the trend towards earlier peak bloom dates quite well.

We see that the model with the linear time trend and the predictor `t2m_max_winter` does predict the data better than with only the linear time trend. We can assume that a model with `t2m_max_spring` would explain the data even better, but unfortunately, we don't have data for the spring of this year.

We also see that the average temperature in Kyoto is too mild to explain much of the variance in peak bloom times.

Let's fit a model that uses both the average maximum winter temperature and the average maximum spring temperature

```{r}
#| label: fit-lin-t2m_max_winter-t2m_max_spring

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:t2m_max_winter + location:t2m_max_spring, 
    data = cherry_seasons
  )
```

Create fitted values for the bast observations

```{r}
#| label: predictions-t2m_max_winter-t2m_max_spring

mean_pred <- predict(object = lm_fit, new_data = cherry_seasons)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-t2m_max_winter-t2m_max_spring
#| fig-cap: "Predictions from average maximum temperatures in winter and in spring"

cherry_seasons |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

Clearly, this model is able to fit the data even better. Unfortunately, even with weather forecasts, we will only know a couple days of March in advance, far from the three month period from March to May required to calculate `t2m_max_spring` for the year 2025.

We can try using monthly, weekly, and daily aggregates as predictors too.

```{r}
cherry_months |>
  ggplot(aes(x = year, y = t2m_max_2)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum temperature in February (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```


```{r}
cherry_weeks |>
  ggplot(aes(x = year, y = t2m_max_8)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum temperature in week 8 (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```


```{r}
cherry_days |>
  ggplot(aes(x = year, y = t2m_max_12)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum temperature on day 12 (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```


### Graphical

Plot a linear time trend

```{r}
cherry_seasons |>
  ggplot(aes(x = year, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

Plot peak dates vs aggregated growing degree days in spring

```{r}
cherry_seasons |>
  ggplot(aes(x = agdd_spring, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```


### Correlations

We use the `{corr}` package to check the correlations of the predictors.

```{r}
cherry_seasons |>
  correlate() |>
  gt() |>
  fmt_number() |>
  tab_header(title = "correlations") |>
  tab_footnote(
  footnote = ""
)
```

### Decision tree

Fit a decision tree for exploratory data analysis. We use the `{parsnip}` package to create a model that runs on the `"rpart"` engine, and we use the `rpart.plot` package to plot the decision tree.

```{r}
set.seed(456)

# Create a {parsnip} model that runs on the "rpart" engine
cart_specs <- decision_tree(
  tree_depth = 10,
  cost_complexity = 0.001
  ) |>
  set_mode("regression") |>
  set_engine("rpart")

# fit the model 
cart_fit <- cart_specs |>
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_days
  )

# Visualize with {rpart.plot}
cart_fit$fit |> 
  rpart.plot()
```

On a seasonal aggregation level, aggregated growing degree days (AGDD) are the most important predictor.

### Variable importance plot (VIP)

Use the `{vip}` package to create a variable importance plot (VIP)

```{r}
cart_fit |> 
  vip()
```

## Modeling

### Data spending

We only have `r nrow(cherry_days)` observations of the peak bloom date. We therefore need to be careful how we spend our data.

We need to split these observations randomly into
- a training set that will be used to train the model
- a validation set that will be used to calibrate the model
- a test set that will be used to compare the performance different models against each other

Because the number of observations are so small, any random split will have huge consequences. Stratified splitting may somewhat help here.

```{r}
set.seed(123)

data_split <- initial_validation_split(cherry_days)

training <- training(data_split)
validate <- validation(data_split)
testing  <- testing(data_split)
```

### Feature engineering and pre-processing

```{r}
cherry_recipe <- recipe(formula = bloom_doy ~ ., data = training) |>
  step_corr(threshold = 0.90) |>
  step_dummy(all_nominal_predictors()) |>
  step_YeoJohnson(all_nominal_predictors())
```

## Training stage

Within the training stage we will evaluate hyperparameters via cross-validation.
In 10-fold cross validation (CV), we will randomly split the training set into 10 roughly equally-sized subsets, and we will repeatedly fit the model on 9 of these folds, which constitute the *analysis* set, and evaluate the models' performance on the remaining fold, the *asssessment* set.

We also want to ensure the sampling is stratified. This ensures the samples represent all quantiles of bloom_doy. We use the option `strata = bloom_doy` in `vfold_cv()`.

```{r}
set.seed(123)

folds <- vfold_cv(training, v = 10, strata = bloom_doy)
```

Random forest

```{r}
forest_grid <- expand.grid(
  trees = c(20, 50),
  mtry  = c(2, 4, 6)
)

rf_spec <- rand_forest(
  trees = tune(),
  mtry  = tune()
  ) |>
  set_engine("ranger") |>
  set_mode("regression")

# Use the {finetune} function tune_race_anova()
rf_results <- tune_race_anova(
  object = workflow() |>
    # our pre-process
    add_recipe(cherry_recipe) |>
    # our model
    add_model(rf_spec),
  # folds
  resamples = folds,
  # parameters
  grid = forest_grid,
  # racing as control
  control = control_race(),
  # loss function to minimize
  metrics = metric_set(yardstick::rmse)
)
```

Boosted tree

```{r}
## boosted tree ----
lgb_grid <- expand.grid(
  trees      = c(3, 10, 25),
  learn_rate = c(0.5, 0.05, 0.005)
)

lgb_spec <- boost_tree(
  trees      = tune(),
  learn_rate = tune()
  ) |>
  # boosted lgbm model
  set_engine("lightgbm") |>
  set_mode("regression")

lgb_results <- finetune::tune_race_anova(
    workflow() |>
      # our pre-process
      add_recipe(cherry_recipe) |>
      # our model
      add_model(lgb_spec),
    # folds
    resamples = folds,
    # parameters
    grid = lgb_grid,
    # racing as control
    control = control_race(),
    # loss function to minimize
    metrics = metric_set(yardstick::rmse)
)
```

Cubist rules

```{r}
# cubist rules -----
cub_grid <- expand.grid(
  committees      = c(30, 45),
  max_rules       = c(5, 10, 30)
)

cub_spec <- cubist_rules(
  committees = tune(),
  max_rules  = tune()
  ) |>
  # quinlan's cubist
  set_engine("Cubist") |>
  set_mode("regression")

cub_results <- tune_race_anova(
  object = workflow() |>
    # our pre-process
    add_recipe(cherry_recipe) |>
    # our model
    add_model(cub_spec),
  # folds
  resamples = folds,
  # parameters
  grid = cub_grid,
  # racing as control
  control = control_race(),
  # loss function to minimize
  metrics = metric_set(rmse)
)
```



### Model selection

After selecting the model with the best hyperparameter specification based on the results of k-fold CV, we fit the model to the whole training data set.

We then evaluate these models on the test set, and compare their performance.

We can have a look at the best hyperparameter specification for each model based on k-fold CV.

```{r}
select_best(rf_results)
select_best(lgb_results)
select_best(cub_results)
```

Then we fit these models on the whole training data set

Random forest

```{r}
rf_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(rf_spec),
    parameters = select_best(rf_results)
    ),
  data = training
  )
```

Boosted tree

```{r}
lgb_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
        add_recipe(cherry_recipe) |>
        add_model(lgb_spec),
    parameters = select_best(lgb_results)
    ),
  data = training
  )
```

Cubist rules

```{r}
cub_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(cub_spec),
      select_best(cub_results)
    ),
  data = training
  )
```

### Validation stage

The validation set is used to validate, and calibrate the models that were selected using k-fold CV and fitted on the whole training data set.

We use the `{probably}` package to add prediction intervals to our point forecasts.

Post-processing is important to handle any adjustments and bias for predictions.

We then use the calibrated models on our test dataset for a final evaluation.

Create the predictions for validation data

```{r}
rf_vpreds <- predict(rf_fit, validate)
lgb_vpreds <- predict(lgb_fit, validate)
cub_vpreds <- predict(cub_fit, validate)
```

Make a large dataframe with all preds and truth

```{r}
validate_preds <- validate |>
  cbind(
    rf = rf_vpreds$.pred,
    lgb = lgb_vpreds$.pred,
    cub = cub_vpreds$.pred
  ) |>
  relocate(year, location, bloom_doy, rf, lgb, cub) |> 
  pivot_longer(cols = c(rf, lgb, cub), names_to = "model", values_to = "value") |> 
  select(year, location, bloom_doy, model, value)
```

Visualize the certainty in predictions

```{r}
validate_preds |>
  ggplot(aes(x = bloom_doy, y = value, color = model)) +
  geom_point() +
  coord_equal() +
  geom_abline(lty = 2) +
  labs(title = "bloom_doy & predictions")
```

Quick error measurement

```{r}
validate_preds |>
  mutate(abs_error = abs(bloom_doy - value)) |>
  group_by(model) |>
  summarize(avg_abs_error = mean(abs_error)) |> 
  arrange(avg_abs_error)
```

## Calibration Stage

The top performing models on validation are cubist and random forest

Cubist is already pretty precise and accurate but we will calibrate our random forest to see if we increase accuracy on the validation set.

Calibrate random forest with validation data

```{r}
calibrate_rf <- data.frame(
  bloom_doy = validate$bloom_doy,
  .pred = rf_vpreds$.pred
)

rf_fit_cal <- calibrate_rf |>
  cal_estimate_linear(bloom_doy)

calibrated_rf_preds <- calibrate_rf |>
  cal_apply(rf_fit_cal)

calibrated_rf_preds |>
  mutate(abs_error = abs(bloom_doy - .pred)) |>
  summarize(avg_abs_error = mean(abs_error))
```

There is an improvement but note we need to **test** the calibration

### Model with conformal coverage

While we have optimized our point forecasts with pre-processing, tuning, and post-processing, it's also important to capture our model uncertainty and to create robust prediction intervals with the validation set.

Let's create confidence intervals for our models. The simplest form being is the conformal split method implemented in the `{probably}` function `int_conformal_split()`.

```{r}
rf_int_model <- probably::int_conformal_split(rf_fit, validate)
lgb_int_model <- probably::int_conformal_split(lgb_fit, validate)
cub_int_model <- probably::int_conformal_split(cub_fit, validate)
```

We have obtained conformal interval models.

```{r}
rf_int_model
lgb_int_model
cub_int_model
```

## Testing stage

We use the test data set to evaluate the performance of the models that we have trained on the test set and calibrated on the validation set.

Create the predictions and prediction intervals on the test set

```{r}
cubist_preds <- predict(cub_int_model, testing, level = 0.90)

rf_cal_preds <- predict(rf_int_model, testing, level = 0.90) |> 
  cal_apply(rf_fit_cal)

lgb_int_preds <- predict(lgb_int_model, testing, level = 0.90)
```

Combine the test set and the predictions

```{r}
test_df <- testing |>
  cbind(
    cubist = cubist_preds$.pred,
    cubist_upper = cubist_preds$.pred_upper,
    cubist_lower = cubist_preds$.pred_lower,
    rf_cal = rf_cal_preds$.pred,
    rf_cal_upper = rf_cal_preds$.pred_upper,
    rf_cal_lower = rf_cal_preds$.pred_lower,
    lgb = lgb_int_preds$.pred,
    lgb_upper = lgb_int_preds$.pred_upper,
    lgb_lower = lgb_int_preds$.pred_lower
  ) |> 
  select(year, location, bloom_doy, cubist, cubist_upper, cubist_lower, rf_cal, rf_cal_upper, rf_cal_lower, lgb, lgb_upper, lgb_lower)
```


```{r}
test_df |>
  ggplot() +
  geom_point(aes(x = bloom_doy, y = bloom_doy, color = "actual")) +
  geom_errorbar(aes(xmin = lgb_lower, xmax = lgb_upper, y = bloom_doy, color = "lgb_interval"), alpha = 0.4) +
  geom_point(aes(x = cubist, y = bloom_doy, color = "cubist")) +
  geom_point(aes(x = rf_cal, y = bloom_doy, color = "rf_calibrated")) +
  coord_equal() +
  geom_abline(lty = 2) +
  labs(x = "predictions", y = "bloom_doy")
```
Test accuracy

```{r}
test_df |>
  mutate(
    cub_abs_error = abs(bloom_doy - cubist),
    rf_abs_error = abs(bloom_doy - rf_cal),
    lgb_coverage = ifelse(bloom_doy >= lgb_lower & bloom_doy <= lgb_upper, 1, 0),
    n = row_number()
  ) |>
  summarize(
    cub_total_abs_error = sum(cub_abs_error),
    rf_total_abs_error = sum(rf_abs_error),
    lgb_coverage_pct = sum(lgb_coverage) / max(n)
  )
```

## Next Steps

- `{workflowsets}`: methods to test across many engines and recipe combinations
- `{yardstick}`: other loss function or metrics to optimize
- `{dials}`: for creating draws for hyperparameters
- `{autoML}`: more automation for engine selection and ensembling
- `{tailor}`: a new post-processing technique for workflows
- `{stacks}`: we can also ensemble many models to outperform any one model
- meta learners & diagnostics: while in production, we can refit as we have additional observations, potentially add another model to learn best hyperparameter setpoints for best out-of-sample accuracy
- it's important to understand trade-offs of the model in operation over time and when to refit, calibrate, or start from scratch

### Nonlinear models

#### Random forest

Let's fit a random forest

```{r}
set.seed(567)
cores <- parallel::detectCores()

# Split the data with {rsamples}
cherry_split <- initial_split(data = cherry_weeks)
cherry_train <- training(cherry_split)
cherry_test  <- testing(cherry_split)

# Create folds with {rsamples}
folds <- vfold_cv(cherry_train, v = 5)
```

Create model with `{parsnip}`

```{r}
rf_mod <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
  ) |> 
  set_engine("ranger", num.threads = cores, importance = "impurity") |> 
  set_mode("regression")
```

Show what will be tuned

```{r}
extract_parameter_set_dials(rf_mod)
```

Create worfklow with `{workflows}`

```{r}
rf_workflow <- workflow() |> 
  add_model(rf_mod) |> 
  add_formula(bloom_doy ~ .)
```

Create a tuning grid with `{dials}`

```{r}
tree_grid <- grid_regular(
  mtry(range = c(1, 9)),
  min_n(),
  levels = 5
)
```

Train the model on the folds

```{r}
tree_res <- rf_workflow |> 
  tune_grid(
    resamples = folds,
    grid = tree_grid,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(rmse)
  )
```

Show the metrics

```{r}
tree_res |> 
  collect_metrics()
```

Show the model with the best RMSE

```{r}
tree_res |> 
  show_best(metric = "rmse")
```

Select the parameter combination with the best k-fold CV performance

```{r}
best_tree <- tree_res |> 
  select_best(metric = "rmse")
```

Finalize the model

```{r}
final_wf <- rf_workflow |> 
  finalize_workflow(parameters = best_tree)
```

Show the final model

```{r}
final_wf
```

Fit the model best model on all the training data with the `{tune}` function `last_fit()`

```{r}
final_fit <- final_wf |> 
  last_fit(cherry_split)
```

Show the metrics of the final fit

```{r}
final_fit |> 
  collect_metrics()
```

Plot the predictions versus the actual values

```{r}
final_fit |> 
  collect_predictions() |> 
  ggplot(mapping = aes(x = .pred, y = bloom_doy)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  scale_x_continuous(limits = c(80, 110)) +
  scale_y_continuous(limits = c(80, 110)) +
  coord_equal(ratio = 1) +
  theme_bw()
```

```{r}
# To predict new data points, use {workflows} extract_worfklow()
final_tree <- extract_workflow(final_fit)

final_tree
```

Plot the variable importance plot (VIP)

```{r}
final_tree |> 
  extract_fit_parsnip() |> 
  vip()
```



#### LightGBM

```{r}

```


#### XGboost

```{r}

```


## Submission

To submit your entries, enter the predicted bloom dates from your models in the **submission form**.

```{r}
#| label: tbl-predictions-for-submission
#| tbl-cap: "Predictions for Submission"

predictions |>
  filter(year == 2025) |>
  mutate(prediction_date = strptime(paste(year, prediction), "%Y %j") |> as_date()) |>
  kable() |> 
  kable_styling()
```

**NOTE**: The question is if we need to round the point estimates or if we can submit partial days.