---
title: "Peak Bloom Prediction"
author: "econmaett"
lang: en-US
format:
  html:
    embed-resources: true
fig-cap-location: bottom
tbl-cap-location: bottom
toc: true
editor: source
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, error = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "80%"
)
```

We start by loading all the packages necessary to run this analysis

```{r}
#| label: load-pkgs
#| output: false

invisible(Sys.setlocale("LC_ALL","English"))

library(tidymodels)
library(kableExtra)
library(rvest)
library(httr2)
library(jsonlite)
library(nasapower)
library(sf)
library(mapview)
```

## Data

We begin by loading the data necessary for this analysis

### Clean data

The data for the five sites is provided as a simple text file in CSV format. Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site. The six columns in each data file are:
- `location`: a human-readable location identifier (`string`).
- `lat`: (approximate) latitude of the cherry trees (`double`).
- `long`: (approximate) longitude of the cherry trees (`double`).
- `alt`: (approximate) altitude of the cherry trees (`double`).
- `year`: year of the observation (`integer`).
- `bloom_date`: date of peak bloom of the cherry trees ([ISO 8601](https://xkcd.com/1179/) date `string`). The "peak bloom date" may be defined differently for different sites
- `bloom_doy`: days since January 1st until peak bloom (`integer`). January 1st is `1`.

#### Kyoto

The file `kyoto.csv` has been obtained from Yasuyuki AONO (`aono(at)envi.osakafu-u.ac.jp`) at <http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/>.

The geographical location at `35.01198°N` (latitude), `135.6761°E` (longitude), `44` meters above sea levels (altitude) roughly corresponds to the **Nakanoshima** area of the **Arashiyama Park** in Kyoto, Japan.

The peak bloom date of the **Prunus jamasakura** is determined by a local news paper in Arashiyama (Kyoto, Japan). Data prior to 1888 is extracted from various descriptions or estimated:
- Data from the 9th to the 14th centuries was acquired and analyzed by Aono and Saito (2010; International Journal of Biometeorology, 54, 211-219).
- Phenology for 15th to 21st centuries was acquired and analyzed by Aono and Kazui (2008; International Journal of Climatology, 28, 905-914).

Note that the data set for Kyoto includes implicit missing values (`NA`s). [Chapter 18 "Missing Values"](https://r4ds.hadley.nz/missing-values) of Hadley Wickham's book "R for Data Science (2e)" describes strategies for dealing with implicit `NA`s.

Let's load the data for Kyoto, Japan

```{r}
#| label: load-kyoto

kyoto <- read_csv("data/kyoto.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  fill(location, lat, long, alt, .direction = "downup")
```

Let's plot the data for Kyoto, Japan

```{r}
#| label: fig-kyoto
#| fig-cap: "Peak cherry blossom days in Kyoto, Japan"

fig_kyoto <- kyoto |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 0.5) +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(800, 2100, by = 100), NA)) +
  labs(title = "Peak cherry blossom days in Kyoto, Japan", x = NULL, y = NULL) +
  theme_light()

print(fig_kyoto)

ggsave(filename = "figures/fig_kyoto.png", plot = fig_kyoto, bg = "white")
```

Here we have used **locally estimated scatterplot smoothing (LOESS)** or **locally weighted scatterplot smoothing (LOWESS)** to who how the average peak bloom day changed over the years. 

We could explicitly model this with a **local level (+ trend)** state-space model that we fit with the **Kalman filter and Smoother**, which also allows us to effectively deal with **missing observations**.

MeteoSwiss suggests using a **Gaussian Kernel** with a *20-year window* to extract the trend, which is equivalent to using a **low-pass filter** that shuts down all higher frequencies.

It also looks like the observations exhibit **heteroskedasticity**, that is, the variance changes over time, and it looks like the variance has become more narrow over time.

We can use **changepoint detection** methods to test for the changing time trend and the changing variance.

Assuming that the data in the past is correct, there do seem to be seasonal cycles, perhaps connected to solar activity or the tilt of the earth.

There is a clear trend towards earlier blooming, perhaps connected to anthropogenic climate change and urbanization. Studies attribute about half of the earlier blooming times in recent years to the **urban heat island effect**, and the other half to **climate change**.

Modelling the local level and the trend would be useful if we are to forecast a couple years into the future, since an individual prediction may be incorrect, but on average, they may be good. 

However, since we are only interested in forecasting a single moment in time, which may be greatly influenced by local weather, we probably are better off using historical weather data that we augment with forecasts, either by professional forecasters or by extrapolating the previous predictors.

### Liestal

The file `liestal.csv` contains observations for **a single wild cherry (Prunus avium) tree** in **Liestal-Weideli**, at `7.4814°N` (latitude), `7.730519°E` (longitude) and `350` meters above sea levels (altitude). 

The **peak bloom date** is defined as the day when **25%** of the blossoms are in full bloom. The date is determined by [MeteoSwiss](https://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html).

This means that 70% of buds would be open a couple days later. This raises the question of whether we should model the locations separately.

The data is provided by **Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss**.

```{r}
#| label: load-liestal

liestal <- read_csv("data/liestal.csv")
```

Let's plot the data for Liestal, Switzerland:

```{r}
#| label: fig-liestal
#| fig-cap: "Peak cherry blossom days for Liestal, Switzerland"

fig_liestal <- liestal |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#ff0080", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Liestal, Switzerland", 
    x = NULL, y = NULL
  ) +
  theme_light()

print(fig_liestal)

ggsave(filename = "figures/fig_liestal.png", plot = fig_liestal, bg = "white")
```

For Liestal you can see the impact of the infamous heat summer of 2003 which I remember from my childhood, although interestingly, there's another record in 1990.

### New York

The file `newyorkcity.csv` contains **only a single observation** for the year 2024. It concerns the **Yoshino Cherry (Prunus x yedoensis)** tree in **Washington Square Park**, New York City, at `40.7304 °N` (latitude), `-73.99809°E` (longitude) and `8.5` meters above sea level (altitude).

Let's load the data for New York City

```{r}
#| label: load-newyorkcity

newyorkcity <- read_csv("data/nyc.csv")
```

Luckily, there are some historical observations dating *back to 2019* in the data provided by **USA-NPN**.

The Washington Square Park has site id `32789` and the **Yoshino** cherry you should predict has species id `228`.

We have been provided with USA-NPN data on individual cherry trees by the competition organizers, and we can load it:

```{r}
#| label: load-historical-data-nyc

nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789, Species_ID == 228) |> 
  mutate(Observation_Date = mdy(Observation_Date))
```

This file now contains individual observations of the phenophase and not the actual peak bloom date.

We use the variable `Intensity_Value` which is either `-9999`, `95% or more`, `Less than 5%`, `25-49%`, `75-94%`, `50-74%`.

Since according to the variable `Individual_ID`, only one tree, `183140` is being observed, we can use the first date where `Intensity_Value` is `50-74%` as our `bloom_doy`.

```{r}
#| warning: true

nyc_data <- nyc_data_npn |>
  arrange(Observation_Date) |>
  mutate(year = year(Observation_Date)) |>
  group_by(year) |>
  summarize(
    first_flower_index = min(which((Intensity_Value %in% c("75-94%", "95% or more")))),
    bloom_date = as_date(strftime(Observation_Date[first_flower_index], format = "%Y-%m-%d")),
    bloom_doy = Day_of_Year[first_flower_index],
    .groups = "drop"
  ) |>
  filter(!is.na(bloom_doy)) |>
  select(-first_flower_index) |>
  mutate(location = "newyorkcity") |> 
  arrange(year)
```

Then we combine the data sets

```{r}
#| label: augment-nyc-data

newyorkcity <- newyorkcity |>
  # Don't add a year twice!
  bind_rows(nyc_data |> filter(year != 2024)) |>  
  fill(lat, long, alt)
```

We now have a total of four observations for New York City

```{r}
#| label: tbl-nyc-data
#| tbl-cap: "Augmented NYC observations"

newyorkcity |> 
  filter(location == "newyorkcity") |> 
  kable() |> 
  kable_styling()
```

### Vancouver

Let's load the data for Vancouver, BC

```{r}
#| label: load-vancouver

vancouver <- read_csv("data/vancouver.csv")
```

The file `vancouver.csv` contains observations for cherry trees of the cultivar  **Akebono** in **Maple Grove Park**, Vancouver, BC located approximately at `49.2236916°N` (latitude), `-123.1636251°E` (longitude), `24` meters above sea levels (altitude).

Unfortunately, there are **only three observations** for 2022, 2023, and 2024, respectively. But casual observations have been recorded in the form of **photos posted** to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/). You can **search the forum** for the keywords `"Akebono"` (i.e., the name of the cultivar) and `"Maple Grove Park"` (i.e., the location of the trees).

After searching through the web forum, I have decided to add the following observations to the data set:

```{r}
#| label: define-vancouver_forum

vancouver_forum <- tribble(
  ~ year, ~ bloom_date,
  2021, "2021-03-31", # Start: 2021-03-21, Full: 2021-03-31, Full: 2021-04-07
  2020, "2020-04-03", # Full bloom
  2018, "2018-03-18", # Full bloom
  2017, "2017-04-11", # Full bloom 
  2015, "2015-03-19", # Full bloom
  2011, "2011-04-01", # Full bloom was over on 2011-04-15
  2010, "2010-03-14", # 60% bloom on 2010-03-07
  2008, "2008-04-14" # Start: 2008-04-06, Full: 2008-04-14, finished: 2008-04-22
  ) |> 
  mutate(
    bloom_date = as_date(bloom_date),
    bloom_doy = as.numeric(1 + as_date(bloom_date) - as_date(paste0(year, "-01-01"))),
    location = "vancouver"
  )
```

We can do some additional plausibility checks. For example, spring 2010 was indeed unusually warm, [causing cherry blossoms to bloom during the Vancouver 2010 Winter Olympics](https://www.upi.com/News_Photos/view/upi/33393f9d20159ecb347de9aa1d670922/2010-Winter-Olympics-in-Vancouver/). National Geographic claimed it was the [warmest Olympics to date](https://www.nationalgeographic.com/adventure/article/100212-vancouver-2010-warmest-winter-olympics).

We add these observations to the data set

```{r}
#| label: augment-vancouver

vancouver <- vancouver |> 
  bind_rows(vancouver_forum) |> 
  fill(lat, long, alt) |> 
  arrange(year)
```

Now we have a total of eleven observations for Vancouver, BC.

```{r}
#| label: tbl-augmented-vancouver-data
#| tbl-cap: "Augmented Vancouver data"

vancouver |> 
  filter(location == "vancouver") |> 
  kable() |> 
  kable_styling()
```

### Washington

The file `washingtondc.csv` has been obtained from the EPA's Climate Change Indicators in the United States <https://www.epa.gov/climate-indicators/cherry-blossoms>.

The geographical location at `38.88535°N` (latitude), `-77.03863°E` (longitude) and `0` meters above sea level (altitude) correspond to the location of the [Tidal Basin in Washington, DC](https://www.nps.gov/articles/dctidalbasin.htm).

The peak bloom date is defined as the day when **70%** of the **Yoshino Cherry (Prunus x yedoensis)** are in full bloom, as determined by the [National Park Service](https://www.nps.gov/subjects/cherryblossom/bloom-watch.htm).

Let's load the data for Washington DC

```{r}
#| label: load-washingtondc

washingtondc <- read_csv("data/washingtondc.csv")
```

Let's plot the data for Washington D.C

```{r}
#| label: fig-washingtondc
#| fig-cap: "Peak cherry blossom days for Washington DC, USA"

fig_washingtondc <- washingtondc |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#ff0080", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(title = "Peak cherry blossom days for Washington DC, USA", x = NULL, y = NULL) +
  theme_light()

print(fig_washingtondc)

ggsave(filename = "figures/fig_washingtondc.png", plot = fig_washingtondc, bg = "white")
```

Now that we have loaded the datasets for the five locations, we can combine them into a single CSV file called `cherry.csv`.

```{r}
#| label: combine-cherry-csv
#| eval: false

cherry <- bind_rows(kyoto, liestal, newyorkcity, vancouver, washingtondc)

write_csv(cherry, file = "data/cherry.csv")
```

Now we can load the `cherry.csv` file from our `data/` folder

```{r}
#| label: load-cherry

cherry <- read_csv("data/cherry.csv")
```

We can show the last three observation for each location

```{r}
#| label: tbl-01
#| tbl-cap: "latest 3 observations for each location"

cherry |> 
  group_by(location) |> 
  slice_tail(n = 3) |> 
  kable() |> 
  kable_styling()
```

We can visualize the time series for our five main locations next to each other. Notice the use of the `{stringr}` function `str_to_title()` to convert the first letter of the locations to uppercase.

```{r}
#| label: fig-cherries_1880
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at five different sites.

df_cherry_1880 <- cherry |> 
  filter(year >= 1880)

fig_cherries <- ggplot(data = df_cherry_1880, aes(year, bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab") +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 40), NA)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(
    title = "Peak bloom (since Jan 1st)",
    x = NULL, 
    y = NULL
  ) +
  theme_light()

print(fig_cherries)

ggsave(filename = "figures/fig_cherries.png", plot = fig_cherries, bg = "white")
```

Create summary statistics to get an overview over the data. Specifically, we are interested in the range of the `year` variable, the total number of observations at each location, and the mean, median, and standard deviation of the `bloom_doy` variable.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for the main data set"

cherry |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location) |> 
  summarise(
    min_year = min(year),
    max_year = max(year),
    n_obs = n(),
    mean = round(mean(bloom_doy)),
    median = round(median(bloom_doy)),
    sd = round(sd(bloom_doy))
  ) |> 
  kable() |> 
  kable_styling()
```

We see that the standard deviation for Liestal is much larger than the rest, where peak bloom dates are on average about a weak apart from their historical average.

We check for missing observations in the most recent years:

```{r}
cherry |> 
  filter(location == "kyoto") |> 
  filter(is.na(bloom_doy)) |> 
  tail() |> 
  kable() |> 
  kable_styling()
```

Kyoto has multiple implicit `NA`s, including in 1730, 1872, 1895, 1919, 1921, and 1945. So if we use data since 1880, as the demo analysis suggests, we still have four `NA`s. For time series analysis, we usually assume equally-spaced observations, and therefore need to either cut the data to start in 1946, or to replace the `NA`s with a sensible value. The other four locations do not contain any implicit `NA`.

Let's plot the distribution of the peak bloom days for the locations:

```{r}
#| label: fig-histogram
#| fig-cap: "Histogram of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location)) +
  geom_histogram() +
  labs(title = "Histogram of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's use a density plot instead:

```{r}
#| label: fig-density-plot
#| fig-cap: "Denisty plots of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location, colour = location)) +
  geom_density(alpha = 0.5) +
  labs(title = "Denisty plots of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's plot all five time series in the same plot.

```{r}
#| label: fig-cherry-blossoms-combined
#| fig-cap: "Peak cherry blossom days for different locations"

fig_combined <- cherry |> 
  filter(year >= 1880) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy, colour = location)) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  geom_step(linetype = "dotted", lwd = 0.5) +
  geom_point(alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(title = "Peak cherry blossom days by locations", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())

print(fig_combined)

ggsave(filename = "figures/fig_combined.png", plot = fig_combined, bg = "white")
```

This graphic highlights the negative linear time trend for all locations.

It also highlights that a linear time trend will likely underestimate the accelerating trend towards earlier peak bloom dates.

A simple method such as **ETS**, which places a lower weight on observations far in the past, may be better at modelling the recent trend.

In any case, we should compare all our models against **benchmark models** such as the **naive model**, where the prediction is equal to the last observation, and a simple **AR(1)** model.

## Additional data

The competition organizers provide additional data sets for sites other than the main sites relevant for the competition. 

We can use these time series in your modeling to help with spatial and temporal extrapolation. 

For example, you can add predictors to these data sets and train models on additional sites.

### Japan

The data file `japan.csv` is provided by the [Japanese Meteorological Agency (JMA)](https://www.data.jma.go.jp/sakura/data/pdf/005.pdf) and contains recorded peak bloom dates for cherry trees at various sites across Japan. The sample trees are located within a 5km radius of the location indicated in the data file. 

We can load the data set, make implicit missing values explicit and change the `location` variable to only include the city names in lowercase letters.

```{r}
#| label: read-japan

japan <- read_csv("data/japan.csv") |>  
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Japan/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `japan.csv` file contains observations between `r min(japan$year)` and `r max(japan$year)` for `r length(unique(japan$location))` locations in Japan.

Let's plot peak bloom dates for Kyoto, Japan

```{r}
#| label: fig-bloom-kyoto2
#| fig-cap: "Peak bloom dates in Kyoto, Japan"

fig_kyoto2 <- japan |>
  filter(location == "kyoto") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 10), NA)) +
  labs(title = "Peak bloom (since Jan 1st) in Kyoto, Japan", x = NULL, y = NULL) +
  theme_light()

print(fig_kyoto2)

ggsave(filename = "figures/fig_kyoto2.png", plot = fig_kyoto2, bg = "white")
```

### Korea

The file `south_korea.csv` contains **first flowering dates** of cherry trees at various sites across South Korea, curated by the **Korean Meteorological Administration (KMA)**.

Because these are **first flowering dates**, they are likely to occur earlier than hypothetical full bloom dates would be.

Trees in South Korea were assumed to be of the **Prunus Yedoensis** variety based on previous work (Hur, Ahn, and Shim 2014; Chung et al. 2009) and its frequent occurrence.

Again, we make implicit missing values explicit, and I change the location names to lowercase so that to match the style of the main data set.

```{r}
#| label: load-korea

korea <- read_csv("data/south_korea.csv") |> 
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "South Korea/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `south_korea.csv` file contains observations between `r min(korea$year)` and `r max(korea$year)` at `r length(unique(korea$location))` locations in South Korea.

Let's plot the first bloom dates in Seoul, South Korea:

```{r}
#| label: fig-seoul
#| fig-cap: "First bloom in Seoul, South Korea"

fig_seoul <- korea |>
  filter(location == "seoul") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1980, 2020, by = 5), NA)) +
  labs(title = "First bloom (since Jan 1st) in Seoul, Korea", x = NULL, y = NULL) +
  theme_light()

print(fig_seoul)

ggsave(filename = "figures/fig_seoul.png", plot = fig_seoul, bg = "white")
```

### MeteoSwiss

The data file `meteoswiss.csv` contains peak bloom dates for various sites across Switzerland, obtained from <https://opendata.swiss/en/dataset/phanologische-beobachtungen>.

Note that when you use the `{utils}` function `read.csv()` you need to specify the option `encoding = "UTF-8"` to allow for special characters that are common in the German language.

```{r}
#| label: load-meteoswiss

meteoswiss <- read_csv("data/meteoswiss.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Switzerland/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `meteoswiss.csv` file contains observations between `r min(meteoswiss$year)` and `r max(meteoswiss$year)` at `r length(unique(meteoswiss$location))` locations.

Plot the data for Zurich, Switzerland

```{r}
#| label: fig-meteoswiss
#| fig-cap: "Peak bloom (since Jan 1st) in Zurich, Switzerland"

fig_meteoswiss <- meteoswiss |> 
  filter(str_detect(location, "zürich-")) |>
  mutate(location = str_remove(location, "zürich-")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 20), NA)) +
  labs(
    title = "Peak bloom (since Jan 1st) in Zurich, Switzerland",
    x = NULL, y = NULL
  ) +
  theme_light() +
  facet_grid(rows = vars(str_to_title(location)))

print(fig_meteoswiss)

ggsave(filename = "figures/fig_meteoswiss.png", plot = fig_meteoswiss, bg = "white")
```

### NPN

Two additional files are provided by the **USA National Phenology Network (USA-NPN)** and the many participants who contribute to its *Nature’s Notebook* program.

The file `USA-NPN_individual_phenometrics_data.csv`contains observations on **individual phenometrics** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-individual-phenometrics

npn_ind_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  mutate(Observation_Date = mdy(Observation_Date))

npn_ind_des <- read_csv("data/USA-NPN_individual_phenometrics_datafield_descriptions.csv")
```

The file `USA-NPN_status_intensity_data.csv` contains observations on the **status and intensity** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-status-intensity

npn_int_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv")

npn_int_des <- read_csv("data/USA-NPN_status_intensity_datafield_descriptions.csv")
```

These files can help augment the existing data sets for which you don't have many observations such as New York City (and perhaps even Vancouver). 

They also contain some useful **predictors** such as weather variables from [Daymet](http://daymet.ornl.gov). We ignore variables measured in degrees Fahrenheit (`Tmin_in_F`, `Tmax_in_F` and `AGDD_in_F`).

- `Observation_ID`: The unique identifier of each phenophase status record (subsequently referred to as "status record") in the database.

- `Update_Datetime`: The date and time the status record was last updated after original submission online. `"-9999"` indicates the record has not been updated since this field was established in July 2014.

- `Site_ID`: The unique identifier of the site at which the status record was made.

- `Latitude`, `Longitude` and `Elevation_in_Meters`: The site at which the status record was made. Calculated from the [Google Maps API](https://developers.google.com/maps) with a datum of `WGS84`. `"-9999"` indicates the elevation could not be calculated.

- `State`: The U.S. state or territory, Mexican state or Canadian province in which the site is located. The state is calculated from lat/long by the [Google Maps Geocoding API](https://developers.google.com/maps/documentation/geocoding/intro). `"-9999"` means the site is outside North America.

- `Species_ID`: Identifier of the species for which the status record was made.

- `Genus`, `Species` and `Kingdom`: Taxonomic genus, species and kingdom of the plant for which the status record was made following the [Integrated Taxonomic Information System](http://itis.gov). Subspecies are appended to the species name after a hyphen (e.g. *Cornus florida-appalachianspring*).

- `Common_Name`: The common name of the species for which the status record was made followwing the [USDA PLANTS Database](http://plants.usda.gov).

- `Individual_ID`: Individuals plants are tracked separately with a unique identifier. 

- `Observation_Date`: The date the status observation was made in the field.

- `Day_of_Year`: The day of year, ranging from `1` to `366` when the status observation was made.

- `Phenophase_Status`: Status (presence or absence) of the phenophase at the time of observation.

- `Intensity_Category_ID`: The degree to which a phenophase is expressed. `"-9999"` indicates there was no intensity category assigned to this phenophase for this species at observation time.

- `Intensity_Value`: The value reported in response to the intensity category question assigned to the phenophase for this species at observation time.

- `AGDD`: **Accumulated growing degree days (in °C)** on the `Observation_Date`. This is the **sum of growing degree days (GDD) since January 1st**, where GDD is calculated as the number of degrees C by which each day's average temperature exceeds 0 degrees C (i.e., (Tmax + Tmin)/2 - 0 degrees C).

- `Tmax_Winter`: Average maximum temperature (in °C) for the winter season of the year of the `Observation_Date` (*December of previous year to February of the year of the Observation_Date*).

- `Tmax_Spring`: Average maximum temperature (in °C) for the spring season of the year of the `Observation_Date` (*March-May*). 

- `Tmax_Summer`: Average maximum temperature (in °C) for the summer season of the year of the `Observation_Date` (*June-August*).

- `Tmax_Fall`: Average maximum temperature (in °C) for the previous year's fall season (*September-November*), relative to the year of the `Observation_Date`. 

- `Tmax`: Maximum temperature (in °C) on the `Observation_Date`.

- `Tmin_Winter`: Average minimum temperature (in °C) for the winter season of the year of the `Observation_Date` (*December of previous year to February of the year of the Observation_Date*).

- `Tmin_Spring`:  Average minimum temperature (in °C) for the spring season of the year of the `Observation_Date` (*March-May*).

- `Tmin_Summer`: Average minimum temperature (in °C) for the summer season of the year of the `Observation_Date` (*June-August*).

- `Tmin_Fall`: Average minimum temperature (in °C) for the previous year's fall season (*September-November*), relative to the year of the `Observation_Date`.

- `Tmin`: Minimum temperature (in °C) on the `Observation_Date`.

- `Prcp_Winter`: Accumulated precipitation (in mm) for the winter season of the year of the `Observation_Date` (*December of previous year to February of the year of the Observation_Date*).

- `Prcp_Spring`: Accumulated precipitation (in mm) for the spring season of the year of the `Observation_Date` (*March-May*).

- `Prcp_Summer`: Accumulated precipitation (in mm) for the summer season of the year of the ^ (*June-August*).

- `Prcp_Fall`: Accumulated precipitation (in mm) for the previous year's fall season (*September-November*), relative to the year of the `Observation_Date`.

- `Prcp`: Precipitation (in mm) on the `Observation_Date`

- `Accum_Prcp`: Accumulated precipitation (in mm) *from January 1st* to the `Observation_Date`.

- `Daylength`: Number of seconds of daylight on the `Observation_Date`.

## Maps

To explore the data, it can be helpful to create a simple map using the `{sf}` and `{mapview}` packages. 

We can use a third variable, such as the altitude or the number of observations to help our understanding of the dataset.

### Main locations

First, we plot the five locations in the `cherry.csv` file

```{r}
#| label: map-main-locations
#| eval: true

cherry_coord <- cherry |> 
  group_by(location) |> 
  slice_tail(n = 1)

cherry_map <- cherry_coord |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(cherry_map, zcol = "alt")
```

### Japan

Next, we plot the map of all the locations in the `japan.csv` file

```{r}
#| label: map-japan
#| eval: true

crd_ch <- japan |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_ch <- crd_ch |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_ch, zcol = "n")
```

### Korea

Second, we plot the locations in the `south_korea.csv` file

```{r}
#| label: map-korea
#| eval: true

crd_korea <- korea |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_korea <- crd_korea |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_korea, zcol = "n")
```

### Switzerland

Third, we plot the locations in the `meteoswiss.csv` file

```{r}
#| label: map-meteoswiss
#| eval: true

crd_ch <- meteoswiss |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_ch <- crd_ch |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_ch, zcol = "n")
```

## Predictors

### POWER

The goal of this section is to find additional publicly-available data that will improve your predictions.

We use the `{nasapower}` package described here <https://docs.ropensci.org/nasapower/> to access the [NASA Prediction Of Worldwide Energy Resources (POWER) API](https://power.larc.nasa.gov/).

We use data from the **Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2)** which includes data at a daily frequency since January 1st, 1980 and simulated values untile December 31st, 2029.

The database is documented here <https://gmao.gsfc.nasa.gov/reanalysis/MERRA-2/>.

The POWER data base is documented here <https://power.larc.nasa.gov/docs/>.

The proper citation is here <https://gmao.gsfc.nasa.gov/reanalysis/MERRA-2/citing_MERRA-2/>.

The Parameter Dictionary includes all the available variables <https://power.larc.nasa.gov/beta/parameters/>.

We create a custom function called `get_power()` to repeatedly query the API with the `nasapower::get_power()` function.

```{r}
#| label: get_power

get_power <- function(vars, start = "1981-01-01", end = as.character(Sys.Date()), freq = "daily", lat, long) {
  
  power <- map2(
      .x = long,
      .y = lat,
      .f = function(long = .x, lat = .y) {
        res <- nasapower::get_power(
          community = "ag",
          lonlat = c(long, lat),
          pars = vars,
          dates = c(start, end),
          temporal_api = freq
        )
        return(res)
      }
    ) |>
    list_rbind()
  
  return(power)
}
```

Specify the parameters of the NASA POWER API request

We load the following variables:

- `WS10M`: Wind Speed at 10 Meters (m/s) 

- `FROST_DAYS`: Frost days. A frost day occurs when the 2m temperature cools to the dew point temperature and both are less than 0 C or 32 F.

- `TO3`: Total Column Ozone. The total amount of ozone in a column extending vertically from the earth's surface to the top of the atmosphere.

- `WS2M`: Wind Speed at 2 Meters. The average of wind speed at 2 meters above the surface of the earth.

- `DAY_HOURS`: Day Hours. The Climatological average number of hours when solar irradiance is present.

- `QV2M`: Specific Humidity at 2 Meters. The ratio of the mass of water vapor to the total mass of air at 2 meters.

- `RH2M`: Relative Humidity at 2 Meters. The ratio of vapor pressure to the saturation vapor pressure with respect to a plane surface of pure water, expressed in percent.

- `PRECTOTCORR`: Precipitation Corrected. The average MERRA-2 bias corrected total precipitation at the surface of the earth.

- `T2M`: Temperature at 2 Meters. The average air (dry bulb) temperature at 2 meters above the surface of the earth.

- `T2M_MAX:` Temperature at 2 Meters Maximum. The maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `T2M_MIN`: Temperature at 2 Meters Minimum. The minimum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `T2M_RANGE`: Temperature at 2 Meters Range. The difference between the minimum and maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest. Degrees Celsius (C).

- `CDD0`: Cooling Degree Days Above 0 C. The daily accumulation of degrees above a 0-degree-Celsius threshold when the daily mean temperature is above the threshold.

- `CDD18_3`: Cooling Degree Days Above 18.3 C. The daily accumulation of degrees above a 0-degree-Celsius threshold when the daily mean temperature is above the threshold.

- `HDD0`: Heating Degree Days Below 0 C. The daily accumulation of degrees below a 0-degree-Celsius threshold when the daily mean temperature is below the threshold.

- `HDD18_3`: Heating Degree Days Below 18.3 C. The daily accumulation of degrees below an 18.3-degree-Celsius threshold when the daily mean temperature is below the threshold.

```{r}
#| label: specify-vars

vars <- c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "CDD0",
  "CDD18_3",
  "HDD0",
  "HDD18_3",
  "FROST_DAYS",
  "PRECTOTCORR",
  "QV2M",
  "RH2M",
  "TO3",
  "PS",
  "WS10M"
)
```

We retrieve the **coordinates** which we will use to match to the nearest weather stations

```{r}
#| label: get-coordinates

crd_kyoto        <- distinct(kyoto, lat, long)
crd_liestal      <- distinct(liestal, lat, long)
crd_nyc          <- distinct(newyorkcity, lat, long)
crd_vancouver    <- distinct(vancouver, lat, long)
crd_washingtondc <- distinct(washingtondc, lat, long)

crd_japan      <- distinct(japan, lat, long)
crd_korea      <- distinct(korea, lat, long)
crd_meteoswiss <- distinct(meteoswiss, lat, long)

crd_npn1 <- npn_ind_obs |> 
  select(lat = Latitude, long = Longitude) |> 
  distinct()

crd_npn2 <- npn_int_obs |> 
  select(lat = Latitude, long = Longitude) |> 
  distinct()

crd_npn <- bind_rows(crd_npn1, crd_npn2) |> 
  distinct()
```

Load NASA POWER data for Kyoto, Japan

```{r}
#| label: load-main-nasa-power

power_kyoto <- get_power(vars, lat = crd_kyoto$lat, long = crd_kyoto$long)
power_kyoto$location <- "kyoto"

power_liestal <- get_power(vars, lat = crd_liestal$lat, long = crd_liestal$long)
power_liestal$location <- "liestal"

power_nyc <- get_power(vars, lat = crd_nyc$lat, long = crd_nyc$long)
power_nyc$location <- "newyorkcity"

power_vancouver <- get_power(vars, lat = crd_vancouver$lat, long = crd_vancouver$long)
power_vancouver$location <- "vancouver"

power_washingtondc <- get_power(vars, lat = crd_washingtondc$lat, long = crd_washingtondc$long)
power_washingtondc$location <- "washingtondc"

power_cherry <- bind_rows(
  power_kyoto,
  power_liestal,
  power_nyc,
  power_vancouver,
  power_washingtondc
)

write_csv(power_cherry, file = "data/power_cherry.csv")
```

We also load predictors for the additional data sets. Because these take a long time to load, instead load data from a previous run

```{r}
#| label: load-additional-nasa-power
#| eval: false

power_japan <- get_power(vars, lat = crd_japan$lat, long = crd_japan$long)
write_csv(power_japan, file = "data/power_japan.csv")

power_korea <- get_power(vars, lat = crd_korea$lat, long = crd_korea$long)
write_csv(power_korea, file = "data/power_korea.csv")

power_meteoswiss <- get_power(vars, lat = crd_meteoswiss$lat, long = crd_meteoswiss$long)
write_csv(power_meteoswiss, file = "data/power_meteoswiss.csv")

power_npn <- get_power(vars, lat = crd_npn$lat, long = crd_npn$long)
write_csv(power_npn, file = "data/power_npn.csv")
```

Let's have a look at the NASA POWER dataset for the five locations

```{r}
power_cherry <- read_csv("data/power_cherry.csv")
```

We change some the column names to lowercase letters

```{r}
colnames(power_cherry) <- tolower(colnames(power_cherry))

power_cherry <- power_cherry |> 
  rename(date = yyyymmdd, prcp = prectotcorr) |> 
  select(-c(lat, lon, year, mm, dd, doy)) |> 
  relocate(location, date)
```

## Feature engineering

### APRCP

Let's use the variable `prcp` to calculate a new variable, `aprcp`, the **accumulated precipitation (in mm) starting from January 1st of the year** of the observation date. 

```{r}
#| label: create-aprcp

power_cherry <- power_cherry |> 
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(aprcp = cumsum(prcp)) |> 
  ungroup() |> 
  select(-year)
```

### AGDD

Let's create the **Accumulated growing degree days (AGDD)** (in °C) as the **sum of growing degree days (GDD) since January 1st**, where `GDD` is calculated as the number of degrees C by which each day's average temperature exceeds `0` degrees C (i.e., `gdd` = (`ts_max` + `ts_min`)/2 - `0` degrees C).

Note that we are using a **base temperature of zero degrees Celsius** instead of `4.3` or `4.4` degrees Celsius which is commonly assumed to be the base temperature for Yoshino cherry trees.

```{r}
#| label: create-agdd
t_base <- 4.3

power_cherry <- power_cherry |> 
  mutate(t2m_range = (t2m_max + t2m_min) / 2) |> 
  mutate(gdd = pmax(t2m_range - t_base, 0)) |>
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(agdd = cumsum(gdd)) |> 
  ungroup() |>
  select(-c(year, t2m_range))
```

### ACDD

Let's also calculate **Chilling Degree Days (CDD)** and **Accumulated Chilling Degree Days (ACDD)**. 

Common choices for the **start dates** are September 1st or October 1st, or even the first date when temperatures fall below a certain **threshold**, usually `7` degrees Celsius. 

We calculate these variables **until December 31st of each year**. 

I again need to introduce a temporary variable `CDD2` which is zero when values are missing. I save the new data types `CDD` and `ACDD` in a temporary file `historic_acdd` and add it to the `historic_noaa` file.

```{r}
#| label: create-acdd

power_acdd <- power_cherry |> 
  mutate(year = year(date)) |> 
  filter(date >= paste0(year, "-09-01") & date <= paste0(year, "-12-31")) |> 
  mutate(cdd = pmax(0, 7 - t2m_min)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(acdd = cumsum(cdd)) |> 
  ungroup() |> 
  select(-year)

power_cherry <- power_cherry |> 
  left_join(y = power_acdd |> select(location, date, cdd, acdd), join_by(location, date))

power_cherry <- power_cherry |> 
  mutate(
    cdd = if_else(is.na(cdd), 0, cdd),
    acdd = if_else(is.na(acdd), 0, acdd)
  )
```

Let's plot the daily average maximum temperature obtained from the NASA POWER project

```{r}
#| label: fig-power-max-temp
#| fig-cap: "Average maximum temperature (°C)"

fig_tmax <- power_cherry |>
  ggplot(aes(x = date)) +
  geom_hline(yintercept = 0, colour = "grey") +
  geom_line(aes(y = t2m_max), colour = "#d62828") +
  labs(
    title = "Average maximum temperature (°C)", 
    x = NULL, y = "Degrees (°C)"
  ) +
  theme_minimal() +
  facet_grid(rows = vars(location))

print(fig_tmax)

ggsave(filename = "figures/fig_tmax.png", plot = fig_tmax, bg = "white")
```

Let's plot the daily average minimum temperature obtained from the NASA POWER project

```{r}
#| label: fig-power-min-temp
#| fig-cap: "Average minimum temperature (°C)"

fig_tmin <- power_cherry |>
  ggplot(aes(x = date)) +
  geom_hline(yintercept = 0, colour = "grey") +
  geom_line(aes(y = t2m_min), colour = "#00b4d8") +
  labs(
    title = "Average minimum temperature (°C)",
    x = NULL, y = "Degrees (°C)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_tmin)

ggsave(filename = "figures/fig_tmin.png", plot = fig_tmin, bg = "white")
```

Plot `prcp` from NASA POWER:

```{r}
#| label: fig-avg-prcp
#| fig-cap: "Average preciptation"

fig_prcp <- power_cherry |>
  ggplot(aes(x = date, y = prcp)) +
  geom_line(colour = "#4361ee") +
  labs(
    title = "Average preciptation (in mm)",
    x = NULL, y = "Average preciptation (in mm)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_prcp)

ggsave(filename = "figures/fig_prcp.png", plot = fig_prcp, bg = "white")
```

Let's plot `APRCP`:

```{r}
#| label: fig-aprcp
#| fig-cap: "Average aggregated preciptation since January 1st (in mm)"

fig_aprcp <- power_cherry |>
  ggplot(aes(x = date, y = aprcp)) +
  geom_line(colour = "#4361ee") +
  labs(
    title = "Average aggregated preciptation since January 1st (in mm)",
    x = NULL, y = "Average agregated preciptation (in mm)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_aprcp)

ggsave(filename = "figures/fig_aprcp.png", plot = fig_aprcp, bg = "white")
```


Let's plot the Growing Degree Days (GDD) from NASA POWER

```{r}
#| label: fig-gdd
#| fig-cap: "Growing Degree Days (GDD) calculated from January 1st"

fig_gdd <- power_cherry |>
  ggplot(aes(x = date, y = gdd)) +
  geom_line(colour = "#d62828") +
  labs(
    title = "Growing Degree Days (GDD)",
    subtitle = "Calculated from January 1st",
    x = NULL, y = "Growing Degree Days (GDD)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_gdd)

ggsave(filename = "figures/fig_gdd.png", plot = fig_gdd, bg = "white")
```

Let's plot Accumulated Growing Degree Days (AGDD) from NASA POWER

```{r}
#| label: fig-agdd
#| fig-cap: "Accumulated Growing Degree Days (AGDD) calculated from January 1s"

fig_agdd <- power_cherry |>
  ggplot(aes(x = date, y = agdd)) +
  geom_step(colour = "#d62828") +
  labs(
    title = "Accumulated Growing Degree Days (AGDD)", 
    subtitle = "Calculated from January 1st",
    x = NULL, y = "Accumulated Growing Degree Days (AGDD)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_agdd)

ggsave(filename = "figures/fig_agdd.png", plot = fig_agdd, bg = "white")
```


Let's plot the Chilling Degree Days (CDD) from NASA POWER

```{r}
#| label: fig-plt-cdd
#| fig-cap: "Chilling Degree Days (CDD) calculated from September 1st"

fig_cdd <- power_cherry |>
  ggplot(aes(x = date, y = cdd)) +
  geom_line(colour = "#00b4d8") +
  labs(
    title = "Chilling Degree Days (CDD)",
    subtitle = "Calculated from September 1st",
    x = NULL, y = "Chilling Degree Days",
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_cdd)

ggsave(filename = "figures/fig_cdd.png", plot = fig_cdd, bg = "white")
```

Let's plot Accumulated Chilling Degree Days (ACDD) from NASA POWER

```{r}
#| label: fig-plt-acdd
#| fig-cap: "Accumulated Chilling Degree Days (ACDD) calculated from September 1st"

fig_acdd <- power_cherry |>
  ggplot(aes(x = date, y = acdd)) +
  geom_line(colour = "#00b4d8") +
  labs(
    title = "Accumulated Chilling Degree Days (ACDD)",
    subtitle = "Calculated from September 1st",
    x = NULL, y = "Chilling Degree Days",
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_acdd)

ggsave(filename = "figures/fig_acdd.png", plot = fig_acdd, bg = "white")
```

## Temporal aggregation

```{r}
#| label: aggregate-to-season

df_seasons <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .month = month(date)) |> 
  mutate(
    season = case_when(
      .month %in% c(12, 1, 2) ~ "winter",
      .month %in% c(3, 4, 5) ~ "spring",
      .month %in% c(6, 7, 8) ~ "summer",
      .month %in% c(9, 10, 11) ~ "winter"
    )
  ) |> 
  mutate(.year = if_else(.month == 12, .year - 1, .year)) |>  
  mutate(.year_season = paste0(.year, "-", season)) |> 
  relocate(location, date, .year, .month, season, .year_season, variable, value) |> 
  group_by(location, .year, season, .year_season, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, season),
    names_glue = "{variable}_{season}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year) |> 
  filter(year > 1980)

cherry_seasons <- cherry |> 
  inner_join(y = df_seasons, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))

cherry_seasons
```


```{r}
#| label: aggregate-to-monthly

df_months <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .month = month(date)) |> 
  mutate(.year_month = paste0(.year, "-", .month)) |> 
  relocate(location, date, .year, .month, .year_month, variable, value) |> 
  group_by(location, .year, .month, .year_month, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .month),
    names_glue = "{variable}_{.month}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_months <- cherry |> 
  inner_join(y = df_months, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))

cherry_months
```

```{r}
#| label: aggregate-to-weekly

df_weeks <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .week = week(date)) |> 
  mutate(.year_week = paste0(.year, "-", .week)) |> 
  relocate(location, date, .year, .week, .year_week, variable, value) |> 
  group_by(location, .year, .week, .year_week, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .week),
    names_glue = "{variable}_{.week}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_weeks <- cherry |> 
  inner_join(y = df_weeks, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))

cherry_weeks
```

```{r}
#| label: aggregate-to-daily

df_days <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .day = day(date)) |> 
  mutate(.year_month = paste0(.year, "-", .day)) |> 
  relocate(location, date, .year, .day, .year_month, variable, value) |> 
  group_by(location, .year, .day, .year_month, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .day),
    names_glue = "{variable}_{.day}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_days <- cherry |> 
  inner_join(y = df_days, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))

cherry_days
```


## AccuWeather

It is likely that the weather in March and April is a very important predictor for the peak bloom date of cherry blossoms.

However, because the submission date is February 28, we won't have any data available for these two months.

We can either train our model exclusively on data from January 1st 1981 to January 27th of the current year.

Or we can include data for April and March and use Weather forecasts.

Any weather forecast can become a flower forecast by applying the law of the flowering plants. In this section, we use the AccuWeather forecast to predict the day a hypothetical lilac will bloom in Brussels in 2023. AccuWeather forecasts daily maximum and minimum temperatures three months into the future. We do not evaluate the quality of these forecasts. The purpose of this section is to simply convert them into flower forecasts.

We use the AccuWeather forecast as it appeared on the webpage [AccuWeather.com](https://www.accuweather.com/en/be/brussels/27581/january-weather/27581?year=2023) on February 19, 2023. AccuWeather reports the forecast for each month on a separate webpage. For reproducibility, we saved each page on the [Internet Archive](https://web.archive.org/web/20230219151906/https://www.accuweather.com/en/be/brussels/27581/january-weather/27581?year=2023&unit=c). The following R code creates the function `get_weather_table()` to retrieve each page we saved, extract the forecast contained within that page, and arrange the data as a tibble. The `get_weather_table()` function combines several functions from the `{rvest}` package, which is yet another member of the `{tidyverse}`. In particular, the forecast on each page is contained within the div `monthly-calendar` and can be extracted with the `html_nodes()` and `html_text2()` functions.

Applying the `get_weather_table()` function to the url for each page yields a five column tibble `temp_br`, with columns defined in the same way as the tibble `temp`, discussed in previous sections. The first 10 rows are below; the data are also available on the author's [GitHub](https://raw.githubusercontent.com/jauerbach/miscellaneous/main/temp_br.csv).

We now predict the day the lilacs will bloom. The R code below uses the `doy_prediction` and `doy_last_frost` functions created in earlier sections and displays the prediction in a table. 

At the time of our writing, the predicted date is **April 19**. The forecast is easily updated by providing the url to the updated AccuWeather webpage. (You might use the url [https://web.archive.org/save](https://web.archive.org/save) to save a webpage to the Internet Archive to ensure your work is reproducible.)

## Prediction

A simple method to predict peak bloom date in the future is to fit a least-squares line through the observed dates and extrapolate the regression function.

We want to have a separate line for each location, hence we tell R to estimate *interaction* effects. Note that R will automatically exclude `NA`s from the regression.

We only use data **from 1880** to fit the trends, as prior data may not be as reliable/relevant.

Thus we fit a model **without intercept** and **five dummy variables** for every location, and we create **interaction effects**. 

This case is equivalent to fitting five separate models with a linear time trend, where each location has a different intercept and a different slope.

```{r}
#| label: fit-linear-model

# Fit a linear time trend without intercept, dummy variables and interaction effect
ls_fit <- lm(
  formula = bloom_doy ~ 0 + location + location:year,
  data = cherry, 
  subset = year >= 1880
)
```

This simple linear regression functions suggest a trend toward earlier peak bloom at the sites.

```{r}
#| label: summary-linear-model
summary(ls_fit)
```

Unsurprisingly, neither the intercept nor the time trend are statistically significant for Vancouver and New York City, where we have few observations.

We can compute predictions for both the historical observations and for the current year with the `predict()` function. We first use the `expand_grid()` function to create all possible combinations for the years between 1880 and 2025 and the five locations, so that we have the input variables for both the fitted values for the past years and the prediction for the current year

```{r}
#| label: create-grid

cherry_grid <- expand_grid(
  location = unique(cherry$location),
  year = 1880:2025
)
```

Then we use these input variables to create both point estimates and 90% confidence intervals with `predict()`

```{r}
#| label: create-predictions

predictions <- cherry_grid |>
  bind_cols(predict(
    object = ls_fit,
    newdata = cherry_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr)
```

Plot only the fitted values, i.e., the predictions for the actual observations between 2015 and 2024

```{r}
#| label: fig-fitted
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from simple linear regression models fitted to
#|   five sites.

cherry |>
  right_join(predictions, by = c("year", "location")) |>
  filter((location == "vancouver" & year >= 2015) | (location != "vancouver" & year >= 2015)) |>
  filter((location == "newyorkcity" & year >= 2019) | (location != "newyorkcity" & year >= 2015)) |> 
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that the slope of the time trend has become steeper since 2015. If we assume that this trend continues, our prediction for the current year will be too high.

We create a function called `day_to_date()` that converts `bloom_doy` into an ISO date

```{r}
#| label: day_to_date

#' Small helper function to convert the day of year to the actual date.
#'
#' @param year year as an integer
#' @param doy day of the year as integer (1 means January 1st)
#' @return date string

doy_to_date <- function(year, doy) {
  strptime(paste(year, doy, sep = "-"), "%Y-%j") |> # create date object
    strftime("%Y-%m-%d") # translate back to date string in ISO 8601 format
}
```

Since we treated the outcome variable as continuous, we need to use `floor()` and `ceiling()` to create upper and lower bounds for our point estimate, for which we use `round()`.

Based on this simple model, the peak bloom dates at these sites in 2025 are predicted to occur on

```{r}
#| label: tbl-predictions
#| tbl-cap: "Predicted peak bloom dates at the four sites"

predictions |>
  filter(year == 2025) |>
  mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  ) |> 
  kable() |> 
  kable_styling()
```

### Averaging

We have only a few observations for Vancouver, BC and for New York City. Although the climate in Vancouver and NYC is different from the other locations, the simplest way to borrow information from the other locations is to average across these three sites.

Hence, we want to fit a straight line through the peak bloom dates, ignoring the actual site. That is, we fit a simple linear time trend on all available observations, but we want to use larger **weights** for observations from the target sites.

#### Vancouver

To run Weighted Least Squares (WLS), the `weights` inside `stats::lm()` argument must be a vector of the same length as the input data and we can use the names of the `location` column to create it. 

We could have created a new variable explicitly with:

```{r}
#| eval: false

cherry$weight <- (cherry$location == "vancouver") + 0.2 * (cherry$location != "vancouver")

cherry$weight <- NULL
```

So the weight put on observations from Vancouver are while observations from other locations receive a weight of only one fifth.

```{r}
#| label: fit-wls-vancouver

ls_fit_for_van <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "vancouver") + 0.2 * (location != "vancouver")
)
```

Create a grid with the input variables for Vancouver only

```{r}
#| label: grid-vancouver

vancouver_grid <- tibble(location = "vancouver", year = 2008:2025)
```

Then we use these inputs to create predictions for Vancouver, 

```{r}
#| label: predict-vancouver

predictions_vancouver <- vancouver_grid |>
  bind_cols(predict(
    object = ls_fit_for_van,
    newdata = vancouver_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  )
```

Not surprisingly, the predicted peak bloom date for Vancouver and NYC is now very similar to the other 3 sites:

```{r}
#| label: tbl-predictions-vancouver
#| tbl-cap: "Predictions for Vancouver"

predictions_vancouver |> 
  kable() |> 
  kable_styling()
```

Due to **rounding** and the **slope** not being very steep, the predictions are almost the same for every year.

We can plot the fitted values against the actual data from previous years

```{r}
#| label: fig-fitted-vancouver
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for Vancouver using data from all four sites.

cherry |>
  right_join(y = predictions_vancouver, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2008:2025) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that without including weather data, we are not able to capture the unusually warm springs in Vancouver, such as the one from 2010.

We have made this very cautious prediction since we don't know much about the conditions in Vancouver.

If we are satisfied with these weighted predictions for Vancouver, we can use them to replace the predictions that were made using only data from Vancouver.

```{r}
#| label: replace-vancouver-predictions

predictions <- predictions |>
  filter(location != "vancouver") |>
  bind_rows(predictions_vancouver)
```

#### New York

We once again fit a model on the whole data set, weighting observations from New York City with one and observations from other locations with one fifth.

```{r}
#| label: tbl-predictions-nyc
#| tbl-cap: "Predictions for New York City"

ls_fit_for_nyc <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "newyorkcity") + 0.2 * (location != "newyorkcity")
)

# Add inputs since 2019 and the current year
nyc_grid <- tibble(location = "newyorkcity", year = 2019:2025)

predictions_nyc <- nyc_grid |>
  bind_cols(predict(
    object = ls_fit_for_nyc,
    newdata = nyc_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr) |> 
    mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  )

predictions_nyc |> 
  kable() |> 
  kable_styling()
```

Due to rounding, and because the slope is not very steep, the predictions are the same for every year.

We plot the fitted values for New York City and the actual observations

```{r}
#| label: fig-fitted-nyc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for NYC using data from all five sites.

cherry |>
  right_join(predictions_nyc, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2019:2024) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

Again, this model does not capture the steepness of the slope of the time trend accurately.

If we are satisfied with the averaged predictions we can replace the predictions that were made using only data from New York City

```{r}
#| label: replace-nyc-predictions

predictions <- predictions |>
  filter(location != "newyorkcity") |>
  bind_rows(predictions_nyc)
```


## Submission

To submit your entries, enter the predicted bloom dates from your models in the **submission form**.

```{r}
#| label: tbl-predictions-for-submission
#| tbl-cap: "Predictions for Submission"

predictions |>
  filter(year == 2025) |>
  mutate(prediction_date = strptime(paste(year, prediction), "%Y %j") |> as_date()) |>
  kable() |> 
  kable_styling()
```

**NOTE**: The question is if we need to round the point estimates or if we can submit partial days.

## Using predictors

### Seasonal predictors

A simple model may average weather data of the seasons into account:

Plot `t2m_max_winter`

```{r}
#| label: fig-avg_tmax
#| fig-cap: "Average maximum winter temperature (Dec. 1st of the previous year until end of February)"

cherry_seasons |>
  ggplot(aes(x = year, y = t2m_max_winter)) +
  geom_line() +
  labs(x = NULL, y = "Average winter temperature (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```

We fit a linear model with different intercepts and time trends for each location and a weather effect. We can add interaction effects to the temperature data too if we want.

```{r}
#| label: fit-lin-season

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_seasons
  )
```

Create fitted values

```{r}
#| label: predictions-avg_tmax

mean_pred <- predict(object = lm_fit, new_data = cherry_seasons)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-avg_tmax
#| fig-cap: "Predictions from seasonal predictors"

cherry_seasons |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

### Monthly predictors

Plot the average maximum temperature in February

```{r}
cherry_months |>
  select(year, location, bloom_doy, t2m_max_2) |> 
  ggplot(aes(x = year, y = t2m_max_2)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum temperature (°C) in February") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```


Now let's use a linear model with monthly data

```{r}
#| label: fit-lin-months

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_months
  )
```

Create fitted values

```{r}
#| label: predictions-avg_tmax

mean_pred <- predict(object = lm_fit, new_data = cherry_months)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-months
#| fig-cap: "Predictions from monthly predictors"

cherry_months |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

### Weekly predictors

Now let's use a linear model with weekly data

```{r}
#| label: fit-lin-weeks

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_weeks
  )
```

Create fitted values

```{r}
#| label: predictions-avg_tmax

mean_pred <- predict(object = lm_fit, new_data = cherry_weeks)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-months
#| fig-cap: "Predictions from monthly predictors"

cherry_weeks |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

## Daily predictors

Now let's use a linear model with daily data

```{r}
#| label: fit-lin-days

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_days
  )
```

Create fitted values

```{r}
#| label: predictions-avg_tmax

mean_pred <- predict(object = lm_fit, new_data = cherry_days)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-months
#| fig-cap: "Predictions from monthly predictors"

cherry_days |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```


Let's fit a random forest

```{r}
cores <- parallel::detectCores()
cores

# Split the data with {rsamples}
set.seed(123)
cherry_split <- initial_split(data = cherry_weeks)
cherry_train <- training(cherry_split)
cherry_test  <- testing(cherry_split)

# Create folds with {rsamples}
set.seed(345)
folds <- vfold_cv(cherry_train, v = 10)
folds

# Create model with {parsnip}
rf_mod <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
  ) |> 
  set_engine("ranger", num.threads = cores, importance = "impurity") |> 
  set_mode("regression")

# show what will be tuned
extract_parameter_set_dials(rf_mod)

# Create worfklow with {workflows}
rf_workflow <- workflow() |> 
  add_model(rf_mod) |> 
  add_formula(bloom_doy ~ .)

# Create a tuning grid with {dials}
tree_grid <- grid_regular(
  mtry(range = c(1, 9)),
  min_n(),
  levels = 5
)

# Train the model on the folds
tree_res <- rf_workflow |> 
  tune_grid(
    resamples = folds,
    grid = tree_grid,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(rmse, mae)
  )

tree_res |> 
  collect_metrics()

tree_res |> 
  show_best(metric = "rmse")

best_tree <- tree_res |> 
  select_best(metric = "rmse")

# Finalize the model
final_wf <- rf_workflow |> 
  finalize_workflow(parameters = best_tree)

final_wf

# Fit the model best model on all the training data
# with the {tune} function last_fit()
final_fit <- final_wf |> 
  last_fit(cherry_split)

final_fit |> 
  collect_metrics()

final_fit |> 
  collect_predictions() |> 
  ggplot(mapping = aes(x = bloom_doy, y = .pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  coord_equal(ratio = 1) +
  theme_bw()

# To predict new data points, use {workflows} extract_worfklow()
final_tree <- extract_workflow(final_fit)

final_tree

final_tree |> 
  extract_fit_parsnip() |> 
  vip()
```




The following plot shows a comparison of predictions for Vancouver using the two methods described in this demo.

```{r}
#| label: fig-pedictions-vancouver
#| fig-cap: "Predicted peak bloom (days since Jan 1st) for Vancouver"

predictions_vancouver |>
  left_join(predictions_from_temp,
    by = c("location", "year")
  ) |>
  select(year, pred_temporal = prediction, pred_temp = pred_bloom) |>
  pivot_longer(cols = -year) |>
  mutate(name = if_else(name == "pred_temporal",
    "Method 1: location-based model",
    "Method 2: temperature-based model"
  )) |>
  ggplot() +
  aes(x = year, y = value, linetype = name) +
  geom_line() +
  scale_x_continuous(breaks = 2008:2025) +
  labs(
    x = NULL, linetype = "",
    y = "Predicted peak bloom (days since Jan 1st) for Vancouver"
  ) +
  theme_light() +
  theme(legend.position = "bottom")
```
