---
title: "Peak Bloom Prediction"
author: "econmaett"
lang: en-US
format:
  html:
    embed-resources: true
fig-cap-location: bottom
tbl-cap-location: bottom
toc: true
editor: source
---

```{r}
#| label: setup
#| include: false
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, error = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "80%"
)
```

## Packages

First we load all the packages necessary to run this analysis.

```{r}
#| label: load-pkgs
#| output: false
#| warning: false
#| message: false

invisible(Sys.setlocale("LC_ALL","English"))

library(tidymodels)
library(tidyverse)
library(arrow)
library(bonsai)
library(rules)
library(finetune)
library(infer)
library(probably)
library(kernlab)
library(corrr)
library(rpart.plot)
library(vip)
library(gt)
library(nasapower)
library(openmeteo)

tidymodels_prefer()
```

## Data

Next, we load the data for this analysis.

### Clean data

The data for the five sites is provided in CSV format. Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site. The six columns in each data file are:
- `location`: a human-readable location identifier (`string`).
- `lat`: (approximate) latitude of the cherry trees (`double`).
- `long`: (approximate) longitude of the cherry trees (`double`).
- `alt`: (approximate) altitude of the cherry trees (`double`).
- `year`: year of the observation (`integer`).
- `bloom_date`: date of peak bloom of the cherry trees ([ISO 8601](https://xkcd.com/1179/) date `string`). The "peak bloom date" may be defined differently for different sites
- `bloom_doy`: days since January 1st until peak bloom (`integer`). January 1st is `1`.

#### Kyoto

The file `kyoto.csv` has been obtained from Yasuyuki Aono at <http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/>. The geographical location at `35.01198°N` (latitude), `135.6761°E` (longitude), `44` meters above sea levels (altitude) roughly corresponds to the **Nakanoshima** area of the **Arashiyama Park** in Kyoto, Japan. Peak bloom is defined as **80%** of flowers being in full bloom.

The peak bloom date of the **Prunus jamasakura (Japanese mountain cherry)** is determined by a local news paper in Arashiyama (Kyoto, Japan). Data prior to 1888 is extracted from various descriptions or estimated:
- Data from the 9th to the 14th centuries was acquired and analyzed by Aono and Saito (2010; International Journal of Biometeorology, 54, 211-219).
- Phenology for 15th to 21st centuries was acquired and analyzed by Aono and Kazui (2008; International Journal of Climatology, 28, 905-914).

Note that this data set includes implicit missing values (`NA`s). [Chapter 18 "Missing Values"](https://r4ds.hadley.nz/missing-values) of Hadley Wickham's book "R for Data Science (2e)" describes strategies for dealing with implicit `NA`s.

```{r}
#| label: load-kyoto
#| output: false
#| warning: false
#| message: false

kyoto <- read_csv("data/kyoto.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(country = "japan") |> 
  fill(location, lat, long, alt, .direction = "downup")
```

Plot the time series for Kyoto.

```{r}
#| label: fig-kyoto
#| fig-cap: "Peak cherry blossom days in Kyoto, Japan"
#| warning: false
#| message: false

fig_kyoto <- kyoto |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 0.5) +
  geom_point(color = "#ff0054", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(800, 2000, by = 100), NA)) +
  scale_y_continuous(expand = c(0, 0)) +
  labs(
    title = "DOY of peak blossom (80%) in Kyoto, Japan from 812 to 2024", 
    subtitle = "Prunus jamasakura (Japanese mountain cherry)\n35.0120°N 135.6761°E	44m",
    x = NULL, y = NULL,
    caption = "Source: Yasuyuki Aono\nhttp://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/"
  ) +
  theme_light()

print(fig_kyoto)

ggsave(filename = "figures/fig_kyoto.png", plot = fig_kyoto, bg = "white")
```

Here we have used **locally estimated scatterplot smoothing (LOESS)** or **locally weighted scatterplot smoothing (LOWESS)** to highlight how the average peak bloom day changed over the years. 

MeteoSwiss suggests using a **Gaussian Kernel** with a *20-year window* to extract the trend, which is equivalent to using a **low-pass filter** that shuts down all higher frequencies.

It also looks like the observations exhibit **heteroskedasticity**, that is, the variance changes over time, and it looks like the variance has become more narrow over time.

We can use **changepoint detection** methods to test for the changing time trend and the changing variance.

Assuming that the data in the past is correct, there do seem to be seasonal cycles, perhaps connected to **solar activity** or the **tilt of the earth**.

There is a clear trend towards earlier blooming, perhaps connected to anthropogenic climate change and urbanization. Studies attribute about half of the earlier blooming times in recent years to the **urban heat island effect**, and the other half to **climate change**.

Modelling the local level and the trend would be useful if we are to forecast a couple years into the future, since an individual prediction may be incorrect, but on average, they may be good. 

#### Liestal

The file `liestal.csv` contains observations for **a single wild cherry (prunus avium) tree** in **Liestal-Weideli**, at `7.4814°N` (latitude), `7.730519°E` (longitude) and `350` meters above sea levels (altitude). The **peak bloom date** is defined as the day when **25%** of the blossoms are in full bloom. The date is determined by [MeteoSwiss](https://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html). The data is provided by **Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss**. 

```{r}
#| label: load-liestal
#| output: false
#| warning: false
#| message: false

liestal <- read_csv("data/liestal.csv") |> 
  mutate(country = "switzerland")
```

Plot the time series for Liestal, Switzerland.

```{r}
#| label: fig-liestal
#| fig-cap: "Peak cherry blossom in Liestal-Weideli, Switzerland"
#| warning: false
#| message: false

fig_liestal <- liestal |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(
    title = "DOY of peak blossom (25%) in Liestal-Weideli, Switzerland from 1894 to 2024",
    subtitle = "Prunus avium (wild cherry)\n47.4814°N 7.730519°E 350m",
    x = NULL, y = NULL,
    caption = "Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss\nhttps://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html"
  ) +
  theme_light()

print(fig_liestal)

ggsave(filename = "figures/fig_liestal.png", plot = fig_liestal, bg = "white")
```

For Liestal you can see the impact of the infamous heat summer of 2003 which I remember from my childhood, although interestingly, there's another record in 1990.

#### New York

The file `newyorkcity.csv` contains **only a single observation** for the year 2024. It concerns the **Yoshino cherry (prunus × yedoensis)** tree in **Washington Square Park**, New York City, at `40.7304 °N` (latitude), `-73.99809°E` (longitude) and `8.5` meters above sea level (altitude). Peak bloom is defined as **70%** of flowers being in full bloom.

Luckily, there are some historical observations dating *back to 2019* in the data provided by **USA-NPN**. The Washington Square Park has site id `32789` and the **Yoshino** cherry you should predict has species id `228`.

This file now contains individual observations of the phenophase and not the actual peak bloom date. We use the variable `Intensity_Value` which is either `-9999`, `95% or more`, `Less than 5%`, `25-49%`, `75-94%`, `50-74%`. Since according to the variable `Individual_ID`, only one tree, `183140` is being observed, we can use the first date where `Intensity_Value` is `50-74%` as our `bloom_doy`.

```{r}
#| label: load-newyorkcity
#| output: false
#| warning: false
#| message: false

newyorkcity <- read_csv("data/nyc.csv") |> 
  mutate(country = "unitedstates")

nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789, Species_ID == 228) |> 
  mutate(Observation_Date = mdy(Observation_Date))

nyc_data <- nyc_data_npn |>
  arrange(Observation_Date) |>
  mutate(year = year(Observation_Date)) |>
  group_by(year) |>
  summarize(
    first_flower_index = min(which((Intensity_Value %in% c("75-94%", "95% or more")))),
    bloom_date = as_date(strftime(Observation_Date[first_flower_index], format = "%Y-%m-%d")),
    bloom_doy = Day_of_Year[first_flower_index],
    .groups = "drop"
  ) |>
  filter(!is.na(bloom_doy)) |>
  select(-first_flower_index) |>
  mutate(country = "unitedstates", location = "newyorkcity") |> 
  arrange(year)

newyorkcity <- newyorkcity |>
  bind_rows(nyc_data |> filter(year != 2024)) |>  
  fill(lat, long, alt)
```

#### Vancouver

The file `vancouver.csv` contains observations for cherry trees of the cultivar  **Prunus × yedoensis "Akebono" (daybreak cherry)** in **Maple Grove Park**, Vancouver, BC located approximately at `49.2236916°N` (latitude), `-123.1636251°E` (longitude), `24` meters above sea levels (altitude). Peak bloom is defined as **70%** of flowers being in full bloom.

Unfortunately, there are **only three observations** for 2022, 2023, and 2024, respectively. But casual observations have been recorded in the form of **photos posted** to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/). You can **search the forum** for the keywords `"Akebono"` (i.e., the name of the cultivar) and `"Maple Grove Park"` (i.e., the location of the trees).

We can do some additional plausibility checks. For example, spring 2010 was indeed unusually warm, [causing cherry blossoms to bloom during the Vancouver 2010 Winter Olympics](https://www.upi.com/News_Photos/view/upi/33393f9d20159ecb347de9aa1d670922/2010-Winter-Olympics-in-Vancouver/). National Geographic claimed it was the [warmest Olympics to date](https://www.nationalgeographic.com/adventure/article/100212-vancouver-2010-warmest-winter-olympics).

```{r}
#| label: load-vancouver
#| output: false
#| warning: false
#| message: false

vancouver <- read_csv("data/vancouver.csv") |> 
  mutate(country = "canada")

vancouver_forum <- tribble(
  ~ year, ~ bloom_date,
  2021, "2021-03-31", # Start: 2021-03-21, Full: 2021-03-31, Full: 2021-04-07
  2020, "2020-04-03", # Full bloom
  2018, "2018-03-18", # Full bloom
  2017, "2017-04-11", # Full bloom 
  2015, "2015-03-19", # Full bloom
  2011, "2011-04-01", # Full bloom was over on 2011-04-15
  2010, "2010-03-14", # 60% bloom on 2010-03-07
  2008, "2008-04-14" # Start: 2008-04-06, Full: 2008-04-14, finished: 2008-04-22
  ) |> 
  mutate(
    bloom_date = as_date(bloom_date),
    bloom_doy = as.numeric(1 + as_date(bloom_date) - as_date(paste0(year, "-01-01"))),
    location = "vancouver",
    country = "canada"
  )

vancouver <- vancouver |> 
  bind_rows(vancouver_forum) |> 
  fill(lat, long, alt) |> 
  arrange(year)
```

#### Washington

The file `washingtondc.csv` has been obtained from the EPA's Climate Change Indicators in the United States <https://www.epa.gov/climate-indicators/cherry-blossoms>.

The geographical location at `38.88535°N` (latitude), `-77.03863°E` (longitude) and `0` meters above sea level (altitude) correspond to the location of the [Tidal Basin in Washington, DC](https://www.nps.gov/articles/dctidalbasin.htm).

The peak bloom date is defined as the day when **70%** of the **Prunus × yedoensis "Somei-yoshino"** are in full bloom, as determined by the [National Park Service](https://www.nps.gov/subjects/cherryblossom/bloom-watch.htm).

```{r}
#| label: load-washingtondc
#| message: false
#| warning: false
#| output: false

washingtondc <- read_csv("data/washingtondc.csv") |> 
  mutate(country = "unitedstates")
```

Let's plot the data for Washington D.C.

```{r}
#| label: fig-washingtondc
#| fig-cap: "Peak cherry blossom in Washington DC, USA"
#| warning: false
#| message: false

fig_washingtondc <- washingtondc |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 0.75) +
  geom_point(color = "#ff0054", alpha = 0.5, size = 1.5) +
  scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(
    title = "DOY of peak blossom (70%) in Washington DC, USA from 1921 to 2024",
    subtitle = "Yoshino cherry (prunus × yedoensis) 'Somei-yoshino'\n38.8853°N –77.0386°E 0m",
    x = NULL, y = NULL,
    caption = "Environmental Protection Agency (EPA) Climate Change Indicators in the United States\nhttps://www.epa.gov/climate-indicators/cherry-blossoms"
  ) +
  theme_light()

print(fig_washingtondc)

ggsave(filename = "figures/fig_washingtondc.png", plot = fig_washingtondc, bg = "white")
```

Now that we have loaded the datasets for the five locations, we can combine them into a single CSV file called `cherry.csv`.

```{r}
#| label: combine-cherry-csv
#| output: false
#| warning: false
#| message: false

cherry <- bind_rows(kyoto, liestal, newyorkcity, vancouver, washingtondc) |> 
  relocate(year, location, country, lat, long, alt, bloom_date)
```

Show the time series for our five main locations next to each other.

```{r}
#| label: fig-cherries_1880
#| warning: false
#| message: false
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at five different sites.

df_cherry_1880 <- cherry |> 
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  filter(year >= 1880)

fig_cherries <- ggplot(data = df_cherry_1880, aes(year, bloom_doy)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 5), colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab") +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 40), NA)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(
    title = "Peak bloom (since Jan 1st)",
    x = NULL, 
    y = NULL
  ) +
  theme_light()

print(fig_cherries)

ggsave(filename = "figures/fig_cherries.png", plot = fig_cherries, bg = "white")
```

### Additional data

Additional data sets are provided with data from sites other than the main sites relevant for the competition.

#### Japan

The data file `japan.csv` is provided by the [Japanese Meteorological Agency (JMA)](https://www.data.jma.go.jp/sakura/data/) and contains recorded peak bloom dates for cherry trees at various sites across Japan. The sample trees are located within a 5km radius of the location indicated in the data file. 

Load the data set, make implicit missing values explicit and change the `location` variable to only include the city names in lowercase letters. I choose one location with an average peak bloom date similar to that of Vancouver, Canada.

```{r}
#| label: read-japan
#| warning: false
#| message: false
#| output: false

japan <- read_csv("data/japan.csv") |>  
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Japan/")) |> 
  mutate(location = tolower(location), country = "japan") |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))

fukuoka <- japan |>
  filter(location == "fukuoka", alt == 2.5)
```

Plot peak bloom dates for selected locations in Japan

```{r}
#| label: fig-bloom-japan
#| fig-cap: "Peak bloom dates for selected locations Japan"
#| warning: false
#| message: false

fig_japan <- japan |>
  filter(location %in% c("fukuoka")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 20), NA)) +
  facet_wrap(facets = vars(location)) +
  labs(
    title = "Peak bloom (since Jan 1st) for selected locations in Japan",
    subtitle = NULL,
    x = NULL, y = NULL,
    caption = "Japanese Meteorological Agency (JMA)\nhttps://www.data.jma.go.jp/sakura/data/"
  ) +
  theme_light()

print(fig_japan)

ggsave(filename = "figures/fig_japan.png", plot = fig_japan, bg = "white")
```

#### MeteoSwiss

The data file `meteoswiss.csv` contains peak bloom dates for various sites across Switzerland, obtained from <https://opendata.swiss/en/dataset/phanologische-beobachtungen>.

Note that when you use the `{utils}` function `read.csv()` you need to specify the option `encoding = "UTF-8"` to allow for special characters that are common in the German language.

```{r}
#| label: load-meteoswiss
#| warning: false
#| message: false
#| output: false

meteoswiss <- read_csv("data/meteoswiss.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Switzerland/")) |> 
  mutate(location = tolower(location), country = "switzerland") |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location)) |> 
  mutate(location = str_remove(location, "basel-")) |> 
  mutate(location = str_remove(location, "zürich-"))

binningen <- meteoswiss |>
  filter(location == "binningen", alt == 315)

laufenburg <- meteoswiss |>
  filter(location == "laufenburg", alt == 330)

albisgutli <- meteoswiss |>
  filter(location == "albisgütli", alt == 530)

meteoschweiz <- meteoswiss |>
  filter(location == "meteoschweiz", alt == 555)

witikon <- meteoswiss |>
  filter(location == "witikon", alt == 620)
```

The locations have up to 70 observations, but most are at very high altitudes, much higher than the five main locations. I therefore select two locations which have a similar altitude to Liestal and are in close geographical proximity. I also select the three locations closest to my home.

Plot the data for selected locations in Switzerland.

```{r}
#| label: fig-bloom-japan
#| fig-cap: "Peak bloom dates for selected locations Japan"
#| warning: false
#| message: false

fig_meteoswiss <- meteoswiss |>
  filter(location %in% c("binningen", "laufenburg", "albisgütli", "meteoschweiz", "meteoschweiz", "witikon")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 20), NA)) +
  facet_wrap(facets = vars(location)) +
  labs(
    title = "Peak bloom (since Jan 1st) for selected locations in Switzerland",
    subtitle = NULL,
    x = NULL, y = NULL,
    caption = "MeteoSwiss\nhttps://opendata.swiss/en/dataset/phanologische-beobachtungen"
  ) +
  theme_light()

print(fig_meteoswiss)

ggsave(filename = "figures/fig_meteoswiss.png", plot = fig_meteoswiss, bg = "white")
```

I enhance the dataset of the five main locations with observations from the selected locations in Japan and Switzerland. I do not use data from Korea because the observations are few and the phenological event observed is first bloom and not peak bloom.

```{r}
#| label: enhance-cherry-with-selected-locations
#| warning: false
#| message: false

cherry <- bind_rows(
  kyoto,
  liestal,
  newyorkcity,
  vancouver,
  washingtondc,
  fukuoka,
  binningen,
  laufenburg,
  albisgutli,
  meteoschweiz,
  witikon
)
```

## Predictors

We have seen that a linear time trend alone is not sufficient to explain the variance in the peak bloom times. It therefore makes sense to include external data on local weather conditions in our model. 

### POWER

I use the `{nasapower}` package described at <https://docs.ropensci.org/nasapower/> to access the [NASA Prediction Of Worldwide Energy Resources (POWER) API](https://power.larc.nasa.gov/).

```{r}
#| label: get_power

get_power <- function(
    vars, 
    start = "1981-01-01", 
    end = as.character(Sys.Date()), 
    freq = "daily", 
    lat, long
  ) {
  power <- map2(
    .x  = long,
    .y = lat,
    .f = function(long = .x, lat = .y) {
      res <- nasapower::get_power(
        community = "ag",
        lonlat = c(long, lat),
        pars = vars,
        dates = c(start, end),
        temporal_api = freq
        )
      return(res)
      }
    ) |>
    list_rbind()
  return(power)
}
```

Load the following variables:

- `T2M`: Temperature at 2 Meters. The average air (dry bulb) temperature at 2 meters above the surface of the earth.
- `T2M_MAX:` Temperature at 2 Meters Maximum. The maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.
- `T2M_MIN`: Temperature at 2 Meters Minimum. The minimum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.
- `PRECTOTCORR` Corrected precipitation.
- `RH2M`: Relative Humidity at 2 Meters. The ratio of vapor pressure to the saturation vapor pressure with respect to a plane surface of pure water, expressed in percent.

```{r}
#| label: specify-vars

vars <- c("T2M", "T2M_MAX", "T2M_MIN", "PRECTOTCORR", "RH2M")
```

Retrieve the **coordinates** which we will use to match to the nearest weather stations.

```{r}
#| label: get-coordinates

crd_kyoto        <- distinct(kyoto, location, country, lat, long, alt)
crd_liestal      <- distinct(liestal, location, country, lat, long, alt)
crd_nyc          <- distinct(newyorkcity, location, country, lat, long, alt)
crd_vancouver    <- distinct(vancouver, location, country, lat, long, alt)
crd_washingtondc <- distinct(washingtondc, location, country, lat, long, alt)

crd_fukuoka   <- distinct(fukuoka, location, country, lat, long, alt)

crd_binningen    <- distinct(binningen, location, country, lat, long, alt)
crd_laufenburg   <- distinct(laufenburg, location, country, lat, long, alt)
crd_albisgutli   <- distinct(albisgutli, location, country, lat, long, alt)
crd_meteoschweiz <- distinct(meteoschweiz, location, country, lat, long, alt)
crd_witikon      <- distinct(witikon, location, country, lat, long, alt)
```

Load the predictors for the locations.

```{r}
#| label: load-nasa-power
#| eval: true

# Five main locations
power_kyoto <- get_power(vars, lat = crd_kyoto$lat, long = crd_kyoto$long) |> 
  left_join(crd_kyoto, by = c("LAT" = "lat", "LON" = "long"))

power_liestal <- get_power(vars, lat = crd_liestal$lat, long = crd_liestal$long) |> 
  left_join(crd_liestal, by = c("LAT" = "lat", "LON" = "long"))

power_nyc <- get_power(vars, lat = crd_nyc$lat, long = crd_nyc$long) |> 
  left_join(crd_nyc, by = c("LAT" = "lat", "LON" = "long"))

power_vancouver <- get_power(vars, lat = crd_vancouver$lat, long = crd_vancouver$long) |> 
  left_join(crd_vancouver, by = c("LAT" = "lat", "LON" = "long"))

power_washingtondc <- get_power(vars, lat = crd_washingtondc$lat, long = crd_washingtondc$long) |> 
  left_join(crd_washingtondc, by = c("LAT" = "lat", "LON" = "long"))

# Selected locations in Japan
power_fukuoka <- get_power(vars, lat = crd_fukuoka$lat, long = crd_fukuoka$long) |>
  left_join(crd_fukuoka, by = c("LAT" = "lat", "LON" = "long"))

# Selected locations in Switzerland
power_binningen <- get_power(vars, lat = crd_binningen$lat, long = crd_binningen$long) |>
  left_join(crd_binningen, by = c("LAT" = "lat", "LON" = "long"))

power_laufenburg <- get_power(vars, lat = crd_laufenburg$lat, long = crd_laufenburg$long) |>
  left_join(crd_laufenburg, by = c("LAT" = "lat", "LON" = "long"))

power_albisgutli <- get_power(vars, lat = crd_albisgutli$lat, long = crd_albisgutli$long) |>
  left_join(crd_albisgutli, by = c("LAT" = "lat", "LON" = "long"))

power_meteoschweiz <- get_power(vars, lat = crd_meteoschweiz$lat, long = crd_meteoschweiz$long) |>
  left_join(crd_meteoschweiz, by = c("LAT" = "lat", "LON" = "long"))

power_witikon <- get_power(vars, lat = crd_witikon$lat, long = crd_witikon$long) |>
  left_join(crd_witikon, by = c("LAT" = "lat", "LON" = "long"))
```

Combine the downloaded data sets and save them to a CSV file.

```{r}
#| label: combine-power_cherry
#| eval: true

power_cherry <- bind_rows(
  power_kyoto,
  power_liestal,
  power_nyc,
  power_vancouver,
  power_washingtondc,
  power_fukuoka,
  power_binningen,
  power_laufenburg,
  power_albisgutli,
  power_meteoschweiz,
  power_witikon
  ) |> 
  rename_with(.fn = tolower, .cols = everything()) |> 
  rename(date = yyyymmdd, long = lon, month = mm, day = dd, prcp = prectotcorr) |> 
  relocate(location, country, date, lat, long, alt, year, month, day, doy)

write_csv(power_cherry, file = "data/power_cherry.csv")
```

### Open Meteo

Use the [`{openmeteo}`](https://CRAN.R-project.org/package=openmeteo) package to retrieve weather forecasts from the [Open Meteo API](https://open-meteo.com/). We can only get weather forecasts 15 days ahead, and we are about 3 days behind with the NASA POWER data. I excluded the precipitation variable because it does not align with the corrected precipitation variable from NASA POWER. Relative humidity is re-weighted by a factor of 28/26.

```{r}
#| label: load-weather-forecasts
#| warning: false
#| message: false
#| eval: true

get_fcst <- function(loc_gc) {
  fcst <- weather_forecast(
    location = c(loc_gc$lat, loc_gc$long), 
    start = as.character(today() - days(3)), 
    end = as.character(today() + days(12)), 
    hourly = c("temperature_2m", "relative_humidity_2m"),
    daily = c("temperature_2m_max", "temperature_2m_min")
  )
  fcst$location <- tolower(loc_gc$location)
  return(fcst)
}

locations <- list(
  crd_kyoto,
  crd_liestal,
  crd_nyc,
  crd_vancouver,
  crd_washingtondc,
  crd_fukuoka,
  crd_binningen,
  crd_laufenburg,
  crd_albisgutli,
  crd_meteoschweiz,
  crd_witikon
)

locations_fcst <- map_dfr(.x = locations, .f = get_fcst) |> 
  rename(
    t2m = hourly_temperature_2m,
    t2m_max = daily_temperature_2m_max,
    t2m_min = daily_temperature_2m_min,
    rh2m = hourly_relative_humidity_2m
  ) |> 
  mutate(rh2m = rh2m * 28/26) |> 
  group_by(location, date) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  relocate(location, date, t2m, t2m_max, t2m_min) |> 
  mutate(across(where(is.numeric), ~ round(.x, 2))) |> 
  mutate(prcp = NA)

write_csv(locations_fcst, file = "data/locations_fcst.csv")
```

Read in the weather forecasts from Open Meteo and combine them with the historical climate indicators from NASA POWER.

```{r}
#| label: combine-power-openmeteo
#| output: false
#| warning: false
#| message: false

power_cherry <- read_csv("data/power_cherry.csv")

locations_fcst <- read_csv("data/locations_fcst.csv")

last_obs_date <- power_cherry |> 
  filter(!is.na(t2m)) |> 
  slice_max(order_by = date) |> 
  select(date) |> 
  pull() |> 
  unique()

power_cherry_fcst <- bind_rows(
  power_cherry,
  locations_fcst |> 
    filter(date > last_obs_date)
  ) |> 
  filter(!is.na(t2m))


power_cherry_fcst <- power_cherry_fcst |> 
  mutate(year = year(date), month = month(date), day = day(date), doy = yday(date)) |> 
  group_by(location) |> 
  fill(country, lat, long, alt, .direction = "down") |> 
  ungroup()

write_csv(power_cherry_fcst, file = "data/power_cherry_fcst.csv")
```

## Feature engineering

### Midrange temperature

Create the midrange temperature, the average of the maximum and the minimum average temperature measured on a specific day.

```{r}
#| label: create-t2m_range

power_cherry_fcst <- power_cherry_fcst |> 
  mutate(t2m_range = (t2m_max + t2m_min) / 2) |> 
  relocate(location, date, t2m, t2m_max, t2m_min, t2m_range)
```

### Aggregated precipitation since January 1st

```{r}
#| label: create-aprcp

power_cherry_fcst <- power_cherry_fcst |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(
    aprcp = cumsum(prcp)
  ) |> 
  ungroup()
```

### Aggregated growing degree days since January 1st

Create the accumulated growing degree days (AGDD) as the sum of growing degree days (GDD) since January 1st, where GDD are calculated as the number of degrees C by which the midrange temperature exceeds a base temperature, which can be 0 degrees C, or a species-specific base temperature, such as 4 degrees C for cherry trees. If the midrange temperature does not exceed the base temperature on a given day, no growing degree days are counted.

```{r}
#| label: create-agdd

base_temp <- 4.0

power_cherry_fcst <- power_cherry_fcst |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(
    gdd = pmax(t2m_range - base_temp, 0),
    agdd = cumsum(gdd)
  ) |> 
  ungroup()
```

## Select time range

I select only days between January 1st of each year until the day of year of the last available forecast date of the current year. This ensures that the model does not rely on data which is not available on the day of the submission deadline.

```{r}
#| label: select-time-range
#| warning: false
#| message: false

last_fcst_date <- power_cherry_fcst |> 
  filter(!is.na(t2m_max)) |> 
  select(date) |>
  pull() |> 
  max()

power_cherry_fcst <- power_cherry_fcst |> 
  relocate(location, country, date, year, month, day, doy, lat, long, alt) |> 
  filter(doy <= yday(last_fcst_date))
```

## Temporal aggregation

I have experimented with seasonal, monthly, weekly and daily aggregates of the predictors. I choose to use monthly aggregates because it makes for easily interpretable variables that are also used by other professional forecasters such as the Japanese Meterorological Agency (JMA), the Washington Post Weather Gang and the National Park Service of the United States.

```{r}
#| label: aggregate-to-months

df_months <- power_cherry_fcst |> 
  select(-c(day, doy)) |>
  mutate(month = tolower(month(date, label = TRUE, abbr = TRUE))) |> 
  group_by(country, location, year, month) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)), .groups = "drop") |> 
  ungroup() |> 
  pivot_longer(cols = -c(country, location, year, month, lat, long, alt), names_to = "variable", values_to = "value") |> 
  mutate(variable = paste0(variable, "_", month)) |> 
  select(-c(month)) |> 
  pivot_wider(names_from = variable, values_from = value) |> 
  mutate(
    prcp_feb = if_else(is.na(prcp_feb), mean(prcp_feb, na.rm = TRUE), prcp_feb),
    aprcp_feb = if_else(is.na(aprcp_feb), mean(aprcp_feb, na.rm = TRUE), aprcp_feb),
    prcp_mar = if_else(is.na(prcp_mar), mean(prcp_mar, na.rm = TRUE), prcp_mar),
    aprcp_mar = if_else(is.na(aprcp_mar), mean(aprcp_mar, na.rm = TRUE), aprcp_mar)
  )

cherry_months <- cherry |> 
  full_join(y = df_months, join_by(country, location, year, lat, long, alt)) |> 
  relocate(bloom_date, bloom_doy, country, location, year, lat, long, alt) |>
  group_by(country, location, lat, long, alt) |> 
  arrange(bloom_date) |> 
  fill(country, location, lat, long, alt, .direction = "down") |> 
  ungroup() |> 
  filter(year >= 2000) |> 
  mutate(year2 = year^2, year3 = year^3) |> 
  select(-c(bloom_date, lat, long))

cherry_months_2024 <- cherry_months |> 
  filter(!is.na(bloom_doy))

cherry_months_2025 <- cherry_months |> 
  filter(year == 2025)

cherry_months <- bind_rows(
  cherry_months_2024,
  cherry_months_2025
)

write_csv(cherry_months, file = "data/cherry_months.csv")
```

```{r}
#| label: plot-polynomial-time-trend
#| warning: false
#| message: false

cherry_months |> 
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  ggplot(aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 3)) +
  geom_point() +
  geom_step() +
  facet_grid(cols = vars(location))
```


## Exploratory data analysis (EDA)

### Linear model

A simple linear model can help us understand how predictive a feature is. We plot the average maximum temperature in February.

```{r}
#| label: fig-t2m_max_feb
#| fig-cap: "Average maximum winter temperature in February"

cherry_months |>
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  ggplot(aes(x = year, y = t2m_max_feb)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum temperature in February (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1980, 2025)) +
  theme_light()
```

We fit a linear model with different intercepts and time trends for each location and a weather effect. We can add interaction effects to the temperature data too if we want.

```{r}
#| label: fit-lin-t2m_max_feb

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:t2m_max_feb, 
    data = cherry_months
  )

mean_pred <- predict(object = lm_fit, new_data = cherry_months)
```

Plot the predictions and the actual data.

```{r}
#| label: fig-predictions-t2m_max_feb
#| fig-cap: "Predictions from linear time trend and average maximum temperature in February"
#| warning: false

cherry_months |> 
  bind_cols(mean_pred) |> 
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

Because `t2m_max_feb` shows a clear upwards trend over the past twenty years, we are able model the trend towards earlier peak bloom dates relatively well, except for Kyoto, where the climate is mild throughout the year.

It is clear that `t2m_max_mar` would be an even better predictor.

### Graphical

Plot a linear time trend

```{r}
#| warning: false

cherry_months |>
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  ggplot(aes(x = year, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

Plot peak dates vs aggregated growing degree days in February

```{r}
#| warning: false

cherry_months |>
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |> 
  ggplot(aes(x = agdd_feb, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```


### Correlations

We use the `{corr}` package to check the correlations of the predictors.

```{r}
#| label: correlations
#| warning: false
#| message: false

cherry_months |>
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc")) |>
  correlate() |>
  gt() |>
  fmt_number() |>
  tab_header(title = "correlations") |>
  tab_footnote(footnote = "") |> 
  as_raw_html()
```

### Decision tree

Fit a decision tree for exploratory data analysis. We use the `{parsnip}` package to create a model that runs on the `"rpart"` engine, and we use the `rpart.plot` package to plot the decision tree.

```{r}
#| label: decision-tree
#| warning: false
#| message: false

set.seed(123)

cart_specs <- decision_tree(
  tree_depth = 10,
  cost_complexity = 0.001
  ) |>
  set_mode("regression") |>
  set_engine("rpart")

cart_fit <- cart_specs |>
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_months |>
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc"))
  )

cart_fit$fit |> 
  rpart.plot()
```

Aggregated growing degree days (AGDD) are the most important predictor.

### Variable importance plot (VIP)

Use the `{vip}` package to create a variable importance plot (VIP)

```{r}
cart_fit |> 
  vip() +
  theme_bw()
```

## Modeling

### Data spending

Split the data into a training set and a test set and split the training set into folds for cross-validation.

```{r}
#| label: initial-split
#| eval: true

set.seed(123)

cherry_data <- cherry_months |> 
  filter(year != 2025)

cherry_split <- initial_split(data = cherry_data, prop = 2/3)

cherry_train <- training(cherry_split)
cherry_test  <- testing(cherry_split)

cherry_recipe <- recipe(formula = bloom_doy ~ ., data = cherry_train) |>
  step_dummy(all_nominal_predictors()) |> 
  step_harmonic(year, frequency = 1 / 11, cycle_size = 1) |> 
  step_zv(all_nominal_predictors()) |> 
  step_normalize(all_numeric_predictors())

folds <- vfold_cv(cherry_train, v = 10)
```

Specify the models.

```{r}
#| label: model-specs

ridge_tune_mod <- linear_reg(
  mixture = 0,
  penalty = tune()
  ) |>
  set_mode("regression") |>
  set_engine("glmnet")

lasso_tune_mod <- linear_reg(
  mixture = 1,
  penalty = tune()
  ) |>
  set_mode("regression") |>
  set_engine("glmnet")

rf_tune_mod <- rand_forest(
  trees = tune(),
  mtry = tune(),
  min_n = tune()
  ) |> 
  set_engine("ranger", num.threads = 6, importance = "impurity") |> 
  set_mode("regression")

xgb_tune_mod <- boost_tree(
  mtry = tune(),
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
  ) |>
  set_engine("xgboost", num.threads = 6) |> 
  set_mode("regression")

lgb_tune_mod <- boost_tree(
  mtry = tune(),
  trees = tune(),
  tree_depth = tune(),
  learn_rate = tune(),
  loss_reduction = tune()
  ) |> 
  set_engine("lightgbm", num.threads = 6) |> 
  set_mode("regression")

cub_tune_mod <- cubist_rules(
  committees = tune(),
  neighbors = tune(),
  max_rules = tune()
  ) |>
  set_engine("Cubist") |>
  set_mode("regression")
```

Create the search grids for the hyperparameter space.

```{r}
#| label: parameter-grids

ridge_grid <- grid_random(
  penalty(range = c(-10, 10)),
  size = 50
)

lasso_grid <- grid_random(
  penalty(range = c(-10, 10)),
  size = 50
)

rf_grid <- grid_random(
  trees(range = c(1000, 2000)),
  mtry(range = c(1, 10)),
  min_n(range = c(1, 10)),
  size = 30
)

xgb_grid <- grid_random(
  trees(range = c(1000, 2000)),
  tree_depth(range = c(1, 5)),
  learn_rate(range = c(0.1, 0.005)),
  mtry(range = c(1, 10)),
  loss_reduction(range = c(0, 0.1)),
  size = 30
)

lgb_grid <- grid_random(
  trees(range = c(1000, 2000)),
  tree_depth(range = c(1, 5)),
  learn_rate(range = c(0.1, 0.005)),
  mtry(range = c(3, 10)),
  loss_reduction(range = c(0, 0.1)),
  size = 30
)

cub_grid <- grid_random(
  committees(range = c(30, 45)),
  neighbors(range = c(1, 10)),
  max_rules(range = c(5, 30)),
  size = 30
)
```

Tune the models using the on the grids.

```{r}
#| label: tune-models
#| output: false
#| warning: false
#| message: false

ridge_tune_res <- tune_grid(
  object = workflow() |> 
    add_recipe(cherry_recipe) |> 
    add_model(ridge_tune_mod),
  resamples = folds, 
  grid = ridge_grid,
  control = control_race(),
  metrics = metric_set(mae)
)

lasso_tune_res <- tune_grid(
  object = workflow() |> 
    add_recipe(cherry_recipe) |> 
    add_model(lasso_tune_mod),
  resamples = folds, 
  grid = lasso_grid,
  control = control_race(),
  metrics = metric_set(mae)
)

rf_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(rf_tune_mod),
    resamples = folds,
    grid = rf_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

xgb_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(xgb_tune_mod),
    resamples = folds,
    grid = xgb_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

lgb_tune_res <- tune_race_anova(
  object = workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lgb_tune_mod),
    resamples = folds,
    grid = lgb_grid,
    control = control_race(),
    metrics = metric_set(mae)
  )

cub_tune_res <- tune_race_anova(
  object = workflow() |>
    add_recipe(cherry_recipe) |>
    add_model(cub_tune_mod),
  resamples = folds,
  grid = cub_grid,
  control = control_race(),
  metrics = metric_set(mae)
)
```

Use `finalize_workflow()` to fit the models with the best hyperparameter combination as assessed using k-fold cross validation to the whole training set.

```{r}
#| label: finalize-workflows
#| output: false
#| warning: false
#| message: false

final_ridge_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(ridge_tune_mod), 
      select_best(ridge_tune_res)
      ), 
  data = cherry_train
  )

final_lasso_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lasso_tune_mod), 
      select_best(lasso_tune_res)
      ), 
  data = cherry_train
  )

final_rf_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(rf_tune_mod), 
      select_best(rf_tune_res)
      ), 
  data = cherry_train
  )

final_xgb_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(xgb_tune_mod), 
      select_best(xgb_tune_res)
      ), 
  data = cherry_train
  )

final_lgb_fit <- fit(
  object = finalize_workflow(
      workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(lgb_tune_mod), 
      select_best(lgb_tune_res)
      ), 
  data = cherry_train
  )

final_cub_fit <- fit(
  object = finalize_workflow(
    workflow() |>
      add_recipe(cherry_recipe) |>
      add_model(cub_tune_mod),
    select_best(cub_tune_res)
    ),
  data = cherry_train
  )

final_ridge_res <- predict(final_ridge_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)

final_lasso_res <- predict(final_lasso_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)

final_rf_res <- predict(final_rf_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)

final_xgb_res <- predict(final_xgb_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)

final_lgb_res <- predict(final_lgb_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)

final_cub_res <- predict(final_cub_fit, new_data = cherry_test) |> 
  bind_cols(cherry_test)
```

Compute the MAE for the "final" model with the `{yardstick}` functions `mae()`.

```{r}
#| label: compute-final-mae

final_ridge_mae <- mae(final_ridge_res, truth = bloom_doy, estimate = .pred)

final_lasso_mae <- mae(final_lasso_res, truth = bloom_doy, estimate = .pred)

final_rf_mae <- mae(final_rf_res, truth = bloom_doy, estimate = .pred)

final_xgb_mae <- mae(final_xgb_res, truth = bloom_doy, estimate = .pred)

final_lgb_mae <- mae(final_lgb_res, truth = bloom_doy, estimate = .pred)

final_cub_mae <- mae(final_cub_res, truth = bloom_doy, estimate = .pred)

tribble(
  ~ model, ~ mae,
  "ridge", final_ridge_mae$.estimate,
  "lasso", final_lasso_mae$.estimate,
  "rf", final_rf_mae$.estimate,
  "xgb", final_xgb_mae$.estimate,
  "lgb", final_lgb_mae$.estimate,
  "cub", final_cub_mae$.estimate
  ) |> 
  arrange(mae)
```


```{r}
vip(final_cub_fit)
```

The Cubist model was the best and it gave the altitude variable a high weight. I want to use the Cubist model to make the predictions for my submission

### Prediction

I need point forecasts and prediction intervals

```{r}
gc()

cub_pred <- predict(object = final_cub_fit, new_data = cherry_months_2025)

# Bootstrap
set.seed(123)
boot_resamples <- bootstraps(cherry_months_2024, times = 200)

# Fit the models on each resample
boot_preds <- boot_resamples %>%
  mutate(
    model = map(splits, ~ fit(object = final_cub_fit, data = analysis(.x))),
    pred = map(model, ~ predict(.x, new_data = cherry_months_2025)$.pred)
  )

# Compute 90% prediction interval
predictions <- boot_preds |> 
  unnest(pred) |> 
  bind_cols(tibble(location = rep(cherry_months_2025 |> select(location) |> pull(), times = 200))) |> 
  group_by(location) |> 
  summarize(
    bloom_doy_pred = mean(pred),
    lower_90 = quantile(pred, 0.05),
    upper_90 = quantile(pred, 0.95)
  ) |> 
  filter(location %in% c("kyoto", "liestal", "newyorkcity", "vancouver", "washingtondc"))
```


```{r}
#| label: save-predicitons
predictions |> 
  rename(prediction = bloom_doy_pred, lower = lower_90, upper = upper_90) |> 
  write_csv("cherry-predictions.csv")
```


```{r}
#| label: tbl-predictions
#| tbl-cap: "Peak bloom predictions"

doy_to_date <- function (year, doy) {
  strptime(paste(year, doy, sep = '-'), '%Y-%j') |> # create date object
    strftime('%Y-%m-%d') # translate back to date string in ISO 8601 format
}

predictions |> 
  rename(prediction = bloom_doy_pred, lower = lower_90, upper = upper_90) |> 
  mutate(
    year = year(Sys.Date()),
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  ) |> 
  gt() |> 
  as_raw_html()
```


