---
title: "Peak Bloom Prediction"
author: "econmaett"
lang: en-US
format:
  html:
    embed-resources: true
fig-cap-location: bottom
tbl-cap-location: bottom
toc: true
editor: source
---

```{r}
#| label: setup
#| include: false

knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, error = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "80%"
)
```

See more at <https://en.wikipedia.org/wiki/Phenology>.

## Packages

We start by loading all the packages necessary to run this analysis

```{r}
#| label: load-pkgs
#| output: false

invisible(Sys.setlocale("LC_ALL","English"))

library(tidymodels)
library(corrr)
library(rpart.plot)
library(vip)
library(tidyverse)
library(readr)
library(gt)
library(kableExtra)
library(rvest)
library(httr2)
library(jsonlite)
library(nasapower)
library(openmeteo)
library(sf)
library(mapview)

tidymodels_prefer()
```

## Data

We begin by loading the data necessary for this analysis

### Clean data

The data for the five sites is provided as a simple text file in CSV format. Each file contains the dates of the peak bloom of the cherry trees at the respective site, alongside the geographical location of the site. The six columns in each data file are:
- `location`: a human-readable location identifier (`string`).
- `lat`: (approximate) latitude of the cherry trees (`double`).
- `long`: (approximate) longitude of the cherry trees (`double`).
- `alt`: (approximate) altitude of the cherry trees (`double`).
- `year`: year of the observation (`integer`).
- `bloom_date`: date of peak bloom of the cherry trees ([ISO 8601](https://xkcd.com/1179/) date `string`). The "peak bloom date" may be defined differently for different sites
- `bloom_doy`: days since January 1st until peak bloom (`integer`). January 1st is `1`.

#### Kyoto

The file `kyoto.csv` has been obtained from Yasuyuki AONO (`aono(at)envi.osakafu-u.ac.jp`) at <http://atmenv.envi.osakafu-u.ac.jp/aono/kyophenotemp4/>.

The geographical location at `35.01198°N` (latitude), `135.6761°E` (longitude), `44` meters above sea levels (altitude) roughly corresponds to the **Nakanoshima** area of the **Arashiyama Park** in Kyoto, Japan.

The peak bloom date of the **Prunus jamasakura** is determined by a local news paper in Arashiyama (Kyoto, Japan). Data prior to 1888 is extracted from various descriptions or estimated:
- Data from the 9th to the 14th centuries was acquired and analyzed by Aono and Saito (2010; International Journal of Biometeorology, 54, 211-219).
- Phenology for 15th to 21st centuries was acquired and analyzed by Aono and Kazui (2008; International Journal of Climatology, 28, 905-914).

Note that the data set for Kyoto includes implicit missing values (`NA`s). [Chapter 18 "Missing Values"](https://r4ds.hadley.nz/missing-values) of Hadley Wickham's book "R for Data Science (2e)" describes strategies for dealing with implicit `NA`s.

Let's load the data for Kyoto, Japan

```{r}
#| label: load-kyoto

kyoto <- read_csv("data/kyoto.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  fill(location, lat, long, alt, .direction = "downup")
```

Let's plot the data for Kyoto, Japan

```{r}
#| label: fig-kyoto
#| fig-cap: "Peak cherry blossom days in Kyoto, Japan"

fig_kyoto <- kyoto |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 0.5) +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(800, 2100, by = 100), NA)) +
  labs(title = "Peak cherry blossom days in Kyoto, Japan", x = NULL, y = NULL) +
  theme_light()

print(fig_kyoto)

ggsave(filename = "figures/fig_kyoto.png", plot = fig_kyoto, bg = "white")
```

Here we have used **locally estimated scatterplot smoothing (LOESS)** or **locally weighted scatterplot smoothing (LOWESS)** to who how the average peak bloom day changed over the years. 

We could explicitly model this with a **local level (+ trend)** state-space model that we fit with the **Kalman filter and Smoother**, which also allows us to effectively deal with **missing observations**.

MeteoSwiss suggests using a **Gaussian Kernel** with a *20-year window* to extract the trend, which is equivalent to using a **low-pass filter** that shuts down all higher frequencies.

It also looks like the observations exhibit **heteroskedasticity**, that is, the variance changes over time, and it looks like the variance has become more narrow over time.

We can use **changepoint detection** methods to test for the changing time trend and the changing variance.

Assuming that the data in the past is correct, there do seem to be seasonal cycles, perhaps connected to solar activity or the tilt of the earth.

There is a clear trend towards earlier blooming, perhaps connected to anthropogenic climate change and urbanization. Studies attribute about half of the earlier blooming times in recent years to the **urban heat island effect**, and the other half to **climate change**.

Modelling the local level and the trend would be useful if we are to forecast a couple years into the future, since an individual prediction may be incorrect, but on average, they may be good. 

However, since we are only interested in forecasting a single moment in time, which may be greatly influenced by local weather, we probably are better off using historical weather data that we augment with forecasts, either by professional forecasters or by extrapolating the previous predictors.

#### Liestal

The file `liestal.csv` contains observations for **a single wild cherry (Prunus avium) tree** in **Liestal-Weideli**, at `7.4814°N` (latitude), `7.730519°E` (longitude) and `350` meters above sea levels (altitude). 

The **peak bloom date** is defined as the day when **25%** of the blossoms are in full bloom. The date is determined by [MeteoSwiss](https://www.meteoswiss.admin.ch/climate/climate-change/vegetation-development/long-term-series-of-phenological-observations.html).

This means that 70% of buds would be open a couple days later. This raises the question of whether we should model the locations separately.

The data is provided by **Landwirtschaftliches Zentrum Ebenrain, Sissach and MeteoSwiss**.

```{r}
#| label: load-liestal

liestal <- read_csv("data/liestal.csv")
```

Let's plot the data for Liestal, Switzerland:

```{r}
#| label: fig-liestal
#| fig-cap: "Peak cherry blossom days for Liestal, Switzerland"

fig_liestal <- liestal |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#ff0080", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(
    title = "Peak cherry blossom days for Liestal, Switzerland", 
    x = NULL, y = NULL
  ) +
  theme_light()

print(fig_liestal)

ggsave(filename = "figures/fig_liestal.png", plot = fig_liestal, bg = "white")
```

For Liestal you can see the impact of the infamous heat summer of 2003 which I remember from my childhood, although interestingly, there's another record in 1990.

#### New York

The file `newyorkcity.csv` contains **only a single observation** for the year 2024. It concerns the **Yoshino Cherry (Prunus x yedoensis)** tree in **Washington Square Park**, New York City, at `40.7304 °N` (latitude), `-73.99809°E` (longitude) and `8.5` meters above sea level (altitude).

Let's load the data for New York City

```{r}
#| label: load-newyorkcity

newyorkcity <- read_csv("data/nyc.csv")
```

Luckily, there are some historical observations dating *back to 2019* in the data provided by **USA-NPN**.

The Washington Square Park has site id `32789` and the **Yoshino** cherry you should predict has species id `228`.

We have been provided with USA-NPN data on individual cherry trees by the competition organizers, and we can load it:

```{r}
#| label: load-historical-data-nyc

nyc_data_npn <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  filter(Site_ID == 32789, Species_ID == 228) |> 
  mutate(Observation_Date = mdy(Observation_Date))
```

This file now contains individual observations of the phenophase and not the actual peak bloom date.

We use the variable `Intensity_Value` which is either `-9999`, `95% or more`, `Less than 5%`, `25-49%`, `75-94%`, `50-74%`.

Since according to the variable `Individual_ID`, only one tree, `183140` is being observed, we can use the first date where `Intensity_Value` is `50-74%` as our `bloom_doy`.

```{r}
#| warning: true

nyc_data <- nyc_data_npn |>
  arrange(Observation_Date) |>
  mutate(year = year(Observation_Date)) |>
  group_by(year) |>
  summarize(
    first_flower_index = min(which((Intensity_Value %in% c("75-94%", "95% or more")))),
    bloom_date = as_date(strftime(Observation_Date[first_flower_index], format = "%Y-%m-%d")),
    bloom_doy = Day_of_Year[first_flower_index],
    .groups = "drop"
  ) |>
  filter(!is.na(bloom_doy)) |>
  select(-first_flower_index) |>
  mutate(location = "newyorkcity") |> 
  arrange(year)
```

Then we combine the data sets

```{r}
#| label: augment-nyc-data

newyorkcity <- newyorkcity |>
  # Don't add a year twice!
  bind_rows(nyc_data |> filter(year != 2024)) |>  
  fill(lat, long, alt)
```

We now have a total of four observations for New York City

```{r}
#| label: tbl-nyc-data
#| tbl-cap: "Augmented NYC observations"

newyorkcity |> 
  filter(location == "newyorkcity") |> 
  kable() |> 
  kable_styling()
```

#### Vancouver

Let's load the data for Vancouver, BC

```{r}
#| label: load-vancouver

vancouver <- read_csv("data/vancouver.csv")
```

The file `vancouver.csv` contains observations for cherry trees of the cultivar  **Akebono** in **Maple Grove Park**, Vancouver, BC located approximately at `49.2236916°N` (latitude), `-123.1636251°E` (longitude), `24` meters above sea levels (altitude).

Unfortunately, there are **only three observations** for 2022, 2023, and 2024, respectively. But casual observations have been recorded in the form of **photos posted** to the [VCBF Neighbourhood Blog for Kerrisdale](https://forums.botanicalgarden.ubc.ca/threads/kerrisdale.36008/). You can **search the forum** for the keywords `"Akebono"` (i.e., the name of the cultivar) and `"Maple Grove Park"` (i.e., the location of the trees).

After searching through the web forum, I have decided to add the following observations to the data set:

```{r}
#| label: define-vancouver_forum

vancouver_forum <- tribble(
  ~ year, ~ bloom_date,
  2021, "2021-03-31", # Start: 2021-03-21, Full: 2021-03-31, Full: 2021-04-07
  2020, "2020-04-03", # Full bloom
  2018, "2018-03-18", # Full bloom
  2017, "2017-04-11", # Full bloom 
  2015, "2015-03-19", # Full bloom
  2011, "2011-04-01", # Full bloom was over on 2011-04-15
  2010, "2010-03-14", # 60% bloom on 2010-03-07
  2008, "2008-04-14" # Start: 2008-04-06, Full: 2008-04-14, finished: 2008-04-22
  ) |> 
  mutate(
    bloom_date = as_date(bloom_date),
    bloom_doy = as.numeric(1 + as_date(bloom_date) - as_date(paste0(year, "-01-01"))),
    location = "vancouver"
  )
```

We can do some additional plausibility checks. For example, spring 2010 was indeed unusually warm, [causing cherry blossoms to bloom during the Vancouver 2010 Winter Olympics](https://www.upi.com/News_Photos/view/upi/33393f9d20159ecb347de9aa1d670922/2010-Winter-Olympics-in-Vancouver/). National Geographic claimed it was the [warmest Olympics to date](https://www.nationalgeographic.com/adventure/article/100212-vancouver-2010-warmest-winter-olympics).

We add these observations to the data set

```{r}
#| label: augment-vancouver

vancouver <- vancouver |> 
  bind_rows(vancouver_forum) |> 
  fill(lat, long, alt) |> 
  arrange(year)
```

Now we have a total of eleven observations for Vancouver, BC.

```{r}
#| label: tbl-augmented-vancouver-data
#| tbl-cap: "Augmented Vancouver data"

vancouver |> 
  filter(location == "vancouver") |> 
  kable() |> 
  kable_styling()
```

#### Washington

The file `washingtondc.csv` has been obtained from the EPA's Climate Change Indicators in the United States <https://www.epa.gov/climate-indicators/cherry-blossoms>.

The geographical location at `38.88535°N` (latitude), `-77.03863°E` (longitude) and `0` meters above sea level (altitude) correspond to the location of the [Tidal Basin in Washington, DC](https://www.nps.gov/articles/dctidalbasin.htm).

The peak bloom date is defined as the day when **70%** of the **Yoshino Cherry (Prunus x yedoensis)** are in full bloom, as determined by the [National Park Service](https://www.nps.gov/subjects/cherryblossom/bloom-watch.htm).

Let's load the data for Washington DC

```{r}
#| label: load-washingtondc

washingtondc <- read_csv("data/washingtondc.csv")
```

Let's plot the data for Washington D.C

```{r}
#| label: fig-washingtondc
#| fig-cap: "Peak cherry blossom days for Washington DC, USA"

fig_washingtondc <- washingtondc |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#ff0080", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1920, 2020, by = 20), NA)) +
  labs(title = "Peak cherry blossom days for Washington DC, USA", x = NULL, y = NULL) +
  theme_light()

print(fig_washingtondc)

ggsave(filename = "figures/fig_washingtondc.png", plot = fig_washingtondc, bg = "white")
```

Now that we have loaded the datasets for the five locations, we can combine them into a single CSV file called `cherry.csv`.

```{r}
#| label: combine-cherry-csv
#| eval: false

cherry <- bind_rows(kyoto, liestal, newyorkcity, vancouver, washingtondc)

write_csv(cherry, file = "data/cherry.csv")
```

Now we can load the `cherry.csv` file from our `data/` folder

```{r}
#| label: load-cherry

cherry <- read_csv("data/cherry.csv")
```

We can show the last three observation for each location

```{r}
#| label: tbl-01
#| tbl-cap: "latest 3 observations for each location"

cherry |> 
  group_by(location) |> 
  slice_tail(n = 3) |> 
  kable() |> 
  kable_styling()
```

We can visualize the time series for our five main locations next to each other. Notice the use of the `{stringr}` function `str_to_title()` to convert the first letter of the locations to uppercase.

```{r}
#| label: fig-cherries_1880
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Time series of peak bloom of cherry trees since 1880 at five different sites.

df_cherry_1880 <- cherry |> 
  filter(year >= 1880)

fig_cherries <- ggplot(data = df_cherry_1880, aes(year, bloom_doy)) +
  geom_smooth(method = "loess", formula = y ~ x, colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab") +
  geom_point(color = "#ff0080", alpha = 0.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 40), NA)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(
    title = "Peak bloom (since Jan 1st)",
    x = NULL, 
    y = NULL
  ) +
  theme_light()

print(fig_cherries)

ggsave(filename = "figures/fig_cherries.png", plot = fig_cherries, bg = "white")
```

Create summary statistics to get an overview over the data. Specifically, we are interested in the range of the `year` variable, the total number of observations at each location, and the mean, median, and standard deviation of the `bloom_doy` variable.

```{r}
#| label: tbl-summary-statistics
#| tbl-cap: "Summary statistics for the main data set"

cherry |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location) |> 
  summarise(
    min_year = min(year),
    max_year = max(year),
    n_obs = n(),
    mean = round(mean(bloom_doy)),
    median = round(median(bloom_doy)),
    sd = round(sd(bloom_doy))
  ) |> 
  kable() |> 
  kable_styling()
```

We see that the standard deviation for Liestal is much larger than the rest, where peak bloom dates are on average about a weak apart from their historical average.

We check for missing observations in the most recent years:

```{r}
cherry |> 
  filter(location == "kyoto") |> 
  filter(is.na(bloom_doy)) |> 
  tail() |> 
  kable() |> 
  kable_styling()
```

Kyoto has multiple implicit `NA`s, including in 1730, 1872, 1895, 1919, 1921, and 1945. So if we use data since 1880, as the demo analysis suggests, we still have four `NA`s. For time series analysis, we usually assume equally-spaced observations, and therefore need to either cut the data to start in 1946, or to replace the `NA`s with a sensible value. The other four locations do not contain any implicit `NA`.

Let's plot the distribution of the peak bloom days for the locations:

```{r}
#| label: fig-histogram
#| fig-cap: "Histogram of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location)) +
  geom_histogram() +
  labs(title = "Histogram of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's use a density plot instead:

```{r}
#| label: fig-density-plot
#| fig-cap: "Denisty plots of peak bloom (since Jan 1st)"

cherry |> 
  filter(year >= 1950) |> 
  filter(!(location %in% c("newyorkcity", "vancouver"))) |> 
  ggplot(mapping = aes(x = bloom_doy, fill = location, colour = location)) +
  geom_density(alpha = 0.5) +
  labs(title = "Denisty plots of peak bloom (since Jan 1st)", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())
```

Let's plot all five time series in the same plot.

```{r}
#| label: fig-cherry-blossoms-combined
#| fig-cap: "Peak cherry blossom days for different locations"

fig_combined <- cherry |> 
  filter(year >= 1880) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy, colour = location)) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE) +
  geom_step(linetype = "dotted", lwd = 0.5) +
  geom_point(alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1880, 2020, by = 20), NA)) +
  labs(title = "Peak cherry blossom days by locations", x = NULL, y = NULL) +
  theme_light() +
  theme(legend.position = "bottom", legend.title = element_blank())

print(fig_combined)

ggsave(filename = "figures/fig_combined.png", plot = fig_combined, bg = "white")
```

This graphic highlights the negative linear time trend for all locations.

It also highlights that a linear time trend will likely underestimate the accelerating trend towards earlier peak bloom dates.

A simple method such as **ETS**, which places a lower weight on observations far in the past, may be better at modelling the recent trend.

In any case, we should compare all our models against **benchmark models** such as the **naive model**, where the prediction is equal to the last observation, and a simple **AR(1)** model.

### Additional data

The competition organizers provide additional data sets for sites other than the main sites relevant for the competition. 

We can use these time series in your modeling to help with spatial and temporal extrapolation. 

For example, you can add predictors to these data sets and train models on additional sites.

#### Japan

The data file `japan.csv` is provided by the [Japanese Meteorological Agency (JMA)](https://www.data.jma.go.jp/sakura/data/pdf/005.pdf) and contains recorded peak bloom dates for cherry trees at various sites across Japan. The sample trees are located within a 5km radius of the location indicated in the data file. 

We can load the data set, make implicit missing values explicit and change the `location` variable to only include the city names in lowercase letters.

```{r}
#| label: read-japan

japan <- read_csv("data/japan.csv") |>  
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Japan/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `japan.csv` file contains observations between `r min(japan$year)` and `r max(japan$year)` for `r length(unique(japan$location))` locations in Japan.

Let's plot peak bloom dates for Kyoto, Japan

```{r}
#| label: fig-bloom-kyoto2
#| fig-cap: "Peak bloom dates in Kyoto, Japan"

fig_kyoto2 <- japan |>
  filter(location == "kyoto") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 10), NA)) +
  labs(title = "Peak bloom (since Jan 1st) in Kyoto, Japan", x = NULL, y = NULL) +
  theme_light()

print(fig_kyoto2)

ggsave(filename = "figures/fig_kyoto2.png", plot = fig_kyoto2, bg = "white")
```

#### Korea

The file `south_korea.csv` contains **first flowering dates** of cherry trees at various sites across South Korea, curated by the **Korean Meteorological Administration (KMA)**.

Because these are **first flowering dates**, they are likely to occur earlier than hypothetical full bloom dates would be.

Trees in South Korea were assumed to be of the **Prunus Yedoensis** variety based on previous work (Hur, Ahn, and Shim 2014; Chung et al. 2009) and its frequent occurrence.

Again, we make implicit missing values explicit, and I change the location names to lowercase so that to match the style of the main data set.

```{r}
#| label: load-korea

korea <- read_csv("data/south_korea.csv") |> 
  complete(year =  full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "South Korea/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `south_korea.csv` file contains observations between `r min(korea$year)` and `r max(korea$year)` at `r length(unique(korea$location))` locations in South Korea.

Let's plot the first bloom dates in Seoul, South Korea:

```{r}
#| label: fig-seoul
#| fig-cap: "First bloom in Seoul, South Korea"

fig_seoul <- korea |>
  filter(location == "seoul") |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = FALSE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1980, 2020, by = 5), NA)) +
  labs(title = "First bloom (since Jan 1st) in Seoul, Korea", x = NULL, y = NULL) +
  theme_light()

print(fig_seoul)

ggsave(filename = "figures/fig_seoul.png", plot = fig_seoul, bg = "white")
```

#### MeteoSwiss

The data file `meteoswiss.csv` contains peak bloom dates for various sites across Switzerland, obtained from <https://opendata.swiss/en/dataset/phanologische-beobachtungen>.

Note that when you use the `{utils}` function `read.csv()` you need to specify the option `encoding = "UTF-8"` to allow for special characters that are common in the German language.

```{r}
#| label: load-meteoswiss

meteoswiss <- read_csv("data/meteoswiss.csv") |> 
  complete(year = full_seq(year, 1), location) |> 
  mutate(location = str_remove(location, "Switzerland/")) |> 
  mutate(location = tolower(location)) |> 
  fill(location, lat, long, alt, .direction = "downup") |> 
  mutate(location = factor(location))
```

The `meteoswiss.csv` file contains observations between `r min(meteoswiss$year)` and `r max(meteoswiss$year)` at `r length(unique(meteoswiss$location))` locations.

Plot the data for Zurich, Switzerland

```{r}
#| label: fig-meteoswiss
#| fig-cap: "Peak bloom (since Jan 1st) in Zurich, Switzerland"

fig_meteoswiss <- meteoswiss |> 
  filter(str_detect(location, "zürich-")) |>
  mutate(location = str_remove(location, "zürich-")) |> 
  ggplot(mapping = aes(x = year, y = bloom_doy)) +
  geom_smooth(method = "loess", colour = "#644128", se = TRUE) +
  geom_step(linetype = "dotted", colour = "#ffa5ab", lwd = 1) +
  geom_point(color = "#f72585", alpha = 0.5, size = 2.5) +
  scale_x_continuous(breaks = c(seq(1950, 2020, by = 20), NA)) +
  labs(
    title = "Peak bloom (since Jan 1st) in Zurich, Switzerland",
    x = NULL, y = NULL
  ) +
  theme_light() +
  facet_grid(rows = vars(str_to_title(location)))

print(fig_meteoswiss)

ggsave(filename = "figures/fig_meteoswiss.png", plot = fig_meteoswiss, bg = "white")
```

#### NPN

Two additional files are provided by the **USA National Phenology Network (USA-NPN)** and the many participants who contribute to its *Nature’s Notebook* program.

The file `USA-NPN_individual_phenometrics_data.csv`contains observations on **individual phenometrics** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-individual-phenometrics

npn_ind_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv") |> 
  mutate(Observation_Date = mdy(Observation_Date))

npn_ind_des <- read_csv("data/USA-NPN_individual_phenometrics_datafield_descriptions.csv")
```

The file `USA-NPN_status_intensity_data.csv` contains observations on the **status and intensity** of cherry trees. It is available at <http://doi.org/10.5066/F78S4N1V>.

```{r}
#| label: load-npn-status-intensity

npn_int_obs <- read_csv("data/USA-NPN_status_intensity_observations_data.csv")

npn_int_des <- read_csv("data/USA-NPN_status_intensity_datafield_descriptions.csv")
```

These files can help augment the existing data sets for which you don't have many observations such as New York City (and perhaps even Vancouver). 

## Maps

### Main locations

First, we plot the five locations in the `cherry.csv` file

```{r}
#| label: map-main-locations
#| eval: false

cherry_coord <- cherry |> 
  group_by(location) |> 
  slice_tail(n = 1)

cherry_map <- cherry_coord |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(cherry_map, zcol = "alt")
```

### Japan

Next, we plot the map of all the locations in the `japan.csv` file

```{r}
#| label: map-japan
#| eval: false

crd_ch <- japan |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_ch <- crd_ch |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_ch, zcol = "n")
```

### Korea

Second, we plot the locations in the `south_korea.csv` file

```{r}
#| label: map-korea
#| eval: false

crd_korea <- korea |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_korea <- crd_korea |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_korea, zcol = "n")
```

### Switzerland

Third, we plot the locations in the `meteoswiss.csv` file

```{r}
#| label: map-meteoswiss
#| eval: false

crd_ch <- meteoswiss |> 
  filter(!is.na(bloom_doy)) |> 
  group_by(location, long, lat, alt) |> 
  summarise(n = n())

map_ch <- crd_ch |> 
  st_as_sf(coords = c("long", "lat"), crs = 4326)

mapview(map_ch, zcol = "n")
```

## Time trend

We can explore models without external predictors that rely solely on past observations and on time trends.

A simple method to predict peak bloom date in the future is to fit a least-squares line through the observed dates and extrapolate the regression function. We want to have a separate line for each location, hence we tell R to estimate *interaction* effects. Note that R will automatically exclude `NA`s from the regression.

We only use data **from 1880** to fit the trends, as prior data may not be as reliable/relevant.

```{r}
#| label: fit-linear-time-trend

ls_fit <- lm(
  formula = bloom_doy ~ 0 + location + location:year,
  data = cherry, 
  subset = year >= 1880
)
```

This simple linear regression functions suggest a trend toward earlier peak bloom at the sites.

```{r}
#| label: summary-linear-model

summary(ls_fit)
```

Unsurprisingly, neither the intercept nor the time trend are statistically significant for Vancouver and New York City, where we have few observations.

We can compute predictions for both the historical observations and for the current year with the `predict()` function. We first use the `expand_grid()` function to create all possible combinations for the years between 1880 and 2025 and the five locations, so that we have the input variables for both the fitted values for the past years and the prediction for the current year

```{r}
#| label: create-grid

cherry_grid <- expand_grid(
  location = unique(cherry$location),
  year = 1880:2025
)
```

Then we use these input variables to create both point estimates and 90% confidence intervals with `predict()`

```{r}
#| label: create-predictions

predictions <- cherry_grid |>
  bind_cols(predict(
    object = ls_fit,
    newdata = cherry_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr)
```

Plot only the fitted values, i.e., the predictions for the actual observations between 2015 and 2024

```{r}
#| label: fig-fitted
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from simple linear regression models fitted to
#|   five sites.

cherry |>
  right_join(predictions, by = c("year", "location")) |>
  filter((location == "vancouver" & year >= 2015) | (location != "vancouver" & year >= 2015)) |>
  filter((location == "newyorkcity" & year >= 2019) | (location != "newyorkcity" & year >= 2015)) |> 
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = c(2015, 2018, 2021, 2024)) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that the slope of the time trend has become steeper since 2015. If we assume that this trend continues, our prediction for the current year will be too high.

We create a function called `day_to_date()` that converts `bloom_doy` into an ISO date

```{r}
#| label: day_to_date

#' Small helper function to convert the day of year to the actual date.
#'
#' @param year year as an integer
#' @param doy day of the year as integer (1 means January 1st)
#' @return date string

doy_to_date <- function(year, doy) {
  strptime(paste(year, doy, sep = "-"), "%Y-%j") |> # create date object
    strftime("%Y-%m-%d") # translate back to date string in ISO 8601 format
}
```

Since we treated the outcome variable as continuous, we need to use `floor()` and `ceiling()` to create upper and lower bounds for our point estimate, for which we use `round()`.

Based on this simple model, the peak bloom dates at these sites in 2025 are predicted to occur on

```{r}
#| label: tbl-predictions
#| tbl-cap: "Predicted peak bloom dates at the four sites"

predictions |>
  filter(year == 2025) |>
  mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  ) |> 
  kable() |> 
  kable_styling()
```

### Averaging

We have only a few observations for Vancouver, BC and for New York City. Although the climate in Vancouver and NYC is different from the other locations, the simplest way to borrow information from the other locations is to average across these three sites.

Hence, we want to fit a straight line through the peak bloom dates, ignoring the actual site. That is, we fit a simple linear time trend on all available observations, but we want to use larger **weights** for observations from the target sites.

#### Vancouver

To run Weighted Least Squares (WLS), the `weights` inside `stats::lm()` argument must be a vector of the same length as the input data and we can use the names of the `location` column to create it. 

We could have created a new variable explicitly with:

```{r}
#| eval: false

cherry$weight <- (cherry$location == "vancouver") + 0.2 * (cherry$location != "vancouver")

cherry$weight <- NULL
```

So the weight put on observations from Vancouver are while observations from other locations receive a weight of only one fifth.

```{r}
#| label: fit-wls-vancouver

ls_fit_for_van <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "vancouver") + 0.2 * (location != "vancouver")
)
```

Create a grid with the input variables for Vancouver only

```{r}
#| label: grid-vancouver

vancouver_grid <- tibble(location = "vancouver", year = 2008:2025)
```

Then we use these inputs to create predictions for Vancouver, 

```{r}
#| label: predict-vancouver

predictions_vancouver <- vancouver_grid |>
  bind_cols(predict(
    object = ls_fit_for_van,
    newdata = vancouver_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr) |> 
  mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  )
```

Not surprisingly, the predicted peak bloom date for Vancouver and NYC is now very similar to the other 3 sites:

```{r}
#| label: tbl-predictions-vancouver
#| tbl-cap: "Predictions for Vancouver"

predictions_vancouver |> 
  kable() |> 
  kable_styling()
```

Due to **rounding** and the **slope** not being very steep, the predictions are almost the same for every year.

We can plot the fitted values against the actual data from previous years

```{r}
#| label: fig-fitted-vancouver
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for Vancouver using data from all four sites.

cherry |>
  right_join(y = predictions_vancouver, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2008:2025) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

We see that without including weather data, we are not able to capture the unusually warm springs in Vancouver, such as the one from 2010.

We have made this very cautious prediction since we don't know much about the conditions in Vancouver.

If we are satisfied with these weighted predictions for Vancouver, we can use them to replace the predictions that were made using only data from Vancouver.

```{r}
#| label: replace-vancouver-predictions

predictions <- predictions |>
  filter(location != "vancouver") |>
  bind_rows(predictions_vancouver)
```

#### New York

We once again fit a model on the whole data set, weighting observations from New York City with one and observations from other locations with one fifth.

```{r}
#| label: tbl-predictions-nyc
#| tbl-cap: "Predictions for New York City"

ls_fit_for_nyc <- lm(
  formula = bloom_doy ~ year,
  data = cherry, subset = year >= 1880,
  weights = (location == "newyorkcity") + 0.2 * (location != "newyorkcity")
)

# Add inputs since 2019 and the current year
nyc_grid <- tibble(location = "newyorkcity", year = 2019:2025)

predictions_nyc <- nyc_grid |>
  bind_cols(predict(
    object = ls_fit_for_nyc,
    newdata = nyc_grid,
    interval = "prediction", level = 0.9
  )) |>
  rename(prediction = fit, lower = lwr, upper = upr) |> 
    mutate(
    prediction = round(prediction),
    lower = floor(lower),
    upper = ceiling(upper),
    prediction_date = doy_to_date(year, prediction)
  )

predictions_nyc |> 
  kable() |> 
  kable_styling()
```

Due to rounding, and because the slope is not very steep, the predictions are the same for every year.

We plot the fitted values for New York City and the actual observations

```{r}
#| label: fig-fitted-nyc
#| fig-width: 8
#| fig-height: 3
#| out-width: 100%
#| fig-cap: |
#|   Predictions and 90% prediction intervals from a simple linear regression model for NYC using data from all five sites.

cherry |>
  right_join(predictions_nyc, by = c("year", "location")) |>
  ggplot(aes(x = year, y = prediction, ymin = lower, ymax = upper)) +
  geom_line(linewidth = 1) +
  geom_ribbon(color = "black", linetype = "22", linewidth = 0.8, fill = NA) +
  geom_point(aes(y = bloom_doy)) +
  scale_x_continuous(breaks = 2019:2024) +
  facet_grid(cols = vars(str_to_title(location))) +
  labs(x = NULL, y = "Peak bloom (days since Jan 1st)") +
  theme_light()
```

Again, this model does not capture the steepness of the slope of the time trend accurately.

If we are satisfied with the averaged predictions we can replace the predictions that were made using only data from New York City

```{r}
#| label: replace-nyc-predictions

predictions <- predictions |>
  filter(location != "newyorkcity") |>
  bind_rows(predictions_nyc)
```


## Predictors

We have seen that a linear time trend alone is not sufficient to explain the variance in the peak bloom times. It therefore makes sense to include external data on local weather conditions in our model. 

### POWER

We use the `{nasapower}` package described here <https://docs.ropensci.org/nasapower/> to access the [NASA Prediction Of Worldwide Energy Resources (POWER) API](https://power.larc.nasa.gov/).

We use data from the **Modern-Era Retrospective analysis for Research and Applications, Version 2 (MERRA-2)** which includes data at a daily frequency since January 1st, 1981.

The database is documented here <https://gmao.gsfc.nasa.gov/reanalysis/MERRA-2/>.

The POWER data base is documented here <https://power.larc.nasa.gov/docs/>.

The proper citation is here <https://gmao.gsfc.nasa.gov/reanalysis/MERRA-2/citing_MERRA-2/>.

The Parameter Dictionary includes all the available variables <https://power.larc.nasa.gov/beta/parameters/>.

We create a custom function called `get_power()` to repeatedly query the API with the `nasapower::get_power()` function.

```{r}
#| label: get_power

get_power <- function(vars, start = "1981-01-01", end = as.character(Sys.Date()), freq = "daily", lat, long) {
  
  power <- map2(
      .x = long,
      .y = lat,
      .f = function(long = .x, lat = .y) {
        res <- nasapower::get_power(
          community = "ag",
          lonlat = c(long, lat),
          pars = vars,
          dates = c(start, end),
          temporal_api = freq
        )
        return(res)
      }
    ) |>
    list_rbind()
  
  return(power)
}
```

Specify the parameters of the NASA POWER API request

We load the following variables:

- `WS10M`: Wind Speed at 10 Meters (m/s) 

- `FROST_DAYS`: Frost days. A frost day occurs when the 2m temperature cools to the dew point temperature and both are less than 0 C or 32 F.

- `TO3`: Total Column Ozone. The total amount of ozone in a column extending vertically from the earth's surface to the top of the atmosphere.

- `WS2M`: Wind Speed at 2 Meters. The average of wind speed at 2 meters above the surface of the earth.

- `DAY_HOURS`: Day Hours. The Climatological average number of hours when solar irradiance is present.

- `QV2M`: Specific Humidity at 2 Meters. The ratio of the mass of water vapor to the total mass of air at 2 meters.

- `RH2M`: Relative Humidity at 2 Meters. The ratio of vapor pressure to the saturation vapor pressure with respect to a plane surface of pure water, expressed in percent.

- `PRECTOTCORR`: Precipitation Corrected. The average MERRA-2 bias corrected total precipitation at the surface of the earth.

- `T2M`: Temperature at 2 Meters. The average air (dry bulb) temperature at 2 meters above the surface of the earth.

- `T2M_MAX:` Temperature at 2 Meters Maximum. The maximum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `T2M_MIN`: Temperature at 2 Meters Minimum. The minimum hourly air (dry bulb) temperature at 2 meters above the surface of the earth in the period of interest.

- `CDD0`: Cooling Degree Days Above 0 C. The daily accumulation of degrees above a 0-degree-Celsius threshold when the daily mean temperature is above the threshold.

- `CDD18_3`: Cooling Degree Days Above 18.3 C. The daily accumulation of degrees above a 0-degree-Celsius threshold when the daily mean temperature is above the threshold.

- `HDD0`: Heating Degree Days Below 0 C. The daily accumulation of degrees below a 0-degree-Celsius threshold when the daily mean temperature is below the threshold.

- `HDD18_3`: Heating Degree Days Below 18.3 C. The daily accumulation of degrees below an 18.3-degree-Celsius threshold when the daily mean temperature is below the threshold.

- `PS`: Surface pressure (kPa)

Learn more about heating degree days at <https://en.wikipedia.org/wiki/Heating_degree_day>.

```{r}
#| label: specify-vars

vars <- c(
  "T2M",
  "T2M_MAX",
  "T2M_MIN",
  "PRECTOTCORR",
  "RH2M",
  "PS",
  "WS10M"
)
```

We retrieve the **coordinates** which we will use to match to the nearest weather stations

```{r}
#| label: get-coordinates

crd_kyoto        <- distinct(kyoto, lat, long)
crd_liestal      <- distinct(liestal, lat, long)
crd_nyc          <- distinct(newyorkcity, lat, long)
crd_vancouver    <- distinct(vancouver, lat, long)
crd_washingtondc <- distinct(washingtondc, lat, long)

crd_japan      <- distinct(japan, lat, long)
crd_korea      <- distinct(korea, lat, long)
crd_meteoswiss <- distinct(meteoswiss, lat, long)

crd_npn1 <- npn_ind_obs |> 
  select(lat = Latitude, long = Longitude) |> 
  distinct()

crd_npn2 <- npn_int_obs |> 
  select(lat = Latitude, long = Longitude) |> 
  distinct()

crd_npn <- bind_rows(crd_npn1, crd_npn2) |> 
  distinct()
```

Load NASA POWER data for Kyoto, Japan

```{r}
#| label: load-main-nasa-power

power_kyoto <- get_power(vars, lat = crd_kyoto$lat, long = crd_kyoto$long)
power_kyoto$location <- "kyoto"

power_liestal <- get_power(vars, lat = crd_liestal$lat, long = crd_liestal$long)
power_liestal$location <- "liestal"

power_nyc <- get_power(vars, lat = crd_nyc$lat, long = crd_nyc$long)
power_nyc$location <- "newyorkcity"

power_vancouver <- get_power(vars, lat = crd_vancouver$lat, long = crd_vancouver$long)
power_vancouver$location <- "vancouver"

power_washingtondc <- get_power(vars, lat = crd_washingtondc$lat, long = crd_washingtondc$long)
power_washingtondc$location <- "washingtondc"

power_cherry <- bind_rows(
  power_kyoto,
  power_liestal,
  power_nyc,
  power_vancouver,
  power_washingtondc
)

write_csv(power_cherry, file = "data/power_cherry.csv")
```

We also load predictors for the additional data sets. Because these take a long time to load, instead load data from a previous run

```{r}
#| label: load-additional-nasa-power
#| eval: false

power_japan <- get_power(vars, lat = crd_japan$lat, long = crd_japan$long)
write_csv(power_japan, file = "data/power_japan.csv")

power_korea <- get_power(vars, lat = crd_korea$lat, long = crd_korea$long)
write_csv(power_korea, file = "data/power_korea.csv")

power_meteoswiss <- get_power(vars, lat = crd_meteoswiss$lat, long = crd_meteoswiss$long)
write_csv(power_meteoswiss, file = "data/power_meteoswiss.csv")

power_npn <- get_power(vars, lat = crd_npn$lat, long = crd_npn$long)
write_csv(power_npn, file = "data/power_npn.csv")
```

Let's have a look at the NASA POWER dataset for the five locations. We change some of the column names and convert the variable names to lowercase letters.

```{r}
#| label: load-power_cherry

power_cherry <- read_csv("data/power_cherry.csv")

colnames(power_cherry) <- tolower(colnames(power_cherry))

power_cherry <- power_cherry |> 
  rename(date = yyyymmdd, prcp = prectotcorr) |> 
  select(-c(lat, lon, year, mm, dd, doy)) |> 
  relocate(location, date)
```

## Open Meteo

Use the `{openmeteo}` package to retrieve weather forecasts from the Open Meteo API at <https://open-meteo.com/>.

- `weather_forecast()` - retrieve weather forecasts for a location
- `weather_history()` - retrieve historical weather observations for a location
- `weather_now()` - simple function to return current weather for a location
- `weather_variables()` - retrieve a shortlist of valid forecast or historical weather variables
provided

Find the weather stations located near the cherry tree locations

```{r}
kyoto_gc <- geocode(location_name = "kyoto", n_results = 1, language = "en", silent = TRUE)

liestal_gc <- geocode(location_name = "liestal", n_results = 1, language = "en", silent = TRUE)

newyorkcity_gc <- geocode(location_name = "new york city", n_results = 1, language = "en", silent = TRUE)

vancouver_gc <- geocode(location_name = "vancouver", n_results = 1, language = "en", silent = TRUE)

washingtondc_gc <- geocode(location_name = "washington", n_results = 1, language = "en", silent = TRUE)

locations_gc <- list(
  kyoto_gc,
  liestal_gc,
  newyorkcity_gc,
  vancouver_gc,
  washingtondc_gc
)
```


Check out the available variables

```{r}
available_variables <- weather_variables()
```

Hourly forecast variables

```{r}
available_variables$hourly_forecast_vars
```

Daily forecast variables

```{r}
available_variables$daily_forecast_vars
```


```{r}
get_fcst <- function(loc_gc) {
  
  fcst <- weather_forecast(
    location = c(loc_gc$latitude, loc_gc$longitude), 
    start = as.character(today() - days(7)), 
    end = "2025-03-09", 
    hourly = c(
      "temperature_2m", 
      "relative_humidity_2m", 
      "pressure_msl",
      "wind_speed_10m"
    ),
    daily = c(
      "temperature_2m_max",
      "temperature_2m_min",
      "precipitation_sum"
    )
  )
  
  fcst$location <- tolower(loc_gc$name)
  
  return(fcst)
}

locations_fcst <- purrr::map_dfr(.x = locations_gc, .f = get_fcst)
View(locations_fcst)
locations_fcst <- locations_fcst |> 
  rename(
    t2m = hourly_temperature_2m,
    t2m_max = daily_temperature_2m_max,
    t2m_min = daily_temperature_2m_min,
    prcp = daily_precipitation_sum,
    rh2m = hourly_relative_humidity_2m,
    ps = hourly_pressure_msl,
    ws10m = hourly_wind_speed_10m
  ) |> 
  group_by(location, date) |> 
  summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  relocate(location, date) |> 
  mutate(
    location = case_match(location,
      "new york" ~ "newyorkcity",
      "washington" ~ "washingtondc", 
      .default = location
    )
  )
```

Add the weather forecast to the predictors

```{r}
power_cherry <- power_cherry |> 
  full_join(
    y = locations_fcst, 
    by = join_by(location, date, t2m, t2m_max, t2m_min, prcp, rh2m, ps, ws10m)
  )
```

Check out if the forecast was appended correctly

```{r}
power_cherry |> 
  tail(n = 20)
```


## AccuWeather

It is likely that the weather in March and April is a very important predictor for the peak bloom date of cherry blossoms. However, because the submission date is February 28, we won't have any data available for these two months.

We can either train our model exclusively on data from January 1st 1981 to January 27th of the current year. Or we can include data for April and March and use Weather forecasts to extend our data set for the current year. 

This is still dangerous, as the weather forecast is likely to be inaccurate. And because the past spring months are actual weather observations and not forecasts, the model will likely give too much weight to the current forecast.

AccuWeather forecasts daily maximum and minimum temperatures three months into the future.  We do not evaluate the quality of these forecasts.

We use the AccuWeather forecast as it appeared on the webpage [AccuWeather.com](https://www.accuweather.com/en/be/brussels/27581/january-weather/27581?year=2023) on February 19, 2023. 

AccuWeather reports the forecast for each month on a separate webpage. 

For reproducibility, we saved each page on the [Internet Archive](https://web.archive.org/web/20230219151906/https://www.accuweather.com/en/be/brussels/27581/january-weather/27581?year=2023&unit=c). 

The following R code creates the function `get_weather_table()` to retrieve each page we saved, extract the forecast contained within that page, and arrange the data as a tibble.

The `get_weather_table()` function combines several functions from the `{rvest}` package.

In particular, the forecast on each page is contained within the div `monthly-calendar` and can be extracted with the `html_nodes()` and `html_text2()` functions.

Applying the `get_weather_table()` function to the URL for each page yields a five column tibble `temp_*`, with columns defined in the same way as the tibble `temp`, discussed in previous sections. 

The first 10 rows are below; the data are also available on the author's [GitHub](https://raw.githubusercontent.com/jauerbach/miscellaneous/main/temp_br.csv).

We now predict the day the lilacs will bloom. The R code below uses the `doy_prediction` and `doy_last_frost` functions created in earlier sections and displays the prediction in a table. 

The forecast is easily updated by providing the url to the updated AccuWeather webpage. 

(You might use the url [https://web.archive.org/save](https://web.archive.org/save) to save a webpage to the Internet Archive to ensure your work is reproducible.)


```{r}
#| label: get_weather_table

get_weather_table <- function(url) {
  read_html(url) |> 
    html_nodes("div.monthly-calendar") |> 
    html_text2() |> 
    str_remove_all("°|Hist. Avg. ") |> 
    str_split(" ", simplify = TRUE) |> 
    parse_number() |> 
    matrix(
      ncol = 3,
      byrow = TRUE,
      dimnames = list(NULL, c("day", "tmax", "tmin"))
    ) |> 
    as_tibble() |> 
    filter(
      row_number() %in%
        (which(diff(day) < 0) %>% (function(x) if (length(x) == 1) seq(1, x[1], 1) else seq(x[1] + 1, x[2], 1)))
    )
}
```

Get the weather forecast for Kyoto, Japan

```{r}
#| label: get_temp_kyoto
#| eval: false

curr_year <- year(Sys.Date())

temp_kyoto <- tibble(
  base_url = "https://web.archive.org/web/20250221214619/https://www.accuweather.com/en/jp/kyoto-shi/2-224436_1_al/",
  month = month.name[2:4],
  year = curr_year,
  url = str_c(base_url, tolower(month), "-weather/2-224436_1_al?year=", year)
  ) |> 
  mutate(temp = map(url, get_weather_table)) |> 
  pull(temp) |> 
  reduce(bind_rows) |> 
  mutate(
    location = "kyoto",
    date = as.Date("2025-02-01"),
    date = date + days(row_number() - 1),
    year = parse_number(format(date, "%Y")),
    t2m_max = round((tmax - 32) * 5/9, 1),
    t2m_min = round((tmin - 32) * 5/9, 1),
    t2m_range = (t2m_max + t2m_min) / 2,
    .keep = "none"
  )
```

Get the weather forecast for Liestal, Switzerland

```{r}
#| label: get_temp_liestal
#| eval: false

temp_liestal <- tibble(
  base_url = "https://web.archive.org/web/20250221224017/https://www.accuweather.com/en/ch/liestal/311994/",
  month = month.name[2:4],
  year = curr_year,
  url = str_c(base_url, tolower(month), "-weather/311994?year=", year)
  ) |> 
  mutate(temp = map(url, get_weather_table)) |> 
  pull(temp) |> 
  reduce(bind_rows) |> 
  mutate(
    location = "liestal",
    date = as.Date("2025-02-01"),
    date = date + days(row_number() - 1),
    year = parse_number(format(date, "%Y")),
    t2m_max = round((tmax - 32) * 5/9, 1),
    t2m_min = round((tmin - 32) * 5/9, 1),
    t2m_range = (t2m_max + t2m_min) / 2,
    .keep = "none"
  )
```

Get the weather forecast for New York City, USA

```{r}
#| label: get_temp_newyorkcity
#| eval: false

temp_newyorkcity <- tibble(
  base_url = "https://web.archive.org/web/20250221230235/https://www.accuweather.com/en/us/new-york/10021/",
  month = month.name[2:4],
  year = curr_year,
  url = str_c(base_url, tolower(month), "-weather/14-349727_1_al?year=", year)
  ) |> 
  mutate(temp = map(url, get_weather_table)) |> 
  pull(temp) |> 
  reduce(bind_rows) |> 
  mutate(
    location = "newyorkcity",
    date = as.Date("2025-02-01"),
    date = date + days(row_number() - 1),
    year = parse_number(format(date, "%Y")),
    t2m_max = round((tmax - 32) * 5/9, 1),
    t2m_min = round((tmin - 32) * 5/9, 1),
    t2m_range = (t2m_max + t2m_min) / 2,
    .keep = "none"
  )
```

Get the weather forecast for Vancouver BC, Canada

```{r}
#| label: get_temp_vancouver
#| eval: false

temp_vancouver <- tibble(
  base_url = "https://web.archive.org/web/20250222014224/https://www.accuweather.com/en/ca/vancouver/v6c/",
  month = month.name[2:4],
  year = curr_year,
  url = str_c(base_url, tolower(month), "-weather/53286?year=", year)
  ) |> 
  mutate(temp = map(url, get_weather_table)) |> 
  pull(temp) |> 
  reduce(bind_rows) |> 
  mutate(
    location = "vancouver",
    date = as.Date("2025-02-01"),
    date = date + days(row_number() - 1),
    year = parse_number(format(date, "%Y")),
    t2m_max = round((tmax - 32) * 5/9, 1),
    t2m_min = round((tmin - 32) * 5/9, 1),
    t2m_range = (t2m_max + t2m_min) / 2,
    .keep = "none"
  )
```

Get the weather forecast for Washington D.C., USA

```{r}
#| label: get_temp_washingtondc
#| eval: false

temp_washingtondc <- tibble(
  base_url = "https://web.archive.org/web/20250221235102/https://www.accuweather.com/en/us/washington/20006/",
  month = month.name[2:4],
  year = curr_year,
  url = str_c(base_url, tolower(month), "-weather/18-327659_1_al?year=", year)
  ) |> 
  mutate(temp = map(url, get_weather_table)) |> 
  pull(temp) |> 
  reduce(bind_rows) |> 
  mutate(
    location = "washingtondc",
    date = as.Date("2025-02-01"),
    date = date + days(row_number() - 1),
    year = parse_number(format(date, "%Y")),
    t2m_max = round((tmax - 32) * 5/9, 1),
    t2m_min = round((tmin - 32) * 5/9, 1),
    t2m_range = (t2m_max + t2m_min) / 2,
    .keep = "none"
  )
```

Combine the weather forecasts for the five locations

```{r}
#| label: combine-temp_fcst
#| eval: false

temp_fcst <- bind_rows(
  temp_kyoto,
  temp_liestal,
  temp_newyorkcity,
  temp_vancouver,
  temp_washingtondc
)

write_csv(temp_fcst, file = "data/temp_fcst.csv")
```

Read in the temperature forecasts for the five main locations

```{r}
#| label: read-temp_fcst

temp_fcst <- read_csv("data/temp_fcst.csv")
```

Create a table of the temperature forecasts

```{r}
temp_fcst |> 
  relocate(year) |> 
  kable(
    dig = 2,
    align = "c",
    col.names = c(
      "location", "year", "date", "maximum temp (°C)",
      "minimum temp (°C)", "midrange temp (°C)"
    )
  ) |> 
  kable_styling() |> 
  scroll_box(width = "100%", height = "400px")
```


Add the minimum and maximum temperature forecast data to the variables. Replace the missing temperature with the midrange temperature forecast.

```{r}
#| label: add-temp_fcst-to-power_cherry

power_cherry <- power_cherry |> 
  full_join(y = temp_fcst, by = join_by(location, date, t2m_max, t2m_min)) |> 
  mutate(t2m = if_else(is.na(t2m), (t2m_max + t2m_min) / 2, t2m))
```

## Feature engineering

We create some additional features from our predictors

### Midrange temperature

We create the mid-range temperature, the average of the maximum and the minimum average temperature measured on a specific day.

```{r}
#| label: create-t2m_range

power_cherry <- power_cherry |> 
  mutate(t2m_range = (t2m_max + t2m_min) / 2)
```

### Aggregated precipitation since January 1st

We use the variable daily measured precipitation (in mm) to calculate accumulated precipitation (in mm) starting from January 1st of the year of the observation date. 

```{r}
#| label: create-aprcp

power_cherry <- power_cherry |> 
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(aprcp = cumsum(prcp)) |> 
  ungroup() |> 
  select(-year)
```

### Aggregated growing degree days since January 1st

We create the accumulated growing degree days (AGDD) as the sum of growing degree days (GDD) since January 1st, where GDD are calculated as the number of degrees C by which each day's mid-range temperature exceeds a base temperature, which can be 0 degrees C, or a species-specific base temperature, such as 4.3 degrees C for Yoshino cherry trees. If the mid-range temperature does not exceed the base temperature on a given day, no growing degree days are counted. See more at <https://en.wikipedia.org/wiki/Growing_degree-day>.

```{r}
#| label: create-agdd

base_temp <- 4.3

power_cherry <- power_cherry |> 
  mutate(gdd = pmax(t2m_range - base_temp, 0)) |>
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(agdd = cumsum(gdd)) |> 
  ungroup() |>
  select(-c(year, gdd))
```

### Aggregated chilling degree days since October 1st

We can calculate chilling degree days (CDD)** and accumulated chilling degree days (ACDD) since October 1st. We count CDDs whenever the mid-range temperature is below a threshold of 7 degrees Celsius, and we accumulate until December 31st of each year. If the mid-range temperature stays above the threshold, no GDDs are counted. See more at <https://en.wikipedia.org/wiki/Chilling_requirement>.

```{r}
#| label: create-acdd

temp_threshold <- 7

power_acdd <- power_cherry |> 
  mutate(year = year(date)) |> 
  filter(date >= paste0(year, "-09-01")) |> 
  mutate(cdd = pmax(0, temp_threshold - t2m_range)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(acdd = cumsum(cdd)) |> 
  ungroup() |> 
  select(-year)

power_cherry <- power_cherry |> 
  left_join(y = power_acdd |> select(location, date, acdd), join_by(location, date))

power_cherry <- power_cherry |> 
  mutate(acdd = if_else(is.na(acdd), 0, acdd))
```


## Select timespan

We may restrict our predictors to a timespan that includes only winter and spring months.

Furthermore, we need to decide if we want to exclude all data before the end of February to make our model more conservative, or if we want to include the months of March and April, and append the data for the current year with weather forecasts.

```{r}
#| eval: false

power_cherry <- power_cherry |> 
  filter(month(date) %in% c(9, 10, 11, 12, 1, 2, 3, 4, 5))
```

## Temporal aggregation

We need to decide at which temporal granularity we want our predictors.

### Seasons

```{r}
#| label: aggregate-to-seasons

df_seasons <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .month = month(date)) |> 
  mutate(
    season = case_when(
      .month %in% c(12, 1, 2) ~ "winter",
      .month %in% c(3, 4, 5) ~ "spring",
      .month %in% c(6, 7, 8) ~ "summer",
      .month %in% c(9, 10, 11) ~ "winter"
    )
  ) |> 
  mutate(.year = if_else(.month == 12, .year - 1, .year)) |>  
  mutate(.year_season = paste0(.year, "-", season)) |> 
  relocate(location, date, .year, .month, season, .year_season, variable, value) |> 
  group_by(location, .year, season, .year_season, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, season),
    names_glue = "{variable}_{season}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year) |> 
  filter(year > 1980)

cherry_seasons <- cherry |> 
  inner_join(y = df_seasons, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Months

```{r}
#| label: aggregate-to-months

df_months <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .month = month(date)) |> 
  mutate(.year_month = paste0(.year, "-", .month)) |> 
  relocate(location, date, .year, .month, .year_month, variable, value) |> 
  group_by(location, .year, .month, .year_month, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .month),
    names_glue = "{variable}_{.month}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_months <- cherry |> 
  inner_join(y = df_months, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Weeks

```{r}
#| label: aggregate-to-weeks

df_weeks <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .week = week(date)) |> 
  mutate(.year_week = paste0(.year, "-", .week)) |> 
  relocate(location, date, .year, .week, .year_week, variable, value) |> 
  group_by(location, .year, .week, .year_week, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .week),
    names_glue = "{variable}_{.week}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_weeks <- cherry |> 
  inner_join(y = df_weeks, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```

### Days

```{r}
#| label: aggregate-to-days

df_days <- power_cherry |> 
  pivot_longer(cols = -c(location, date), names_to = "variable", values_to = "value") |> 
  mutate(.year = year(date), .day = day(date)) |> 
  mutate(.year_month = paste0(.year, "-", .day)) |> 
  relocate(location, date, .year, .day, .year_month, variable, value) |> 
  group_by(location, .year, .day, .year_month, variable) |> 
  summarise(value = mean(value), .groups = "keep") |> 
  ungroup() |> 
  pivot_wider(
    names_from = c(variable, .day),
    names_glue = "{variable}_{.day}",
    values_from = value
  ) |>
  group_by(location, .year) |> 
   summarise(across(where(is.numeric), ~ mean(.x, na.rm = TRUE))) |> 
  ungroup() |> 
  rename(year = .year)

cherry_days <- cherry |> 
  inner_join(y = df_days, join_by(year, location)) |> 
  select(-c(lat, long, alt, bloom_date))
```


## Exploratory data analysis (EDA)

### Linear model

A simple linear model can help us understand how predictive a feature is

Plot `t2m_max_winter`

```{r}
#| label: fig-t2m_max_winter
#| fig-cap: "Average maximum winter temperature (Dec. 1st of the previous year until end of February)"

cherry_seasons |>
  ggplot(aes(x = year, y = t2m_max_winter)) +
  geom_line() +
  labs(x = NULL, y = "Average maximum winter temperature (°C)") +
  facet_grid(rows = vars(location)) +
  scale_x_continuous(limits = c(1970, 2025)) +
  theme_light()
```

We fit a linear model with different intercepts and time trends for each location and a weather effect. We can add interaction effects to the temperature data too if we want.

```{r}
#| label: fit-lin-t2m_max_winter

lm_mod <- linear_reg(mode = "regression", engine = "lm")

lm_fit <- lm_mod |> 
  fit(
    formula = bloom_doy ~ 0 + location + location:year + location:t2m_max_winter, 
    data = cherry_seasons
  )
```

Create fitted values for the bast observations

```{r}
#| label: predictions-avg_tmax

mean_pred <- predict(object = lm_fit, new_data = cherry_seasons)
```

Let's plot the predictions and the actual data

```{r}
#| label: fig-predictions-avg_tmax
#| fig-cap: "Predictions from seasonal predictors"

cherry_seasons |> 
  bind_cols(mean_pred) |> 
  ggplot(aes(x = year)) +
  geom_point(aes(y = bloom_doy)) +
  geom_line(aes(y = .pred)) +
  facet_grid(rows = vars(location)) +
  theme_light()
```

We see that the model with the linear time trend and the predictor `t2m_max_winter` does predict the data better than with only the linear time trend. We can assume that a model with `t2m_max_spring` would explain the data even better, but unfortunately, we don't have data for the spring of this year.

We also see that the average temperature in Kyoto is too mild to explain much of the variance in peak bloom times.


### Graphical

```{r}
cherry_seasons |>
  ggplot(aes(x = year, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

```{r}
cherry_seasons |>
  ggplot(aes(x = agdd_spring, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

```{r}
cherry_seasons |>
  ggplot(aes(x = acdd_winter, y = bloom_doy, color = location)) +
  geom_point() +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE) +
  theme_bw() +
  theme(legend.position = "top")
```

We only see the expected effect of aggregated chilling degrees for Vancouver, which is not a lot. We could refine the calculation of this predictor.

### Correlations

We use the `{corr}` package to check the correlations of the predictors.

```{r}
cherry_seasons |>
  correlate() |>
  gt() |>
  fmt_number() |>
  tab_header(title = "correlations") |>
  tab_footnote(
  footnote = ""
)
```

### Decision tree

Fit a decision tree for exploratory data analysis. We use the `{parsnip}` package to create a model that runs on the `"rpart"` engine, and we use the `rpart.plot` package to plot the decision tree.

```{r}
set.seed(456)

# Create a {parsnip} model that runs on the "rpart" engine
cart_specs <- decision_tree(
  tree_depth = 10,
  cost_complexity = 0.001
  ) |>
  set_mode("regression") |>
  set_engine("rpart")

# fit the model 
cart_fit <- cart_specs |>
  fit(
    formula = bloom_doy ~ ., 
    data = cherry_days
  )

# Visualize with {rpart.plot}
cart_fit$fit |> 
  rpart.plot()
```

On a seasonal aggregation level, aggregated growing degree days (AGDD) are the most important predictor.

### Variable importance plot (VIP)

Use the `{vip}` package to create a variable importance plot (VIP)

```{r}
cart_fit |> 
  vip()
```

AGDDs in winter and spring are the most important predictor.

# section 3

How to apply tabular ml?

## Workflows

- while fitting a one-off model like above for EDA was fine
- if we want to quickly iterate pre-, post- and parameter tuning
- workflow objects via `{tidymodels}` is necessary

## Data Splitting

- ML modeling processes benefit from data splitting
- data splitting usually involves splitting the initial data into at least two parts: training and testing, sometimes a third called validation
- defining data that the model can see (training) and data held away (validation & testing)
- training data will be used to *train* the model
- testing/validation data is held out to *test* and *calibrate* the accuracy of the trained model
- this is an important step to measure over/under fitting and determine post-processes

```{r}
set.seed(31)

data_split <- initial_validation_split(magpie_data)

training <- training(data_split)
validate <- validation(data_split)
testing  <- testing(data_split)
```

## Feature Engineering and Pre-Processing

- a crucial step before training is feature engineering
- feature engineering can come in many flavors
- these steps can be used to scale data, extract components, adjustments, cross-effects, etc.
- some features come from industry knowledge
- some come from EDA, training, testing, iterating
- recipes package is designed to help the ML practitioner with pre-processing

```{r}
magpie_recipe <- recipe(formula = magpie_count ~ ., data = training) |>
  step_corr(threshold = 0.90) |>
  step_dummy(all_nominal_predictors()) |>
  step_YeoJohnson(all_nominal_predictors())
```

## Training Stage, Part 1

- within the training stage we will evaluate hyperparameter estimations, model engines, & final performance
- *that's a lot of moving pieces!*
- this means we will most likely need to split our training data into *many pieces*
- this is done via cross-validation
- this is cost effective, we want to test as many possibilities in the shortest amount of time and compute
- within the training data we will have *analysis* and *assessment* sets

## Training Stage, Part 2

- let's create our cross-validation and fit many models
- below shows hyper-parameter estimations for each model engine
- given all the moving pieces, we'll use racing to quickly drop the bad performers

```{r}
set.seed(72)
# we also want to ensure the sampling is stratified
# this ensures the samples represent all quantiles of our magpie count
folds <- vfold_cv(training, v = 10, strata = magpie_count)

## random forest ----
forest_grid <- expand.grid(
  trees = c(20, 50),
  mtry  = c(2, 4, 6)
)

orf_spec <- rand_forest(
  trees = tune(),
  mtry  = tune()
  ) |>
  # accelerated oblique random forest
  set_engine("aorsf") |>
  set_mode("regression")

orf_results <- finetune::tune_race_anova(
  object = workflow() |>
    # our pre-process
    add_recipe(magpie_recipe) |>
    # our model
    add_model(orf_spec),
  # folds
  resamples = folds,
  # parameters
  grid = forest_grid,
  # racing as control
  control = control_race(),
  # loss function to minimize
  metrics = metric_set(yardstick::rmse)
)

## boosted tree ----
lgb_grid <- expand.grid(
  trees      = c(3, 10, 25),
  learn_rate = c(0.5, 0.05, 0.005)
)

lgb_spec <- boost_tree(
  trees      = tune(),
  learn_rate = tune()
  ) |>
  # boosted lgbm model
  set_engine("lightgbm") |>
  set_mode("regression")

lgb_results <- finetune::tune_race_anova(
    workflow() |>
      # our pre-process
      add_recipe(magpie_recipe) |>
      # our model
      add_model(lgb_spec),
    # folds
    resamples = folds,
    # parameters
    grid = lgb_grid,
    # racing as control
    control = control_race(),
    # loss function to minimize
    metrics = metric_set(yardstick::rmse)
  )

# cubist rules -----
cub_grid <- expand.grid(
  committees      = c(30, 45),
  max_rules       = c(5, 10, 30)
)

cub_spec <- cubist_rules(
  committees = tune(),
  max_rules  = tune()
  ) |>
  # quinlan's cubist
  set_engine("Cubist") |>
  set_mode("regression")

cub_results <- tune_race_anova(
  object = workflow() |>
    # our pre-process
    add_recipe(magpie_recipe) |>
    # our model
    add_model(cub_spec),
  # folds
  resamples = folds,
  # parameters
  grid = cub_grid,
  # racing as control
  control = control_race(),
  # loss function to minimize
  metrics = metric_set(rmse)
)


## svm ----
svm_grid <- expand.grid(
  laplace_sigma = c(0.5, 1, 2, 5)
)

svm_spec <- svm_laplace(
  laplace_sigma = tune()
) |>
  # support vector with laplacian kernel
  set_engine("kernlab") |>
  set_mode("regression")

svm_results <- finetune::tune_race_anova(
    workflow() |>
      # our pre-process
      add_recipe(magpie_recipe) |>
      # our model
      add_model(svm_spec),
    # folds
    resamples = folds,
    # parameters
    grid = svm_grid,
    # racing as control
    control = control_race(),
    # loss function to minimize
    metrics = metric_set(yardstick::rmse)
  )
```

## Training Stage, Part 3

- selecting our best performers based on analysis and assessment
- next is to do a final fit
- taking the best hyperparameter sets and fitting to the entire training data

```{r}
select_best(orf_results)
select_best(lgb_results)
select_best(cub_results)
select_best(svm_results)

# final fits with all workflow details ---
## random forest ----
orf_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
      add_recipe(magpie_recipe) |>
      add_model(orf_spec),
    parameters = select_best(orf_results)
    ),
  data = training
  )

## boosted tree ----
lgb_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
        add_recipe(magpie_recipe) |>
        add_model(lgb_spec),
    parameters = select_best(lgb_results)
    ),
  data = training
  )

# cubist rules ----
cub_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
      add_recipe(magpie_recipe) |>
      add_model(cub_spec),
      select_best(cub_results)
    ),
  data = training
  )

# support vector machine ----
svm_fit <- fit(
  object = finalize_workflow(
    x = workflow() |>
        add_recipe(magpie_recipe) |>
        add_model(svm_spec),
    parameters = select_best(svm_results)
    ),
  data = training
  )
```

## Validation Stage

- Validation data will be used to validate, evaluate, calibrate our models
- add prediction intervals to point forecasts
- post-processing is important to handle any adjustments and bias for predictions
- we can then use the calibrated models on our test dataset for a final evaluation

```{r}
# final models & probably
library(probably) # point calibration and intervals

# predictions for validation data
orf_vpreds <- predict(orf_fit, validate)
lgb_vpreds <- predict(lgb_fit, validate)
cub_vpreds <- predict(cub_fit, validate)
svm_vpreds <- predict(svm_fit, validate)

# making a large dataframe with all preds and truth
validate_preds <- validate |>
  cbind(
    orf = orf_vpreds$.pred,
    lgb = lgb_vpreds$.pred,
    cub = cub_vpreds$.pred,
    svm = svm_vpreds$.pred
  ) |>
  dplyr::select(magpie_count:svm) |>
  tidyr::pivot_longer(-magpie_count)

# visualize the certainty in predictions
ggplotly(
  validate_preds |>
    ggplot() +
    geom_point(aes(x = magpie_count, y = value, color = name)) +
    coord_equal() +
    geom_abline(lty = 2) +
    labs(
      title = "magpie count & predictions",
      caption = "models disagree at high magpie counts"
    )
)
# quick error measurement
validate_preds |>
  mutate(
    abs_error = abs(magpie_count - value)
  ) |>
  group_by(name) |>
  summarize(
    total_abs_error = sum(abs_error)
  )
```

## Calibration Stage

- our top performing models on validation are cubist and random forest
- cubist is already pretty precise and accurate
- but we will calibrate our random forest to see if we increase accuracy on validation data

```{r}
# calibrate ORF with validation data
calibrate_orf <- data.frame(
  magpie_count = validate$magpie_count,
  .pred        = orf_vpreds$.pred
)

orf_fit_cal <- calibrate_orf |>
  cal_estimate_linear(magpie_count)

calibrated_orf_preds <- calibrate_orf |>
  cal_apply(orf_fit_cal)

calibrated_orf_preds |>
  mutate(
    abs_error = abs(magpie_count - .pred)
  ) |>
  # there is an improvement but note we need to 'test' the calibration
  summarize(total_abs_error = sum(abs_error))
```

## Model with Conformal Coverage

- while we have optimized our point forecasts with pre-, tuning, and post-processing
- it's also important to capture our model uncertainty
- create robust prediction intervals via validation data

```{r}
# create confidence intervals for our models

### simplest form being a conformal split method --
svm_int_model <- probably::int_conformal_split(svm_fit, validate)

# just like that, we have a conformal interval model!
svm_int_model
```

## Testing Stage

- Testing data will *finally* be used to see the overall performance

```{r}
# our original top performer in the validation stage
cubist_preds <- predict(cub_fit, testing)

# our 2nd best, calibrated during validation stage
orf_cal_preds <- predict(orf_fit, testing) |> cal_apply(orf_fit_cal)

# a weaker model, to show conformal prediction intervals
svm_int_preds <- predict(svm_int_model, testing, level = 0.875)

test_df <- testing |>
  cbind(
    cubist = cubist_preds$.pred,
    orf_cal = orf_cal_preds$.pred,
    svm_upper = svm_int_preds$.pred_upper,
    svm_lower = svm_int_preds$.pred_lower
  )

ggplotly(
  test_df |>
    ggplot() +
    geom_point(aes(x = magpie_count, y = magpie_count, color = "actual")) +
    geom_errorbar(aes(xmin = svm_lower, xmax = svm_upper, y = magpie_count, color = "svm_interval"), alpha = .4) +
    geom_point(aes(x = cubist, y = magpie_count, color = "cubist")) +
    geom_point(aes(x = orf_cal, y = magpie_count, color = "orf_calibrated")) +
    coord_equal() +
    geom_abline(lty = 2)
)

# test accuracy:
test_df |>
  mutate(
    cub_abs_error = abs(magpie_count - cubist),
    orf_abs_error = abs(magpie_count - orf_cal),
    svm_coverage  = ifelse(magpie_count >= svm_lower & magpie_count <= svm_upper, 1, 0),
    n             = row_number()
  ) |>
  summarize(
    cub_total_abs_error = sum(cub_abs_error),
    orf_total_abs_error = sum(orf_abs_error),
    svm_coverage_pct = sum(svm_coverage) / max(n)
  )
```

## Next Steps

- `{workflowsets}`: methods to test across many engines and recipe combinations
- `{yardstick}`: other loss function or metrics to optimize
- `{dials}`: for creating draws for hyperparameters
- `{autoML}`: more automation for engine selection and ensembling
- `{tailor}`: a new post-processing technique for workflows
- `{stacks}`: we can also ensemble many models to outperform any one model
- meta learners & diagnostics: while in production, we can refit as we have additional observations, potentially add another model to learn best hyperparameter setpoints for best out-of-sample accuracy
- it's important to understand trade-offs of the model in operation over time and when to refit, calibrate, or start from scratch


Let's plot the daily average maximum temperature obtained from the NASA POWER project

```{r}
#| label: fig-power-max-temp
#| fig-cap: "Average maximum temperature (°C)"

fig_tmax <- power_cherry |>
  ggplot(aes(x = date)) +
  geom_hline(yintercept = 0, colour = "grey") +
  geom_line(aes(y = t2m_max), colour = "#d62828") +
  labs(
    title = "Average maximum temperature (°C)", 
    x = NULL, y = "Degrees (°C)"
  ) +
  theme_minimal() +
  facet_grid(rows = vars(location))

print(fig_tmax)

ggsave(filename = "figures/fig_tmax.png", plot = fig_tmax, bg = "white")
```
Let's plot the daily average minimum temperature obtained from the NASA POWER project

```{r}
#| label: fig-power-min-temp
#| fig-cap: "Average minimum temperature (°C)"

fig_tmin <- power_cherry |>
  ggplot(aes(x = date)) +
  geom_hline(yintercept = 0, colour = "grey") +
  geom_line(aes(y = t2m_min), colour = "#00b4d8") +
  labs(
    title = "Average minimum temperature (°C)",
    x = NULL, y = "Degrees (°C)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_tmin)

ggsave(filename = "figures/fig_tmin.png", plot = fig_tmin, bg = "white")
```
Plot `frost_days` from NASA Power: 

Create average number of frost days

```{r}
power_cherry <- power_cherry |> 
  mutate(year = year(date)) |> 
  group_by(location, year) |> 
  arrange(location, year, date) |> 
  mutate(afrost_days = cumsum(frost_days)) |> 
  ungroup() |> 
  select(-year)
```


```{r}
#| label: fig-frost_days
#| fig-cap: "Average preciptation"

fig_prcp <- power_cherry |>
  ggplot(aes(x = date, y = afrost_days)) +
  geom_line(colour = "#4361ee") +
  labs(
    title = "Average preciptation (in mm)",
    x = NULL, y = "Average preciptation (in mm)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_prcp)

ggsave(filename = "figures/fig_prcp.png", plot = fig_prcp, bg = "white")
```


Plot the `prcp` from NASA POWER:

```{r}
#| label: fig-avg-prcp
#| fig-cap: "Average preciptation"

fig_prcp <- power_cherry |>
  ggplot(aes(x = date, y = prcp)) +
  geom_line(colour = "#4361ee") +
  labs(
    title = "Average preciptation (in mm)",
    x = NULL, y = "Average preciptation (in mm)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_prcp)

ggsave(filename = "figures/fig_prcp.png", plot = fig_prcp, bg = "white")
```

Let's plot `APRCP`:

```{r}
#| label: fig-aprcp
#| fig-cap: "Average aggregated preciptation since January 1st (in mm)"

fig_aprcp <- power_cherry |>
  ggplot(aes(x = date, y = aprcp)) +
  geom_line(colour = "#4361ee") +
  labs(
    title = "Average aggregated preciptation since January 1st (in mm)",
    x = NULL, y = "Average agregated preciptation (in mm)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_aprcp)

ggsave(filename = "figures/fig_aprcp.png", plot = fig_aprcp, bg = "white")
```

Plot cooling and heating degree days
  "CDD0", # temp above 0C
  "CDD18_3",# Temp above 18C
  "HDD0", # Temp below 0C
  "HDD18_3", # Temp below 18C

```{r}
power_cherry |> 
  ggplot(aes(x = date)) +
  geom_line(aes(y = hdd18_3), colour = "red") +
  facet_grid(rows = vars(location))
```
  



Let's plot the Growing Degree Days (GDD) from NASA POWER

```{r}
#| label: fig-gdd
#| fig-cap: "Growing Degree Days (GDD) calculated from January 1st"

fig_gdd <- power_cherry |>
  ggplot(aes(x = date, y = gdd)) +
  geom_line(colour = "#d62828") +
  labs(
    title = "Growing Degree Days (GDD)",
    subtitle = "Calculated from January 1st",
    x = NULL, y = "Growing Degree Days (GDD)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_gdd)

ggsave(filename = "figures/fig_gdd.png", plot = fig_gdd, bg = "white")
```

Let's plot Accumulated Growing Degree Days (AGDD) from NASA POWER

```{r}
#| label: fig-agdd
#| fig-cap: "Accumulated Growing Degree Days (AGDD) calculated from January 1s"

fig_agdd <- power_cherry |>
  ggplot(aes(x = date, y = agdd)) +
  geom_step(colour = "#d62828") +
  labs(
    title = "Accumulated Growing Degree Days (AGDD)", 
    subtitle = "Calculated from January 1st",
    x = NULL, y = "Accumulated Growing Degree Days (AGDD)"
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_agdd)

ggsave(filename = "figures/fig_agdd.png", plot = fig_agdd, bg = "white")
```



Let's plot the Chilling Degree Days (CDD) from NASA POWER

```{r}
#| label: fig-plt-cdd
#| fig-cap: "Chilling Degree Days (CDD) calculated from October 1st"

fig_cdd <- power_cherry |>
  ggplot(aes(x = date, y = cdd)) +
  geom_line(colour = "#00b4d8") +
  labs(
    title = "Chilling Degree Days (CDD)",
    subtitle = "Calculated from October 1st",
    x = NULL, y = "Chilling Degree Days",
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_cdd)

ggsave(filename = "figures/fig_cdd.png", plot = fig_cdd, bg = "white")
```

Let's plot Accumulated Chilling Degree Days (ACDD) from NASA POWER

```{r}
#| label: fig-plt-acdd
#| fig-cap: "Accumulated Chilling Degree Days (ACDD) calculated from October 1st"

fig_acdd <- power_cherry |>
  ggplot(aes(x = date, y = acdd)) +
  geom_line(colour = "#00b4d8") +
  labs(
    title = "Accumulated Chilling Degree Days (ACDD)",
    subtitle = "Calculated from October 1st",
    x = NULL, y = "Chilling Degree Days",
  ) +
  theme_light() +
  facet_grid(rows = vars(location))

print(fig_acdd)

ggsave(filename = "figures/fig_acdd.png", plot = fig_acdd, bg = "white")
```


### Nonlinear models

#### Random forest

Let's fit a random forest

```{r}
cores <- parallel::detectCores()
cores

# Split the data with {rsamples}
set.seed(123)
cherry_split <- initial_split(data = cherry_weeks)
cherry_train <- training(cherry_split)
cherry_test  <- testing(cherry_split)

# Create folds with {rsamples}
set.seed(345)
folds <- vfold_cv(cherry_train, v = 10)
folds

# Create model with {parsnip}
rf_mod <- rand_forest(
  mtry = tune(),
  min_n = tune(),
  trees = 1000
  ) |> 
  set_engine("ranger", num.threads = cores, importance = "impurity") |> 
  set_mode("regression")

# show what will be tuned
extract_parameter_set_dials(rf_mod)

# Create worfklow with {workflows}
rf_workflow <- workflow() |> 
  add_model(rf_mod) |> 
  add_formula(bloom_doy ~ .)

# Create a tuning grid with {dials}
tree_grid <- grid_regular(
  mtry(range = c(1, 9)),
  min_n(),
  levels = 5
)

# Train the model on the folds
tree_res <- rf_workflow |> 
  tune_grid(
    resamples = folds,
    grid = tree_grid,
    control = control_grid(save_pred = TRUE),
    metrics = metric_set(rmse, mae)
  )

tree_res |> 
  collect_metrics()

tree_res |> 
  show_best(metric = "rmse")

best_tree <- tree_res |> 
  select_best(metric = "rmse")

# Finalize the model
final_wf <- rf_workflow |> 
  finalize_workflow(parameters = best_tree)

final_wf

# Fit the model best model on all the training data
# with the {tune} function last_fit()
final_fit <- final_wf |> 
  last_fit(cherry_split)

final_fit |> 
  collect_metrics()

final_fit |> 
  collect_predictions() |> 
  ggplot(mapping = aes(x = bloom_doy, y = .pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  coord_equal(ratio = 1) +
  theme_bw()

# To predict new data points, use {workflows} extract_worfklow()
final_tree <- extract_workflow(final_fit)

final_tree

final_tree |> 
  extract_fit_parsnip() |> 
  vip()
```


#### LightGBM

```{r}

```


#### XGboost

```{r}

```


## Submission

To submit your entries, enter the predicted bloom dates from your models in the **submission form**.

```{r}
#| label: tbl-predictions-for-submission
#| tbl-cap: "Predictions for Submission"

predictions |>
  filter(year == 2025) |>
  mutate(prediction_date = strptime(paste(year, prediction), "%Y %j") |> as_date()) |>
  kable() |> 
  kable_styling()
```

**NOTE**: The question is if we need to round the point estimates or if we can submit partial days.